---
title: "Geospatial Analysis with R"
author: "Lyndon Estes & Zhiwen Zhu"
date: "8/30/2017"
output:
  slidy_presentation: default
  ioslides_presentation: default
subtitle: Class 2
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
```

## Things to Cover Today

- Progress thus far
    - Rstudio and git installed and working?
    - `geospaar` installed?
    - able to do this?

```{r, eval = FALSE}
browseVignettes("geospaar")
# or this, for individual vignette?
vignette("syllabus", package = "geospaar")
```

- Your questions
- Key concepts/techniques

## Your questions

Prompts: 

- What was least clear? 
- What seemed obvious/made most sense? 
- Feel free to say out loud, or send slack DM to Zhiwen

## Key concepts/techniques
- Replicability versus reproducibility, and what we mean by reproducible here
- Objects, functions, arguments, assignment
- How to get help
- Where packages live: 
    - after being built
    - when they are being built
- Git, GitHub, and RStudio
    - When you have to use shell scripts

## Examples

### Help
```{r, eval = FALSE}
?matrix  # exact function name
??matrices  # search for matches 
apropos("matrix")  # searches matching functions loaded in memory
help("matrix")
help("bs")
help("bs", try.all.packages = TRUE)
```

http://rseek.org
http://www.google.com
https://duckduckgo.com
https://www.r-project.org/mail.html


### Where Built Packages Live
```{r, eval = FALSE}
.libPaths()
dir(paste0(.libPaths(), "/geospaar"))
R.home()
dir(R.home())
dir(paste0(R.home(), "/library"))
```

