---
title: "Geospatial Analysis with R"
author: "Lyndon Estes & Zhiwen Zhu"
date: "Halloween -1"
output:
  slidy_presentation: 
    default:
    css: custom.css
    fig_height: 4
    fig_width: 6
subtitle: Class 17
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# options(repos = c(CRAN = "http://ftp.ussg.iu.edu/CRAN/"))
```

## Today

- Questions from last week?
- Projects
- Exercises from readings
- Additional exercises

## Projects
```{r, warning=FALSE, message=FALSE}
library(data.table)
projects <- fread("project-list.csv")
projects[order(Topic, Student), ]
```


## Exercises from reading

### Rapporteurs

```{r}
student_names <- read.csv("students.csv", stringsAsFactors = FALSE)
set.seed(1)
sel_names <- student_names[sample(1:nrow(student_names), size = 6, 
                                  replace = FALSE), ]
monday_names <- c("Sarah Gates", "Ali Filipovic")
student_names2 <- student_names$name[!student_names$name %in% 
                                       c(sel_names, monday_names)]
set.seed(100)
sel_names <- student_names2[sample(1:length(student_names2), size = 3)]
sel_names
```

## Exercise Datasets

```{r}
zamr <- raster(x = extent(districts), crs = crs(districts), res = 0.1)
values(zamr) <- 1:ncell(zamr)

districts$ID <- 1:length(districts)
distsr <- rasterize(x = districts, y = zamr, field = "ID")
distsr

farmers <- read.csv(system.file("extdata/farmer_spatial.csv", 
                                package = "geospaar"), stringsAsFactors = FALSE)
farmers2 <- do.call(rbind, lapply(unique(farmers$uuid), function(x) {
  dat <- farmers[farmers$uuid == x, ][1, ]  # select first row only
}))
coordinates(farmers2) <- ~lon + lat
farmers2$ct <- 1  # assign value of 1 to each farmers

zamr2 <- raster(x = extent(districts), crs = crs(districts), res = 0.25)
values(zamr2) <- 1:ncell(zamr2)
farmersr <- rasterize(farmers2, zamr2, field = "ct", fun = sum)
```


## Exercise 6

Mask `farmersr` using `districts`, and plot the district boundaries over the masked result, suppressing the axes and box around the plot 

## Exercise 7

Crop `chirpsz[[1]]` using the extent of `districts[57, ]`, and disaggregate it to 0.01 resolution using both the default and bilinear methods. Plot the results side by side using `plot_noaxes`

## Exercise 8

Use zonal stats on `farmersr_rs` to calculate the total number of farmers per district, and then map them onto the districts as we did above. Remember that `farmersr_rs` is a coarser resolution than `distsr_rs`, but aggregating the districts further would lose a lot information. Disaggregating is also problematic because it messes up the zonal stats. Better to first re-rasterise `farmers2` to the 0.1 resolution, using `distsr_rs` as the reference (to get the right extent) and then calculate and map zonal stats

## Additional exercises - new Zambia data

1. Use `readOGR` to read in this dataset:

```
https://www.dropbox.com/s/wd6doedusc8zv1l/chomafields.sqlite?dl=0
```

2. Reproject it to Albers Equal Area Conic (same as one we use)

3. Calculate area of each field in ha

4. Calculate total area by farmers

## Additional exercises - Git/GitHub

1. Create a new project (R package) with new repo locally (xyza2)
2. Run `devtools::use_vignette()`
3. Copy into: 
    a. vignettes folder you Vignette from assignment 3
    b. R folder your function from assignment 2
4. Build package and commit
5. Create remote repo of same name on GitHub, and push local repo onto that. 
6. Back in local, create new branch "b1"
7. While in branch b1, copy into R folder the cat function from assignment 1, and your vignette from assignment 2 into vignettes
8. Rebuild package and commit
9. checkout your master branch
10. from shell, run `git merge b1`
11. Examine results. 
12. Now delete your GitHub remote and local 










