---
title: "ARS - Spectral Unmixing"
author: "Mike Cecil"
date: "2023-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install `RStoolbox` from source. 

```{r}
library(devtools)
install_github("bleutner/RStoolbox")
```

This may throw an error, in which case install any other dependencies then rerun above. 

```{r}
#install.packages("", "", "") ## install dependencies
```

Load other libraries
```{r cars}
library(dplyr)
library(raster)
library(sf)
library(RStoolbox)
```

Set up file paths 
```{r}
taxonomic_path <- "C:/Users/micha/Downloads/Endmembers/TaxonomicEndmembers.shp"
biophysical_path <- "C:/Users/micha/Downloads/Endmembers/BiophysicalEndmembers.shp"
landsat_path <- "C:/Users/micha/Downloads/Landsat8SR_SD_July2019_scale.tif"
```

Read in data
```{r}
tax_classes <- st_read (taxonomic_path)
#bio_classes <- st_read(biophysical_path)
lsat <- raster::brick(landsat_path)
```

Plot data
```{r}
plot(lsat[[4]])
plot(tax_classes, add = T)
```
Extract pixel values at point locations.

```{r}
tax_em <- extract(lsat, tax_classes %>% st_centroid()) %>% 
  cbind(tax_classes, .) %>% as_tibble()
tax_em
```
Because we have two points per land cover class, we will use `group_by` to average
```{r}
tax_group <- tax_em %>% group_by(class) %>%
  summarise(mean_b1 = mean(Landsat8SR_SD_July2019_scale_1),
            mean_b2 = mean(Landsat8SR_SD_July2019_scale_2),
            mean_b3 = mean(Landsat8SR_SD_July2019_scale_3),
            mean_b4 = mean(Landsat8SR_SD_July2019_scale_4),
            mean_b5 = mean(Landsat8SR_SD_July2019_scale_5)) %>%
  st_drop_geometry()
tax_group
```

We need our spectral table to only have a column for each band. So we'll put the land cover classes into row names (instead of as a column.)

```{r}

class_names <-  tax_group$class

tax_group <- tax_group %>% data.frame() %>%  dplyr::select(-class)
rownames(tax_group) <- class_names
tax_group
```

Perform spectral unmixing. The issue with this algorithm is that the class percents may not add to 1. (this may take 10-20 min)

```{r}

probs_tax <- mesma(lsat,
                 tax_group,
                 method = "NNLS",
                 iterate = 1000)

plot(probs_tax)
```

The code below is used to "rescale" the probabilities so they add to 1. 

```{r}
sum_tax <- sum(probs_tax) - probs_tax[['RMSE']]
probs_tax_adj <- probs_tax
## for loop rescales each band
for(band in class_names){
 # print(band)
  probs_tax_adj[[band]] <- probs_tax_adj[[band]]/sum_tax
}

summary(probs_tax_adj)


```

Let's check that the adjusted values sum to 1.

```{r}
sum_tax_adj <- sum(probs_tax_adj) - probs_tax_adj[['RMSE']] ## sum all bands except RMSE
summary(sum_tax_adj)
plot(sum_tax_adj)
```

Now let's plot the final class percents

```{r}
plot(probs_tax_adj, zlim = c(0,1))
```

And we can save the results to a tif file.

```{r}
output_folder <- "C:/Users/micha/Downloads/"
writeRaster(probs_tax_adj, file = paste0(output_folder, "class_probabilities.tif"))
```

