---
title: "Assessment"
subtitle: "GEOG246-346"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: pygments
    number_sections: no
    toc_depth: 3
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

***
<center>
[Back to **vignette index**](toc.html)
</center>
***


# Overall Grades{#grading}

Final letter grades will be assigned as follows
```{r, echo = FALSE}
gr <- rbind(c(94, 90, 88, 83, 80, 78, 73, 70, 68, 63, 60, 0), 
            c(100, 94, 90, 88, 83, 80, 78, 73, 70, 68, 63, 60))
lt <- c("A ", "A-", "B+", "B ", "B-", "C+", "C ", "C-", "D+", "D ", "D-", "F ")
DF <- cbind(c("lower", "upper", "grade"), rbind.data.frame(gr, lt))
colnames(DF) <- rep("", ncol(DF))
kable(DF)
```

# Participation

Participation is worth 10% of your final grade and has two components.
- Attendance. You are expected to attend each class. If you cannot make it for some reason, please arrange ahead of time. 
- Engagement with the course material, including any coding work assigned for homework. It is essential to read and work through the class material between classes. You will not be able to learn everything just from class. Additionally, you will at times be asked to work on practical problems between classes, for which you should be prepared to share your code and provide examples of errors/obstacles. 
- Reading quizzes. Periodically, short quizzes may be given on core concepts from readings. These quizzes will be closed-book and not exceed 10 minutes. (e.g. explain in general terms how an apply function works). Students receiving 60% or higher on a quiz receive a "Complete" grade.

# For assignments

Assessment: Can we load the data you used, run your code, and arrive at the same result you reported? Does your code conform to the style guide (note: we are a little obsessive about this)

We will grade using this rubric (or one close to it that captures the results/style/reproducibility dimensions).

```{r, echo=FALSE}
mat <- rbind(c("Results all wrong (0-5%)", 0, 
               "Can't be made to run", 0, 
               "Indecipherable code, no comments/documentation", 
               0), 
             c("Results mostly wrong (5-25%)", 4,
               "Doesn't work", 4,
               "Indecipherable code, no comments/documentation", 
               2), 
             c("Results more wrong than right (25-50%)", 8, 
               "Doesn't run without numerous fixes", 8, 
               "Code indecipherable/no comments/documentation", 
               4), 
             c("Results more right than wrong (50-75%)", 12,
               "Runs after a handful of fixes", 12,
               "Code readable but wrong style/some commenting/doc'tation",
               6), 
             c("Results almost entirely correct (75-95%)", 16,
               "Runs after some minor tweaks", 16,
               "Code mostly conformant/adequate commenting/documentation", 8), 
             c("Results all correct (95-100%)", 20, 
               "Runs out of the box", 20,
               "Code conformant and stylish/well commented/documented", 10))
DF <- as.data.frame(mat, stringsAsFactors = FALSE)
colnames(DF) <- c("Accuracy", "Acc. points", "Reproducibility", 
                  "Repro. points",
                  "Elegance", "Ele. points")
for(i in colnames(DF)[c(2, 4, 6)]) DF[, i] <- as.numeric(DF[, i])
# reshape2::dcast(DF, Accuracy + Elegance + Reproducibility ~ Points)
# DF2 <- rbind(t(DF[1:6, 1:3]), DF[1:6, 4])
kable(DF, align = "r")
```

***
<center>
[Back to top](#grading) || [Back to **vignette index**](toc.html)
</center>
***

# Final project overview
For the project overview assignment, assessment will be out of 50 points using simplified and somewhat different categories than the unit assignments, as shown below. The required content for this particular assignment in the [Final Projects section](projects.html).

Note that the minimum number of points assigned per category assume that the final project overview was completed and handed in. 

```{r, echo=FALSE}
quality <- c(
  "No plan (analysis/coding approach and timelines) provided",
  paste0("Plan (analysis/coding approach and timelines) is ",
           "provided but mostly unworkable, many holes or ",
           "inconsistencies in logic or approach"),
  paste0("Probably workable plan, but with a fair number of holes",
           " or inconsistencies that need to be addressed"),
  paste0("A pretty well thought-out and workable plan, with ",
           "just a handful of holes or inconsistencies"),
  "Excellent, well thought plan, no obvious flaws in approach"
)
pts1 <- c(5, 15, 20, 25, 30)

struct <- c(
  "Doesn't follow required structure or provide requested information",
  "Barely follows required structure or provides requested information",
  "Partially follows structure and provides part of the information",
  "Mostly follows structure and provides most of the required information",
  "Perfectly follows structure and provides all of the required information"
)
pts2 <- c(2, 4, 6, 8, 10)

reproducibility <- c(
  "Can't be built/no R package structure", 
  "Runs after many fixes", 
  "Runs after 4-6 fixes",
  "Runs after 1-3 fixes",
  "Runs out of the box/with no fixes needed"
)
pts3 <- pts2


DF <- data.frame(quality, pts1, struct, pts2, reproducibility, pts3)
colnames(DF) <- c("Quality of Plan", "QPts", "Structure/completeness", 
                  "Cpts", "Reproducibility", "RPts")
for(i in colnames(DF)[c(2, 4, 6)]) DF[, i] <- as.numeric(DF[, i])
# reshape2::dcast(DF, Accuracy + Elegance + Reproducibility ~ Points)
# DF2 <- rbind(t(DF[1:6, 1:3]), DF[1:6, 4])
kable(DF, align = "r")
```


***
<center>
[Back to top](#grading) || [Back to **vignette index**](toc.html)
</center>
***

# Final project
Assessment is out of 70 pts (note that there a minimum points floor, which assumes that a final project has been submitted), and will be based on four categories: 

- **Quality** evaluates the **content** of the material in your package, focusing on the information and ideas conveyed in your descriptions, figures, and tables. How well did you understand and execute the project, and how well did you convey what you did? 
- **Progress** evaluates whether you achieved the objectives put forward in your Assignment 6 and final presentation, focusing more on the latter due to course corrections (and incorporating any necessary changes we discussed)
- **Clarity** examines the clarity of the writing (is it easy to understand, no typos or missed words), the visual aspects of figures (does the color choice make sense, legend sensible, size appropriate, text readable?), formatting (did html lists end up as lists? do headings make sense?), and code syntax. This basically covers the stylistic components of your vignettes and code.
- **Reproducibility** relates to the now familiar criteria about whether your package builds for another user without error, and with browsable vignettes.  

Note that the minimum number of points assigned per category assume that the final project was completed and handed in. 

```{r, echo = FALSE, message = FALSE}
# library(dplyr)
# html_table_width <- function(kable_output, width){
#   width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
#   sub("<table>", paste0("<table>\n", width_html), kable_output)
# }
quality <- c(paste("Objectives/methods/results/interpretations unclear,", 
                   "poor understanding conveyed", 
                   "replete with analytical/methodological errors", 
                   "Findings/interpretations/examples absent"), 
             paste("Objectives/methods/results/interpretations cloudy,", 
                   "understanding of project less than half",
                   "many analytical/methodological errors", 
                   "Findings/interpretations/examples present but weak/wrong"), 
             paste("Objectives/methods/results/interpretations 50% clear,", 
                   "project about 50% understood", 
                   "some analytical/methodological errors", 
                   "Findings/interpretations/examples about half wrong"),
             paste("Objectives/methods/results/interpretations 75% clear,", 
                   "project about 75% understood", 
                   "few analytical/methodological errors", 
                   "Findings/interpretations/examples interesting/correct"),
             paste("Objectives/methods/results/interpretations crystal clear,", 
                   "project completely understood", 
                   "no obvious analytical/methodological errors", 
                   "Findings/interpretations/examples novel/really well done"))
pts1 <- c(10, 15, 20, 25, 30)
progress <- c("None since presentation", "Only a little", "Modest progress", 
              "Project mostly complete", "Much greater than expected")
pts2 <- c(5, 7.5, 10, 12.5, 15)
clarity <- c(paste("Vignette(s) a complete mess,",
                   "graphics absent/totally confusing,",
                   "text a word salad,",
                   "narrative impossible to follow", 
                   "code a mess/doesn't follow style"),
             paste("Vignette(s) more unclear than clear;", 
                   "graphics present but hard to decipher;", 
                   "text somewhat confusing;", 
                   "narrative hard to follow", 
                   "code messy/mostly doesn't follow style"), 
             paste("Vignette(s) not unreasonable; graphics so-so,", 
                   "not spectacular, maybe hard to decipher;",
                   "text still a bit distracting, causes some confusion;",
                   "narrative mostly clear", 
                   "code sloppy/half follows style"), 
             paste("Vignette(s) good;", 
                   "graphics mostly clear, no major questions;",
                   "text understood and gets the job done;", 
                   "narrative clear", 
                   "code clear/mostly follows style"), 
             paste("Vignette(s) outstanding; graphics spectacular,", 
                   "tell a clear and easily understood story supporting text;", 
                   "text crystal clear and right amount;", 
                   "gripping narrative", 
                   "code tight, perfect style"))
pts3 <- c(5, 7.5, 10, 12.5, 15)
reproducibility <- c("Can't be built/no R package structure", 
                     "Runs after many fixes", "Runs after 4-6 fixes",
                     "Runs after 1-3 fixes",
                     "Runs out of the box/with no fixes needed")
pts4 <- c(3, 5, 7, 8.5, 10)

DF <- cbind.data.frame(quality, pts1, progress, pts2, clarity, pts3,
                       reproducibility, pts4, stringsAsFactors = FALSE)
colnames(DF) <- c("Quality", "QPts", "Progress", "PrPts", "Clarity",
                  "Cpts", "Reproducbility", "RPts")
y <- knitr::kable(DF, format = "html")
kableExtra::column_spec(y, column = 1:8, width = "10em")
# kable(DF[, 3:4], format = "html") %>% html_table_width(c(50, 20))
# kable(DF[, 5:6], format = "html") %>% html_table_width(c(50, 20))
# kable(DF[, 7:8], format = "html") %>% html_table_width(c(50, 20))
# kable(DF, align = "r") 
```

***
<center>
[Back to top](#grading) || [Back to **vignette index**](toc.html)
</center>
***
