---
title: "Assessment"
author: "Lyndon Estes"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: yes
    toc_depth: 3
    toc: yes
vignette: >
  %\VignetteIndexEntry{Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Grading

__[Under construction]__

## Overall

Final letter grades will be assigned as follows
```{r, echo = FALSE}
gr <- rbind(c(99, 93, 90, 88, 83, 80, 78, 73, 70, 68, 63, 60, 0), 
            c(100, 98, 92, 89, 87, 82, 79, 77, 72, 69, 67, 62, 59))
lt <- c("A+", "A ", "A-", "B+", "B ", "B-", "C+", "C ", "C-", "D+", "D ", "D-",
        "F ")
DF <- cbind(c("lower", "upper", "grade"), rbind.data.frame(gr, lt))
colnames(DF) <- rep("", ncol(DF))
kable(DF)
```


## For assignments

Assessment: Can we load the data you used, run your code, and arrive at the same result you reported?

___[Still under construction]___ We will grade using this rubric (or one close to it that captures the results/style/reproducibility dimensions).

```{r, echo=FALSE}
mat <- rbind(c("Expected results all wrong (0-5%)", 0, 
               "Can't be made to run", 0, 
               "Indecipherable code, no comments or documentation", 
               0), 
             c("Expected results mostly wrong (5-25%)", 4,
               "Doesn't work", 4,
               "Indecipherable code, no comments or documentation", 
               2), 
             c("Expected results more wrong than right (25-50%)", 8, 
               "Doesn't run without numerous fixes", 8, 
               "Code indecipherable or no comments/documentation", 
               4), 
             c("Expected results more right than wrong (50-75%)", 12,
               "Runs after a handful of fixes", 12,
               "Code readable but wrong style/some commenting/documentation",
               6), 
             c("Expected results almost entirely correct (75-95%)", 16,
               "Runs after some minor tweaks", 16,
               "Code pretty clean/decent commenting & documentation", 8), 
             c("Expected results all right (95-100%)", 20, 
               "Runs out of the box", 20,
               "Code elegant and stylish/well commented & documented", 10))
DF <- as.data.frame(mat, stringsAsFactors = FALSE)
colnames(DF) <- c("Accuracy", "Acc. Points", "Elegance", "Ele. Points",
                  "Reproducibility", "Repro. Points")
for(i in colnames(DF)[c(2, 4, 6)]) DF[, i] <- as.numeric(DF[, i])
# reshape2::dcast(DF, Accuracy + Elegance + Reproducibility ~ Points)
# DF2 <- rbind(t(DF[1:6, 1:3]), DF[1:6, 4])
kable(DF, align = "r")
```

## For presentations
___[Under construction]___

## For final paper/report
___[Under construction]___
