---
title: "Helpful functions"
subtitle: "GEOG246-346"
date: "`r Sys.Date()` [Back to home](index.html)"
output:
  rmdformats::readthedown:
    highlight: pygments
    number_sections: yes
    toc_depth: 3
    css: unit.css
vignette: >
  %\VignetteIndexEntry{helpful functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  comment = "#>"
)
library(knitr)
```


# strings{#strings}
 
## paste, paste0{#paste}
Concatenate strings. `paste` includes the option to add a specific separator, like a space or hyphen.
```{r}
paste("a", "b", "c", sep = " ")
paste("a", "b", "c", sep = "-")
```

`paste0` assumes there is no separator.
```{r}
paste0("a", "b", "c")
```

## str_replace{#str_replace}

From `stringr` package. Use to replace strings
```{r}
library(stringr)
v <- "king phillip came over for good soup"
print(v)
w <- stringr::str_replace(v, "soup", "spaghetti" )
print(w)
```

## str_replace_all{#str_replace_all}

Similar to `str_replace`, but `str_replace_all` replaces  
```{r}
library(stringr)
v <- "it was the best of times it was the worst of times"
print(v)
w <- stringr::str_replace(v, "times", "spaghetti" )
print(w) ## only first "times" replaced
x <- stringr::str_replace_all(v, "times", "spaghetti" ) 
print(x) ## all "times" replaced
```

## str_sub{#str_sub}

Create substrings based on character index
```{r}
library(stringr)
v <- "it was the best of times it was the worst of times"
print(v)
w <- stringr::str_sub(v, 1, 10)
print(w)
x <- stringr::str_sub(v, 11, 20)
print(x)
```

# dates{#dates}
## as_date{#as_date}
From `lubridate` package. Converts from character to date.

Dates in "YYYY-MM-DD" format don't need additional information.
```{r}
library(lubridate)
a <- as_date("2020-11-01")
print(a)
```

Dates in other formats may need the `format` parameter. See different format options [here](https://epirhandbook.com/en/working-with-dates.html) or run `?strptime`.


```{r}
date2 <- as_date("3/1/22", format = "%m/%d/%y" )
date2
```

Can also convert from date to character.

```{r}
date2_char <- as.character(date2, format = "%A %B %d, %Y")
date2_char
```

## as_datetime{#as_datetime}
Similar to `as_date`, except you can include a time.

```{r}
date3 <- as_datetime("2000-05-09 10:00:00", tz = "EST")
date3
```



# dplyr{#dplyr}

```{r}
library(dplyr)
```

## pipe operator ( %>% ){#pipe}
Use pipe operator to chain commands. 

```{r}
a <- c(9, 16, 25)
b <- a %>% sqrt()
b
```
Commonly used with tibbles. Note that in `dplyr`, you don't need to use quotes for column names.

Example below groups by "site_id" and summarizes the mean NDVI.
```{r}
library(geospaar)
f <- system.file("extdata/FAOSTAT_maize.csv", package = "geospaar")
maize <- read.csv(f)
site_summary <- maize %>% 
  filter(Element == "Production") %>% 
  group_by(Area) %>% 
  summarise(Value)
site_summary
```

You can use `.` to specify in which argument the ` %>% ` should go to.
Let's say you want to take a sample of size 50 from the numbers 1:100.
```{r}
sample_size <- 50
samples <- sample_size %>% sample(1:100, ., replace = F)
```

## mutate{#mutate}
Creates a new column based on calculations you define.
```{r}
set.seed(1)
tib <- tibble(a = 1:10, b = sample(1:100, 10))
tib <- tib %>% mutate(product = a * b) ## new column is product of columns a, b
tib
```



## dplyr::select{#dplyr::select}
```{r}
set.seed(1)
tib <- tibble(a = 1:10, b = sample(1:100, 10))
tib <- tib %>% mutate(product = a * b) ## new column is product of columns a, b
tib
```

## arrange{#arrange}
Sorts by a column. Default is ascending order. You can also arrange multiple columns.
```{r}
set.seed(1)
tib <- tibble(a = 1:10, b = sample(1:100, 10))
tib <- tib %>% mutate(product = a * b) ## new column is product of columns a, b
## sort tib by product column
tib_sorted <- tib %>% arrange(product)
tib_sorted
```
Use `-` for descending order
```{r}
tib_sorted <- tib %>% arrange(-product)
tib_sorted
```



# control structures{#string-functions}
## for loops{#for_loops}

In for loop, you perform the operations once for each item in the iterator. So if the loop starts `for(k in items)` then `items` is the iterator.
```{r}
set.seed(2)
items <- sample (1:100, 5)
for(k in items){
  print(k)
  print(paste0("This value is ", k))
}
```

## if-else{#if_else}
```{r}
items <- sample(LETTERS, 10)

for(k in items){
  print(k)
  if(k %in% c("A", "E", "I", "O", "U")){
    print("vowel")
  } else {
    print("consonant")
  }
}
```

## lapply{#lapply}
`lapply` returns objects in a list.
```{r}
v1 <- 1:5
v2 <- lapply(v1, function(x){
  y <- x^2  ## y will be returned
}) #
print(v2)
```

## sapply{#sapply}
`sapply` returns elements in a vector (when possible)
```{r}
v1 <- 1:5
v2 <- sapply(v1, function(x){
  y <- x^2  ## y will be returned
}) #
print(v2)
```

## apply{#apply}
`apply` works well fro 2D objects.
```{r}
set.seed(3)
v1 <- sample(1:100, 5)
v2 <- sample(1:100, 5)
DF <- data.frame(v1, v2) ## data frame columns will take names of vectors
DF
```

Use index 1 for rows.
```{r}
## index 1 for rows
rowMax <- apply(DF, 1, FUN = max)
rowMax
```

Use index 2 for columns
```{r}
## index 2 for columns
colMax <- apply(DF, 2, FUN = max)
colMax
```


# sampling{#sampling}
## sample{#sample}
`sample` is used for picking samples from a discrete object, like a vector.
```{r}
v1 <- sample(1:100, 5)
v2 <- sample(letters, 5)
print(v1)
print(v2)
```

## runif{#runif}
`runif` samples from a uniform distribution (equal probability for all values in the defined interval)

The example below picks 5 values from a uniform distribution between 0 and 2.
```{r}
set.seed(4)
v <- runif(5, min = 0, max = 2)
v
```

## rnorm{#rnorm}
`rnorm` uses a normal distribution. You can define the mean and standard deviation.
```{r}
set.seed(4)
v <- rnorm(5, mean = 10, sd = 3)
v
```


# read/write{#read_write}
## read/write csv's{#csv}
You can use Base R `read.csv()`, or readr `read_csv()`
```{r}
f <- system.file("extdata/FAOSTAT_maize.csv", package = "geospaar")
maize <- read.csv(f)
print(class(maize))
```
```{r}
maize2 <- readr::read_csv(f)
print(class(maize2))
maize2
```

## save/load{#save_load}
Saving and loading is used for `RData` objects. Use extension `.rda`.
You can save any R object in this way (data frames, tibbles, lists, rasters etc)

```{r}
f <- system.file("extdata/FAOSTAT_maize.csv", package = "geospaar")
maize <- read.csv(f)
save(maize, file = "~/maize.rda") ## save to your user home
```

When you load data, it will retain the variable name it had.
```{r}
maize <- NULL
load(file = "~/maize.rda") ## data will be loaded into "maize" variable
```


# table indexes{#indexing}
## Base R{#index_base_R}
Use `[ , ]` notation. 
Row conditions (filtering) are to the left of comma. 
Column conditions (dplyr::selecting columns) are to the right.
```{r}
DF <- data.frame(v1 = 1:5, v2 = 6:10)
rownames(DF) <- LETTERS[1:5]
DF
```
```{r}
DF[,'v2'] ## column indexing
```

```{r}
DF[c("A", "B", "D"), ] ## row indexing
```

Subsetting data.
```{r}
DF[ DF$v1 > 3   ,    ] ## get observations (rows) where first column is larger than 3
```

## dplyr{#index_dplyr}
Use `filter` for row conditions and `dplyr::select` to dplyr::select columns.

```{r}
DF <- tibble(v1 = 1:5, v2 = 6:10)
rownames(DF) <- LETTERS[1:5]
DF
```

Filter to rows where v1 is greater than 3.
```{r}
DF_filt <- DF %>% filter(v1 > 3)
DF_filt
```

Same as above but only show column v2.

```{r}
DF_filt <- DF %>% filter(v1 > 3) %>% dplyr::select(v2)
DF_filt
```

## slice{#slice}
`slice` is a `dplyr` function to dplyr::select rows by number.

dplyr::select second and third rows.
```{r}
DF_filt <- DF %>% slice(2:3)
DF_filt
```

## head, tail{#head_tail}
`head` dplyr::selects the first n rows in a data frame or tibble. `tail` dplyr::selects the last n rows.

```{r}
DF_head <- DF %>% head(2)
DF_head
```

```{r}
DF_tail <- DF %>% tail(2)
DF_tail
```


# table functions{#tables}
## cbind, rbind{#cbind_rbind}
## joins{#joins}
## pivot_longer{#pivot_longer}
## pivot_wider{#pivot_wider}



# Other{#other}
## which{#which}
Returns indices (position in vector) where a condition is true.
```{r}
set.seed(1)
a <- sample(1:100, 20)
print(a)
print(which(a > 80)) ## shows indices of elements greater than 80. 
```

## which.min{#which_min}
Finds index of minimum value. Only returns first location of min, even if multiple values exist.
```{r}
v <- c(5, 1, 10, 3, 10, 8, 1)
which.min(v) ## only returns index 2, even though 
```

## which.max{#which_max}
Finds index of maximum value. Only returns first location of max, even if multiple values exist.
```{r}
v <- c(5, 1, 10, 3, 10, 8, 1)
which.max(v) ## only returns index 3
```
## unique{#unique}

`unique` filters an object to unique values
```{r}
set.seed(2)
birthdays <- sample(1:365, 50, replace = T) ## sample 100 birthdays
print(birthdays)
```
```{r}
distinct_birthdays <- (unique(birthdays))
print(distinct_birthdays)
print(paste0(length(distinct_birthdays), " distinct birthdays"))
```



# getting help{#help}
## ?{#help_function}
Use `?` to load help for a function
```{r}
#?dplyr::mutate
#?strptime
```




***
<center>
[Back to home](index.html)
</center>
***
