---
title: "Known issues"
subtitle: "GEOG246-346"
date: "`r Sys.Date()` [Back to home](toc.html)"
output:
  rmdformats::readthedown:
    highlight: pygments
    fig_caption: yes
    number_sections: yes
    toc_depth: 3
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Known-issues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
```

# Fixing a pandoc error
The current version of `geospaar` 0.2.4 was built against `R 3.6.2` and RStudio 1.2.5033. Older versions of RStudio may cause a pandoc failure when building vignettes.  Here is a fix (courtesy of Lei Song). If you get this error:

Run the following lines of code (fix suggested here) from an `R` script:


```{r, eval=FALSE}
# Download pandoc 2.7.1 built with ghc-8.6.4, and instruct
# RStudio + rmarkdown to use it.
local({
  # The directory where Pandoc will be extracted. Feel free
  # to adjust this path as appropriate.
  dir <- "~/rstudio-pandoc"
  # The version of Pandoc to be installed.
  version <- "2.7.1"
  # Create and move to the requested directory.
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  owd <- setwd(dir)
  on.exit(setwd(owd), add = TRUE)
  # Construct path to pandoc.
  root <- "https://s3.amazonaws.com/rstudio-buildtools"
  suffix <- sprintf("pandoc-%s-windows-x86_64.zip", version)
  url <- file.path(root, "pandoc-rstudio", version, suffix)
  # Download and extract pandoc.
  file <- basename(url)
  utils::download.file(url, destfile = file)
  utils::unzip(file)
  unlink(file)
  # Write .Renviron to update the version of Pandoc used.
  entry <- paste("RSTUDIO_PANDOC", shQuote(path.expand(dir)), sep = " = ")
  contents <- if (file.exists("~/.Renviron")) readLines("~/.Renviron")
  filtered <- grep("^RSTUDIO_PANDOC", contents, value = TRUE, invert = TRUE)
  amended <- union(filtered, entry)
  writeLines(amended, "~/.Renviron")
  # Report change to the user.
  writeLines("Updated .Renviron:\n")
  writeLines(amended)
  writeLines("\nPlease restart RStudio for these changes to take effect.")
}) 
```

Then restart Rstudio and proceed again with `devtools::install*` steps. This step can be avoided if you update to the latest Rstudio, however. 

***
<center>
[Back to home](toc.html)
</center>
***
