---
title: "Semester Projects"
author: "Lyndon Estes"
date: "`r Sys.Date()` [Back to home](index.html)"
output:
  rmdformats::readthedown:
    highlight: pygments
    number_sections: yes
    toc_depth: 3
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Projects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview{#overview}

Each student will undertake an R-based final project for this class. This project could be one related to your own thesis work, which we will discuss and agree upon. Otherwise, there is a a list of [potential projects below](projects.html#project-list), which are related to ongoing research projects.  

There are two parts to the final project: 1) the project overview; 2) the final project itself. 

# Project overview assignment 

As the first step in the first assignment purpose of this assignment is to provide a plan for your semester final project.

## Set-up
We say goodbye to the repos/packages used for assignments 1-5 (e.g. your version of `xyza1`). Start a new package with a Git/GitHub repo, naming it something short but descriptive of the project you are doing. Follow the R package naming conventions to setup your repo name. This should be a private repo under the agroimpacts organization. 

As with your previous assignments, your work should be structured as an R package. That does not mean that you will necessarily need to be writing functions that are native to the package (i.e. documented functions living in the R folder that build with the package), but you should be writing your project up and showing your code and analysis within vignettes.  

## Tasks

For this assignment, you should write a single vignette called "overview.Rmd", which has the following sections: 

1. Summary: A brief (up to 250 words) description of the project, and a bullet point or enumerated list of its primary objectives. 

2. Approach and Method: An outline of the analytical methods and code you plan to use, including the names of key packages that you will draw on. This section should be composed of the following sub-sections: 

3. Data: A brief (~250 words) description **and** visualization of the datasets you will be using. That means spatial plots of the main datasets and their key values, and, as a bonus, a plot of summary statistics, e.g. a histogram or boxplot of one of the more importants variables in the dataset.  

4. Code: A bullet point summary of the analysis and coding approach that you propose to follow. For teams, this section should include a description of which member will be responsible for each bullet point. 

5. Timelines: Provide a timeline for when each portion of the analysis will be completed. These timelines should be constructed relative to the time period of presentations (during the last two weeks of class) and final project submission (during exam week). For teams, names should be associated with each step on the timeline. 

6. Anticipated outcomes: Briefly describe, as bullet points, the outcomes you expect for each of your primary project objectives

## An additional note for teams
For this last assignment, it doesn't make sense that team members do separate work in different branches of the repo, with the exception of cases where data fetching and wrangling tasks needed to describe and visualize the data are partitioned between members. In this case, team members should have separate vignettes in their respective branches describing the data gathering and processing steps to date. Otherwise, the expectation is that this assignment is presented as a jointly written vignette that builds with the package in the master branch. Please put your initials next to sections that you were responsible for writing. If you want to get really advanced, you can each jointly work on a version of overview in your own branches, and then do a merge and reconcile of your differing version in the master branch. 

***
<center>
[Back to home](index.html)
</center>
***

# Final project
The following are the requirements and assessment approach for the final project:

## Overall scope

The purpose of the final product is to provide an overview of the results of the project you have been working on during the final unit of the class, building on the work you will have already presented in class during one of the last four sessions. The nature of this product can vary according to the type of project you are pursuing, which I would broadly define into two groups: 

1. Those that are more focused on developing a package that provides a set of R functions that will be more broadly useful; 

2. Those that are more analytical in nature, i.e. using R to answer particular questions of interest. 

In the former case, we will want to see a good presentation of worked examples presented in your package vignette(s). In the latter case, your package vignette(s) should give an informative overview of your key findings and (if applicable) way forward for the project. Some projects may fall somewhere in between these two extremes.    

## Structure and content

Regardless of the type of project you are developing, you will submit it as an R package, as we have been doing all semester. That R package should be fully reproducible, i.e.

```{r, eval = FALSE}
devtools::install_github("agroimpacts/yourproject", build_vignettes = TRUE)
browseVignettes("yourproject")
```

Works perfectly and gives us results that contain everything we need to evaluate the project. Datasets that are too large to be committed to the package repo should be linked to, but the code used to process them should be clear, and any figures built from them should be committed to the repo and incorporated in your package vignettes. Vignettes will contain the material we are most interested in, but we will also examine your DESCRIPTION, NAMESPACE, function documentation (if applicable), i.e. the key components of making an R package and delivering it in a reproducible manner. 

Jointly authored vignettes in group projects should have team member initials indicating which section each member worked on which section. Otherwise, separately authored ones should have team member titles at the top.  

As stated before, if there are clear, separable tasks that lend themselves to separate development efforts, team members should work on those and make frequent commits within their own repo branches. 

R code should follow the [style](http://adv-r.had.co.nz/Style.html) we have been using all along. 

The main item in your package that we will assess is/are the package vignette/vignettes. 

## Vignette(s)

You can choose to have one or several vignettes, depending on how much sense it makes to break down the information into separate documents. Regardless, your vignette or vignettes should contain the following information (the names of the headings can vary though according to what makes sense)

### Introduction (or Overview)
A more detailed and informative overview of the project's purpose and objectives than what your provided in assignment 6. Approximate maximum length of 400 (individual) to 800 (team, assuming a single package vignette) words.  

### Methods (or Approach)

For a more analytical project, provide a complete description of the methods you used to undertake the analysis, complete with informative illustrations and summaries of the input datasets. Approximate maximum length of 400 (individual) to 800 words. 

For a project that is more tools-oriented, use this section to describe the purpose and rationale of the functions/capabilities you are developing, what other packages/software/tools it is building on, and provide informative illustrations and summaries of the dataset(s) you are using to demonstrate your functions. Same approximate length. 

### Results (Worked Examples)

For analytical projects, describe and illustrate your results, using standard scientific reporting conventions: 1) plan on having 2-4 figures; 2) a similar number of statistical summaries; 3) describe the results in your figures and tables in (up to) 400-800 words (range refers to individual versus team efforts). 

For tools-oriented projects, provide and illustrate 2-4 worked examples of how to use the functions in your package, with descriptive accompanying text that will help users to understand what the functions do and how to apply them (up to 400-800 words; range refers to individual versus team efforts). 

### Discussion

For analytical projects, briefly provide your interpretation of the results, any uncertainties/difficulties encountered, and any next steps to be taken. Up to 400-800 words (range refers to individual versus team efforts).

For tools-oriented projects, describe any limitations of the package, improvements that can be made, and any plans to undertake these. Up to 400-800 words (range refers to individual versus team efforts).  


Alternatively, you can choose to undertake one of the projects listed below. 

This list is currently in thumbnail form, and will fill out during the next few weeks. 

# Potential project list{#project-list}

## Urban tree health (HERO)
Develop a Shiny app identifying locations of HERO trees and their health over time. Project would potentially cover both HERO trees (across MA, RI) and trees in arboretum. 

[HERO Research](https://www.clarku.edu/departments/hero-program/research/)

## Ecological sensing in MA via in-situ sound sensors
Building on data from Prof. Sangermano's work on sound sensing, this project would examine relationship morning bird chorus and other variables, including MODIS temperature data and forest fragmentation. 

Ideally, project would also create an interactive map that displays sensing data. 

If interested, students can also work with sound processing libraries (seewave, soundecology).

[Flor presentation](https://vimeo.com/762855200)

## Tracking CAFO installation in Yucatan peninsula
We have a data set of vector polygons identifying Concentrated Animal Feeding Operations (CAFOs) in Yucatan Peninsula Mexico. This project will develop a classification model for identifying CAFO locations, and use historical remote sensing imagery to estimate when CAFO's were built.

[Storymap](https://storymaps.arcgis.com/collections/3e7203cf44cf417c9b5fe1db7a182293)

## Game Design

R is not especially used for games but there are some [nice examples](https://appsilon.com/shiny-videogame-video-tutorial/#:~:text=Did%20you%20know%20you%20can,and%20easy%20to%20play%20games.). This would be an educational geospatial game (similar to "Worldle" and "Carmen SanDiego"). The game should give progressive hints that are based off of geospatial analysis (i.e. predominant climate type, percent forest etc). The user could guess the country by picking a location in the app.  


## Redistricting effects
Congressional maps were redrawn in 2022 after the 2020 Census. Where were the largest changes seen? Ideally you would have a background in political science/demographics.

[Redistricting Data Hub](https://redistrictingdatahub.org/)

## COVID policy impact
Create shiny app to examine trends relating COVID cases and different policies implemented (masking, vaccine requirements, social distancing). Can be at state or country level. 

[COVID tracking](https://covid.cdc.gov/covid-data-tracker/#datatracker-home)

## Modeling climate change
Use climate data from GEE ``ee.ImageCollection("NASA/NEX-DCP30")`` to estimate impacts on heat stress, drought and precipitation. Use the ``rgee`` library to create a Shiny app. [rgee link](https://r-spatial.github.io/rgee/reference/rgee-package.html)

[rgee sample app](https://github.com/MVanDenburg92/RGEE_Shiny)

## USF projects (build on previous work)

### Are neighborhoods with public housing more brightly lit than those without? 

Following up on the previous semester's [work](https://github.com/agroimpacts/USFlite/blob/main/external/notebooks/nightlights-part2.md) (and processed further for the Urban Studies Foundation summer institute) that used VIIRS nightlights data to show that New York neighborhoods with public housing are more brightly lit than comparable neighborhoods without, we want to extend this analysis to other cities. 

This will entail collecting data on public housing projects and building footpoints from other US cities, and then collecting and analyzing the nightlights data following the steps in the previous project. 

## Analyzing Ghanaian croplands (build on previous work)

The Agricultural Impacts Research Group has undertaken several projects focussed on Ghainaian croplands over the past several years. These include the [Mapping Africa](http://mappingafrica.io) project, which has applied an integrated crowdsourcing/machine learning platform to map Ghana's croplands at high resolution. Another project involves work with [Farmerline](https://farmerline.co/), a farmer services provider, for the Enabling Crop Analytics at Scale (ECAAS) project, to [develop innovative methods](https://cropanalytics.net/cropland-mapping/) for collecting ground truth on crop types and using these to map crop types. 

There are several possible projects available from these projects. 

From the Mapping Africa data:

### Assess field size and shape through Ghana
Working with a set of cropland predictions at 3 m resolution throughout Ghana, convert the predictions to individual field boundaries using python scripts plus R, and then estimate field sizes throughout the region. Can this dataset be used to identify center pivot irrigation?  

From the ECAAS project:

### Map crop types using Class 3 labels
Building on an analysis that was already, we want to recreate a crop type map for Northern Ghana using a pure R implementation of a 2-step Random Forests analysis: 

- Step 1: Train and test a Random Forests model
- Step 2: Use the trained model to map predictions throughout Northern Ghana
- Step 3: Identify highest likelihood predictions, and convert these to new training labels
- Step 4: Retrain models, map predictions, and assess change between first and second map. 

The code for this is complex but has already been worked out, so this a good example of adapting existing code to perform an analysis. 

### Filter updated crop type predictions through field boundaries
Use a cropland layers from the Mapping Africa project to filter that predicted crop type map so that the predictions are confined to known fields.  

### Compare crop type predictions to drone labels

Working with an extensive set of manually digitized labels captured in from drone imagery over two areas of Ghana, compare the crop type predictions from the Random Forests model to the drone labels.  Calculate model performance metrics, and map the correspondence. 

## Student research
Use R to improve the analysis and visualization of your current research. 

