---
title: "Unit 2 - Working with Raster Data"
author: "Lyndon Estes, Zhiwen Zhu"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: yes
    toc_depth: 3
    toc: yes
vignette: >
  %\VignetteIndexEntry{Unit 2: Raster data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 38px;
  <!-- color: DarkRed; -->
}
h1 { /* Header 1 */
  font-size: 28px;
  <!-- color: DarkBlue; -->
}
h2 { /* Header 2 */
    font-size: 22px;
  <!-- color: DarkBlue; -->
}
h3 { /* Header 3 */
  font-size: 18px;
  <!-- font-family: "Times New Roman", Times, serif; -->
  <!-- color: DarkBlue; -->
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# Introduction{#rasterintro}

The preamble to this section is basically the same as the one for the previous sections on [vector data](vector-data.html), except here we are working with raster data, and will primarily use the `raster` package (which we started using when working with vectors, because of the convenient functions). We will continue to draw on, directly or indirectly (via package dependencies), `sp`, `rgdal`, and `rgeos`.

The material in this section assumes that the reader is familiar with standard GIS operations and concepts, ranging from projections and transformations to moving windows, raster algebra, terrain analysis, and the like. 

As with the previous vector-focused materials, reading are designed with the expectation that you work through the code as you read through it, and do the exercises (as much as you can) before class. Additionally, you should also look at the help file for each new function that is introduced, in order to familiarize yourself with the arguments. 

[Back to top](#rasterintro)

# Preparation
## Working space

Please continue to use the same working space configuration as described in the [vector data](vector-data.html) vignette. 

## Installations

Let's start by loading the following packages.
```{r, warning=FALSE, message=FALSE}
library(geospaar)
library(raster)
```

# Section 1{#section1}

## Exercises{#unit1exercises} 
- [Exercise #1](#exercise1)
- [Exercise #2](#exercise2)
- [Exercise #3](#exercise3)
- [Exercise #4](#exercise4)
- [Exercise #5](#exercise5) 

## Data
```{r}
farmers <- read.csv(system.file("extdata/farmer_spatial.csv", 
                                package = "geospaar"), stringsAsFactors = FALSE)
head(farmers)
```

## Reading, writing, and creating rasters

## Working with 


# Section 2{#section2}

[Back to top](#rasterintro)

# Section 3{#section3}

[Back to top](#rasterintro)

<span style="color:blue">Exercise 1</span><a name="exercise15"></a>: ___
___ [Exercise index](#unit3exercises)















