---
title: "Geography 246/346 Syllabus"
author: "Lyndon Estes, Zhiwen Zhu"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: yes
    toc_depth: 3
    toc: yes
vignette: >
  %\VignetteIndexEntry{Syllabus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 38px;
  <!-- color: DarkRed; -->
}
h1 { /* Header 1 */
  font-size: 28px;
  <!-- color: DarkBlue; -->
}
h2 { /* Header 2 */
    font-size: 22px;
  <!-- color: DarkBlue; -->
}
h3 { /* Header 3 */
  font-size: 18px;
  <!-- font-family: "Times New Roman", Times, serif; -->
  <!-- color: DarkBlue; -->
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Overview

Originally developed for statistics, free and open source R is now one of the most widely used programming languages in both industry and academia. Among its many applications, R is increasingly used for geospatial analyses, as evidenced by its growing ecosystem of spatial libraries (covering everything from basic vector and raster manipulation to advanced modeling and image processing), the evolution of classes for representing spatial data, its expanding capacity to interact with and leverage the capabilities of other geospatial libraries (GDAL) and GIS (e.g. GRASS, QGIS, PostGIS, SAGA).  Alongside its geospatial capabilities, there has been concurrent development of interactive development environments (IDEs) for R, which facilitate the creation, presentation, and reproducibility of analyses. R has therefore very close to being a one-stop shop for the modern GIScientist. This course will provide students with the skills they need to use R as a GIS. There will be additional emphases on programming, presentation, and reproducibility, which will entail learning to develop R libraries, development of presentations and reports using Rmarkdown, and using version control with github. Students will learn and apply R skills by working on a specific research problem. 

<a href="#top">Back to top</a>

## Meeting Time/Place

Jonas Clark 103, MW 12-13.15

## Philosophy
Although our primary focus is learning how to use `R` for geospatial analysis, the course also has a meta-goal, which is to introduce some additional skills and a mindset that together lead to more efficient and transparent, and thus ___reproducible___, research. Reproducibility is something I (Lyndon) only started to figure out (and am still learning) well after I completed by postdoc, but sincerely wish I had acquired before I started my PhD.  

<a href="#top">Back to top</a>

## Caveats/things to consider
Here are some other aspects regarding this course that you may wish to consider before committing: 

1. ___Assignments in this class are problem-oriented, not recipe-based___: 

    I believe that the best way to learn R (and any human or computational language, for that matter) is to figure out how to use it to solve specific problems. There are usually multiple paths that can be taken to solve a problem, particularly in R, which has a huge number of contributors and over 10,000 packages. Coding recipes that spell out the precise steps needed to arrive at a solution will not take you as far towards R fluency, and will prevent you from learning the diversity of this language. 
    
    I find it very rewarding to figure out programming problems (and I even prefer programming to writing papers), but some people can find it boring or even frustrating. If you fall into the latter camp, please consider the possibility that this course might not be for you. Let's discuss.
    
2. ___The order in which material is introduced will occasionally be non-linear___: 

    Primarily because we are introducing the concept of reproducibility up front in this class, which entails learning about some things that people might ordinarily get around to after they know a bit of R code.  However, I believe that this order of things will enable you to learn R (or any other programming language) more rapidly. It will also (hopefully) make learning R more fun. 

3. ___This is a flipped class___:

    All materials and problems are expected to be done before class.  With few exceptions, there will be no lectures. Class time is for helping you to work through problems that you are having, clarifying concepts that are unclear, or for presenting your work to your peers. That means that we expect you to have completed your work ahead of time so that you are already aware of the problems you are having.  
    
    We have adopted this structure because of the problem-oriented nature of assignments. 
    
4. ___I am not a computer scientist___:

    Full disclosure: I am self-taught programmer. That means that the way I teach this course might differ from the way a formally trained computer or (even data) scientist would teach it. I don't know if that's the case, but those of you have taken computer science courses might notice some differences. 
    
5. ___This is a new course___: 

    We are developing this course for Clark for the first time. Much of it will be developed as the semester unfolds, building out material onto the existing three unit framework. This approach will allow us to adapt the pace and material as needed to match progress.  The disadvantage is that you will not see from Day 1 the exact materials we will be covering for later sections (e.g. Unit 2). 

<a href="#top">Back to top</a>
    
## Required Texts, Reading, and Assignments
There are no required texts for this course, at least not ones that you have to buy. There is a huge amount of well-developed R material that is freely available on the web, and we are going to draw on that for this class. You will find links to that material in the syllabus, and, in some cases, in the [Resources](#Resources) section below.

__Readings__ should be completed ___before___ the class they are listed under. E.g. By the time you arrive for the first class on Monday, August 28th, you should have done the readings. Also, since many of the assigned readings present example code and practical work, you should be running those examples as you read. This is a key part of learning the language, and this part of the course is separate from the practicals. 

___Practical assignments___. During the first two units, we will give a number of practical exercises (see links to individual assignments in the [course structure](#structure) section). Unless otherwise stated, these will be assigned on a Monday and due by midnight on Friday of the same week. These assignments will ask you to apply skills from the readings to solve a particular problem. You will undertake and submit your work through github repositories that exist under the Agricultural Impacts Research Group's github [organization](https://github.com/agroimpacts), where there is a [team](https://github.com/orgs/agroimpacts/teams/geog246346) setup for this class. That means that you will need to join github (it's free!) to be able to submit assignments, as we will need your github name to add you to the team. You will manage your individual assignments under private personal repositories that will be listed under your own individual sub-team.   

__Projects__ Each student will be required to undertake a final project. Please see the [projects](projects.html) page for more detail.

A note on writing code: there are many ways to write code and get the results you want. However, not all ways of writing code are equal. Some code is messy and hard to read. Other code is organized, clean, and easy to read. The latter is what we are aiming for, as it helps to foster reproducibility. In this class, we will follow [Hadley Wickham's style guide](http://adv-r.had.co.nz/Style.html). Please study it.    

<a href="#top">Back to top</a>

## Assessment and engaged time

Your progress in this class will be assessed according as follows: 

```{r, echo = FALSE}
# a <- c(20, 30, 40)
# a / sum(a)
mat <- rbind(c("Practical assignments (n=6)", "33% of grade", "22% of grade"), 
            c("Oral presentation of semester project", "22% of grade", 
              "33% of grade"),
            c("Final report on semester project", "45% of grade", 
              "45% of grade"))
colnames(mat) <- c("Component", "GEOG246", "GEOG346")
kable(mat)                 
```

Grading will be based on the rubrics found under the [Assessment](assessment.html) vignette ___[still under construction]___.

Courses at Clark are worth 4 credit hours, which equates to 180 hours of engaged academic time. The breakdown of that time is estimated to be: 

```{r, echo = FALSE}
gr <- c(37, 12, 40, 65, 10, 16)
ug <- c(37, 14, 50, 59, 8, 12)
DF <- cbind.data.frame(c("Class meetings", "Readings", "Practicals", 
                         "Semester projects-analysis", 
                         "Semester projects-presentation",
                         "Semester projects-final report", "Total hours"), 
                       c(gr, sum(gr)), c(ug, sum(ug)))
colnames(DF) <- c("", "GEOG246 hours", "GEOG346 hours")
kable(DF)                 

```

<a href="#top">Back to top</a>

## Expectations
Since our class is a computer lab, no food or drink, I am sad to say (and can't have my tea :( ). Since class is flipped, this is a time for questions and discussion, between us and you, and often between yourselves. However, please keep any conversations low and necessary to the task at hand if they are one-on-one, as we don't want necessary class-wide conversations drowned out.   

Class attendance is expected. It is the primary time in which to get help on understanding reading materials and assignments (see next section on Communications). No late assignments (including presentations and final report) will be accepted, unless you have a documentable emergency that prevents an on-time submission. 

For assignments in the first two units, students can work together to figure out coding problems and to understand the material, but final assignments should reflect each student's own work and coding effort (i.e. not copying code from someone else). For final projects, many, if not most, of the projects will be team-based (2-3 per team, depending on the nature of the assignment) on some if not all of the available topics.

We abide by, and will enforce, the University's policies on plagiarism and cheating. Please familiarize yourself with the University's [policy](http://www2.clarku.edu/offices/aac/integrity.cfm) on academic integrity, particularly section I. 

### Communications
We will conduct class communications via a Slack channel that you should already be invited to. Please don't send emails. We chose to use Slack because it keeps relevant conversations in defined places, rather than scattered across email inboxes. We will have a few channels that are designated for specific kinds of conversations. The \#general channel will be used for course announcements, etc. There will be a specific channel for discussion about each assignment. Any questions you have about the assignment should be asked in those. You can send private individual or group messages that are restricted to those individuals in the conversation. Use this for confidential questions.  

Slack's most obvious characteristic is instant messaging. However, please don't approach it as an instant messenger, at least with respect to Zhiwen and myself and how quickly you want a response from us. While I always strive to be responsive, given workloads and the number of students enrolled, we have to restrict responses to specific days and times. Since we are running this class as a flipped model, the goal is for all questions to be answered during class (which will be more like lab than lecture). Outside of class times, we will have the following physical and Slack (meaning we will respond to Slack messages during those times) office hours, to which any other course-related communications should be confined. 

Instructor | Office | Office hours | Slack hours
---------|------------------|------------------|-------
Lyndon   | Jefferson 107    | Tuesday 1-3 pm   | Wednesday 2-4 pm 
Zhiwen   | Geography Earth Observation and Remote Sensing Lab (EORS) | Thursday 9-11 am | Friday 10 am-12 pm

Also, please feel free to give any ___constructive___ feedback, positive or negative (and it can be an anonymous note, if you are more comfortable), regarding the course or how we are teaching it.  This will be appreciated, and can help make the course better for everyone. There will should also be an opportunity for feedback during a mid-course survey. 

<a href="#top">Back to top</a>

# Course Structure{#structure}
The following is an overview of the course structure. Note that this structure will evolve during the semester, given that this is its first run, and we will adjust material to match the pace of learning. 

## Unit 1. *An introduction to R and related reproducibility skills*

In this first part of the course, we will learn the basics of working with R, starting with non-spatial data. We will also learn some additional skills that foster ___reproducibility___, which can be loosely defined as the ability for you and others to *easily* repeat the steps of your analysis.  including the use of `git` and [github](https://github.com), how to create an R package, and the use of `Rmarkdown` to document and present your analyses. 

This syllabus provides the general subject for each week and class. The detailed readings and assignments for each week and day can be found in the accompanying [Unit 1 vignette](unit1.html), as well as the overall learning goals for the unit.  

<a href="#top">Back to top</a>

### Week 1. Overview of R and Reproducibility
#### Class 1 (28 Aug). The nature and history of R

#### Class 2 (30 Aug). Reproducibility and its components
- [Assignment 1](assignment1.html)

### Week 2. Reproducibility Continued
#### Class 3 (6 Sep). Rmarkdown and more on R packages

### Week 3. R fundamentals

#### Class 4 (11 Sep). R bestiary (types, structures, functions, object-orientation)

#### Class 5 (13 Sep). Basic operations, indexing, and control structures
- [Assignment 2](assignment2.html)

### Week 4. Data preparation and visualization
#### Class 6 (18 Sep). Working with/manipulating data

#### Class 7 (20 Sep). Visualization 

### Week 5. Basic analytics
#### Class 8  (25 Sep). Common analyses: simple statistics and regressions

#### Class 9 (29 Sep). More regression (including non-linear)
- [Assignment 3](assignment3.html)

### Week 6. Unit 1 wrap-up
#### Class 10 (2 Oct). Trends in R

#### Class 11 (4 Oct). Wrap-up

<a href="#top">Back to top</a>

## Unit 2. Handling and analyzing spatial data with R

In this part of the course we will start to learn to use R as a GIS.  The detailed syllabus can be found in the [Unit 2 vignette](unit2.html).

### Week 7. Introduction
#### Class 12 (Oct 9) - Spatial data classes

___[under construction]___

### Week 8 (Oct 16, 18). 
### Week 9 (Oct 23, 25). 
### Week 10 (Oct 30, Nov 1). 
### Week 11 (Nov 6, Nov 8). <!-- Visualization -->
<!-- ### Key takeaways/skills -->

## Unit 3. Projects

### Weeks 12-13. 
Students will spend this time working on their projects, with a particular focus on working with us to identify and trouble-shoot methods. During this period, each student/student-team will give a lightning (1-2 minute) informal presentation of progress that day (or perhaps *that week*, depending on how much time this ends up taking away from actual work).  

### Weeks 14-15. 
These last two weeks will be used to give final presentations, which will cover progress-to-date and plans for finishing the semester project. Presentations will be on the order of 8-12 minutes per person, thus 8-12 slides. 

# Resources {#Resources}
The following are some helpful resources that you may wish to follow-up on as you learn R. 

## Books etc
[The R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)

[Efficient R Programming](https://csgillespie.github.io/efficientR/index.html#building-the-book)

[Swirl Stats](http://swirlstats.com) and the `swirl` package

```{r, eval = FALSE}
install.packages("swirl")
library(swirl)
```


## The intertubes

### Websites

__R general__
    
[CRAN](https://cran.r-project.org), particularly an [Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html) the Task view [Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html)

[R Exercises](http://www.r-exercises.com)

[ROpenSci](https://ropensci.org)

[Kelly Black's R tutorial](http://www.cyclismo.org/tutorial/R/#)

[Creating reproducible examples](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example) (read this before posting to one of the mailing lists)

[R mailing lists/listservs](https://www.r-project.org/mail.html)

[Speeding up R](http://www.noamross.net/blog/2013/4/25/faster-talk.html)

[tidyverse tutorial video](https://www.youtube.com/watch?v=9q7gssUP8UA)

__Advanced analytics__

[Generalized Additive Models](http://multithreaded.stitchfix.com/blog/2015/07/30/gam/)

[Regression Trees, Random Forests, Gradient Boosting](https://www.analyticsvidhya.com/blog/2016/04/complete-tutorial-tree-based-modeling-scratch-in-python/)

[Support Vector Machines](https://eight2late.wordpress.com/2017/02/07/a-gentle-introduction-to-support-vector-machines-using-r/)

[Change point detection](https://www.r-bloggers.com/change-point-detection-in-time-series-with-r-and-tableau/)

__Reproducibility__

[Add citations to an Rmarkdown document](https://www.rdocumentation.org/packages/citr/versions/0.2.0)

[Rmarkdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)

[Tools for Reproducible Research](http://kbroman.org/Tools4RR/)

[manuscriptPackage](https://github.com/jhollist/manuscriptPackage): An R package that provides a package template for ___writing a Latex manuscript using Rmarkdown___, complete with citations. 

__R versus python__

This is a big topic, but I will post some links here as I find them. 

[Why I'll stick with R](https://md.ekstrandom.net/blog/2016/04/using-r/)

__R spatial__
[A short R spatial tutorial](https://pakillo.github.io/R-GIS-tutorial/), with some good examples for connecting to Google Maps. 

### Blogs/tweeps to follow

[R-bloggers](https://www.r-bloggers.com/)

[R-spatial](http://r-spatial.org/)

[\@RLangTip](https://twitter.com/RLangTip)










