---
title: "Appendix"
subtitle: "GEOG246-346"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    fig_caption: yes
    number_sections: yes
    toc_depth: 4
    toc: yes
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Unit 2 Module Appendix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

***
<center>
[Back to **vignette index**](toc.html)
</center>
***

# Module 1 practice answers{#module-1-practice-answers}
## Practice 1
### Questions
1. Answers [here](https://r-spatial.github.io/sf/articles/sf1.html#simple-feature-geometry-types).

2. Assuming the `tibble` has x and y or lat/long coordinates, you apply the function `st_as_sf` with the "coords" argument set to specify which columns contain the x and y coordinates. 

3. `sf::plot` by default plots one panel per variable. You can create a single panel by specifying the variable you want, or by using the `st_geometry` argument to strip out the geometry from object. It also prevent distortions that sometimes occur when overlaying subsequent features on top of a base map. 

4. For a single point you provide an x and y coordinate, otherwise you give an input matrix containing x and y coordinates. A polygon requires that the last point pair in the matrix is the same as the first point pair, to close the polygon. 

### Code
1. 
```{r, eval=FALSE}
farmers %>% st_geometry()
st_geometry(farmers)
```

2. 
```{r, eval=FALSE}
st_crs(farmers) <- st_crs(districts)
# p <- "path/to/your/project/notebooks/data"
p <- "~/Desktop"
st_write(farmers, dsn = file.path(p, "farmers.sqlite"))
rm(farmers)
st_read(file.path(p, "farmers.sqlite"))
```

3. 
```{r, eval = FALSE}
class(roads)
str(roads)
class(districts)
str(districts)
```

4. 
```{r, eval=FALSE}
plot(roads %>% st_geometry(), col = "blue")
```

5. 
```{r, eval=FALSE}
plot(districts %>% select(distName), main = "Zambia Districts")
```

6. 
```{r, eval=FALSE}
pts <- st_multipoint(x = cbind(x = c(27, 28, 29), y = c(-13, -14, -15)))
plot(districts %>% st_geometry(), col = "grey")
plot(pts, add = TRUE, col = "orange", pch = 16)
```

***
<center>
[Back to top](#module-1-practice-answers) || [Back to **vignette index**](toc.html)
</center>
***


## Practice 2
### Questions
1. At least with this example, pretty negligible--well less than 1% mean absolute error.  It might matter more in other places and with other scales though.  The reason `st_area` knows how to estimate areas is because it invokes `lwgeom::st_geod_area`, which calculates a geodetic surface area.  

2. Because for the time being `sf::plot` is a fair bit faster. However, [this recent twitter thread](https://twitter.com/thomasp85/status/1092412825899810817) suggests that may change soon. 

3. By using `mutate` with `cut` that has break values based on those properties. In our example, we found the breaks using `quantile` and different probabilities/percentile levels that creating tertiles of area. 

### Code
1. 
```{r, eval = FALSE}
set.seed(1)
districts %>% sample_n(20) %>% st_area() %>% units::set_units("ha") %>% mean()
```

2. 
```{r, eval = FALSE}
set.seed(1)
roads %>% sample_n(100) %>% st_length() %>% units::set_units("km") %>% mean()
```

3. 
```{r, eval = FALSE}
plot(st_geometry(districts), col = "lightgrey")
set.seed(1)
farmers %>% filter(season == 2) %>% sample_n(200) %>% st_geometry() %>% 
  plot(col = "red", pch = 20, add = TRUE)
```

4. 
```{r, eval = FALSE}
districts %>% st_transform(st_crs(roads)) %>% st_geometry() %>% 
  plot(col = "lightgrey")
roads %>%  
  mutate(length = as.numeric(st_length(.) / 1000)) %>% 
  filter(length > 50 & length < 100) %>% st_geometry() %>% 
  plot(col = "red", pch = 20, add = TRUE)
```

5.
```{r, eval = FALSE}
deciles <- function(x) quantile(x, probs = seq(0, 1, 0.1))
dist_deciles <- districts %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  mutate(acls = cut(area, breaks = deciles(area), include.lowest = TRUE)) %>% 
  group_by(acls) %>% summarize(sum_area = sum(area))  
dist_deciles
#
# #2
cols <- heat.colors(10)
par(mar = rep(0, 4))
plot(st_geometry(dist_deciles), col = cols)
legend(x = "bottomright", pch = 15, col = cols, bty = "n", 
       legend = paste0(1:10, c("st", "nd", "rd", rep("th", 7))))

```

***
<center>
[Back to top](#module-1-practice-answers) || [Back to **vignette index**](toc.html)
</center>
***


### Practice 3
#### Questions
1. It changes features from one type to another (e.g. POLYGON to MULTIPOLYGON), either one specified by the user or the simplest possible common feature, if left unspecified. Casting is sometimes necessary to avoid mixed feature types that cause failures for subsequent operations. 

2. `st_union` runs under the hood of `summarise.sf`, so a `summarize` operation on an `sf` will result in a merged/dissolved set of spatial features. 

3. It affects the order of the fields in the resulting unioned `sf` object--the fields from the object passed to the "x" argument appear first. 

#### Code
1. 
```{r, eval = FALSE}
coords <- cbind("x" = c(27, 27.5, 27.5, 27, 27), 
                "y" = c(-13, -13, -13.5, -13.5, -13))
pol2 <- st_polygon(x = list(coords)) %>% st_sfc %>% st_sf(ID = 1, crs = 4326)

par(mar = rep(0, 4))
plot(st_geometry(districts), col = "grey")
plot(pol2, col = "blue", add = TRUE)
```

2. 
```{r, eval = FALSE}
pol2_int_dists <- st_intersection(pol2, districts)
districts[1] %>% plot(col = "grey", main = NULL)
pol2_int_dists[2] %>% plot(col = rainbow(nrow(pol2_int_dists)), add = TRUE)
```

3. `pol2` is nearly 26 hectares larger
```{r, eval = FALSE}
pol2_int_dists %>% st_area() %>% as.numeric() %>% sum() / 10000 -
  pol2 %>% st_area() %>% as.numeric() %>% sum() / 10000
```

4. 
```{r, eval = FALSE}
st_difference(districts, pol2)[2] %>% plot(col = "grey", main = NULL)
```

5. 
```{r, eval = FALSE}
# Try compare the
set.seed(1)
farmers_alb %>% filter(season == 2) %>%
  sample_n(size = 5) %>% st_buffer(dist = 30000) %>% st_geometry %>% plot()

# With 
set.seed(1)
farmers_alb %>% filter(season == 2) %>%
  st_sample(size = 5) %>% st_buffer(dist = 30000) %>% st_geometry %>% plot()
```

6.
```{r, eval = FALSE}
roads_gt400 <- roads %>% filter(as.numeric(st_length(.)) / 1000 > 400)
par(mar = rep(0, 4))
st_transform(districts, st_crs(roads)) %>% st_geometry %>% 
  plot(col = "grey", main = NULL)
roads_gt400 %>% st_buffer(25000) %>% plot(col = "green", add = TRUE)
```

7.
```{r, eval = FALSE}
par(mar = rep(0, 4))
st_transform(districts, st_crs(roads)) %>% st_geometry %>% 
  plot(col = "grey", main = NULL)
roads_gt400 %>% st_buffer(25000) %>% plot(col = "tan", add = TRUE)
set.seed(1)
roads_gt400 %>% st_buffer(25000) %>% 
  st_sample(size = rep(20, nrow(.)), exact = TRUE) %>% 
  plot(col = "red", pch = 20, add = TRUE)

```
***
<center>
[Back to top](#module-1-practice-answers) || [Back to **vignette index**](toc.html)
</center>
***

# Module 2 practice answers
## Practice 1
### Questions
1. `raster` uses the S4 object-oriented system, where `sf` uses S4. Slots in S4 objects are accessed using the `@` operator. 

2. `brick`, because `raster` only allows you to read and write a single layer.  

3. `stack` organizes multiple rasters that might be stored in separation locations into a single multi-layer object, whereas `brick` requires a single file on disk. `bricks` take longer to create, but are faster to work with once they exist. 

4. The output of `raster`'s vectorization functions are `sp` objects, so you have to convert them to `sf` objects using `st_as_sf`. 

### Code
1. 
```{r, eval=FALSE}
# recreate r, r2, r3
e <- extent(c("xmin" = 27, "xmax" = 29, "ymin" = -16, "ymax" = -14))  
r <- raster(x = e, res = 0.25, crs = crs(districts))
set.seed(1)  
values(r) <- sample(1:100, size = ncell(r), replace = TRUE)  # 3
r2 <- r > 50
r3 <- r
set.seed(1)
values(r3) <- rnorm(n = ncell(r3), mean = 10, sd = 2)

# 
r4 <- r3
set.seed(1)
values(r4) <- runif(n = ncell(r4), 0, 1)
r5 <- r4 > 0.5

s2 <- list(r, r2, r3, r4, r5) %>% stack
names(s2) <- c("r", "r2", "r3", "r4", "r5")
plot(s2)

```

2. We use `tempdir()` here, but you should use your `notebooks/data` folder.
```{r, eval = FALSE}
b <- brick(s2, file = file.path(tempdir(), "b2.tif"))
```

3. 
```{r, eval = FALSE}
zamr3 <- raster(x = extent(districts), crs = crs(districts), res = 0.2)
values(zamr3) <- 1:ncell(zamr3)

farmersr <- farmers %>% distinct(uuid, .keep_all = TRUE) %>% select(x, y) %>% 
  mutate(count = 1) %>% st_as_sf(coords = c("x", "y")) %>% 
  rasterize(x = ., y = zamr3, field = "count", fun = sum)

par(mar = c(0, 0, 1, 4))
districts %>% st_union %>% plot(col = "grey", border = "grey")
plot(farmersr, add = TRUE)
```

4. 
```{r, eval = FALSE}
zamr_alb <- projectRaster(from = zamr, res = 20000, crs = crs(roads), 
                          method = "ngb")
farmersr_alb <- projectRaster(from = farmersr, to = zamr_alb, 
                              method = "bilinear")

par(mar = c(0, 0, 1, 4), mfrow = c(1, 2))
districts %>% st_union %>% plot(col = "grey", border = "grey")
plot(farmersr, add = TRUE)
districts %>% st_transform(crs(roads)) %>% st_union %>% 
  plot(col = "grey", border = "grey")
plot(farmersr_alb, add = TRUE)
```

5. 
```{r, eval = FALSE}
par(mar = c(0, 0, 0, 0))
farmersr %>% rasterToPolygons(dissolve = TRUE) %>% st_as_sf %>% 
  plot(main = NULL)
```

***
<center>
[Back to top](#module-1-practice-answers) || [Back to **vignette index**](toc.html)
</center>
***

## Practice 2
### Questions
1. `cellStats` provides summary statistics over an entire layer; `zonal` calculates statistics within pre-defined zones; `focal` calculates statistics within a moving window. 

2. Use `method = bilinear`.

3. You don't. You have to resample them to a common resolution and extent first. And then you have to `stack` them. 

### Code
1. 
```{r, eval = FALSE}
as.Date("10-11-2017", "%m-%d-%Y")
as.Date("10-11-17", "%m-%d-%y")
as.Date("101117", "%m%d%y")
as.Date("10112017", "%m%d%Y")
lubridate::mdy("10-11-2017")
lubridate::as_date("20171011")
```

2. 
```{r, eval = FALSE}
farmersr2 <- farmers %>% distinct(uuid, .keep_all = TRUE) %>% 
  mutate(count = 1) %>% select(x, y, count) %>% 
  st_as_sf(coords = c("x", "y")) %>% 
  rasterize(., distsr, field = "count", fun = sum)
```

3. 
```{r, eval = FALSE}
zonal(farmersr2, distsr, fun = sum) %>% data.frame %>% 
  subs(distsr, .) %>% plot_noaxes
```

4.
```{r, eval=FALSE}
wmat3 <- matrix(1, nrow = 3, ncol = 3) 
wmat5 <- matrix(1, nrow = 5, ncol = 5) 
fstack <- list(sd3 = focal(x = chirpsz[[20]], w = wmat3, fun = sd), 
               sd5 = focal(x = chirpsz[[20]], w = wmat5, fun = sd),
               max3 = focal(x = chirpsz[[20]], w = wmat3, fun = max), 
               max5 = focal(x = chirpsz[[20]], w = wmat5, fun = max)) %>% stack
fstack %>% plot_noaxes
```

5. 
```{r, eval = FALSE}
chirps_d57 <- chirpsz[[1]] %>% crop(., extent(districts[57, ]))
s <- list(d1 = disaggregate(chirps_d57, fact = 5), 
          d2 = disaggregate(chirps_d57, fact = 5, method = "bilinear")) %>% 
  stack
plot_noaxes(s)
```

6. 
```{r, eval = FALSE}
s <- lapply(list(mean, cv, median), function(x) {
  calc(chirpsz, fun = x)  
}) %>% stack
names(s) <- c("Mean", "CV", "Median")
plot_noaxes(s, nr = 1)
```

***
<center>
[Back to top](#module-1-practice-answers) || [Back to **vignette index**](toc.html)
</center>
***
