---
title: "Unit 2 - Handling and analyzing spatial data with R"
author: "Lyndon Estes, Zhiwen Zhu"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: yes
    toc_depth: 3
    toc: yes
vignette: >
  %\VignetteIndexEntry{Unit 2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 38px;
  <!-- color: DarkRed; -->
}
h1 { /* Header 1 */
  font-size: 28px;
  <!-- color: DarkBlue; -->
}
h2 { /* Header 2 */
    font-size: 22px;
  <!-- color: DarkBlue; -->
}
h3 { /* Header 3 */
  font-size: 18px;
  <!-- font-family: "Times New Roman", Times, serif; -->
  <!-- color: DarkBlue; -->
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Overview{#unit2overview}

In this unit we finally begin to start working with spatial data. We will learn to use R to do the types of analyses you typically associate with traditional GIS software, such as Idrisi/Terrset, ArcGIS, QGIS, and friends. We will also continue to use and build on the basic R programming and reproducibility skills you learned in [Unit 1](unit1.html). 

Three assignments will be given during this unit. You should also finalize the subject of your final project before the mid-point of this unit, and begin to assemble the data you need to undertake the project, and outline the approach you plan to use. The more effort you put into planning (and even beginning the analysis for) your project, the more R skills you will acquire. Why? Because you will be be thinking about how to make R produce solutions for problems you are interested in solving, which requires more exploration and creativity than simply tackling assignments that I give you. Also, since each of you will be doing different projects, there is more potential for peer-to-peer learning, as each of you will learn about and share different aspects of R.

The learning goals for this unit are as follows - by the end of this unit, you should: 

- ___Understand/know___: the strengths and limitations of working with spatial data in R; possible workarounds for these limitations; the differences between the primary spatial object classes; the three external libraries (GDAL, GEOS, and PROJ.4) that R (and many other GIS) uses to provide spatial capabilities; the fact that R's current `sp`-based classes and methods are in the process of being replaced by [simple features](https://r-spatial.github.io/sf/articles/sf1.html) (via the new `sf` package)

- Be able to do ___the following___: perform basic vector operations (intersections, unions, differences) and manipulations (merges/joins); raster operations (algebra, focal analyses, terrain characteristics) and manipulations (aggregate/disaggregate, resample, reproject); convert rasters to vectors and vice versa; visualize spatial data; extract quantities/statistics from spatial data; run simple spatial models and map their predictions; get help on R-sig-geo (i.e. know how to create a reproducible examples)   

___[under construction]___

# Week 7 - Introduction
## Class 12 (Oct 9). Spatial data classes  
+ Required reading: 
    + Read the scanned pdf of Chapter 3 from Applied Spatial Data Analysis with R, which you can get by download from this [link](http://bit.ly/2gb85kl) (please let me know if this doesn't work). This is primarily a reading assignment (but see next bullet point)--focus on learning about the different types of spatial objects and methods, beginning with section 3, and particularly regarding the different ways that spatial objects are structured, building in complexity from the simplest `Spatial` class up through `SpatialPoints*`, then `SpatialLines*`, and then `SpatialPolygons*` for vectors, and how `data.frame`s are associated with each. For rasters, understand the differences between `SpatialGrids*`, `SpatialPixels*`, and `raster` classes. You also want to pay attention to key "slots" in these various S4 classes (e.g. "bbox", "proj4string", "coords", "data") and the methods that are used to query slots (e.g. `bbox`, `proj4string`, `coordinates`, `slot` [note this list of methods is not a 1:1 map to slots]). 
    + There is code and data associated with this chapter. Because there is a lot of it, you are not expected to work through it all. However, you are expected to try at least some of it, particularly where you are building up simple spatial objects (up through `SpatialPoints*`, so 2.3-2.4), and in the section (2.8) on the `raster` package, where you have to read in and query basic properties of the raster (up until the code block containing the `writeStart` function in it on page 55--which you can of course do if you are interested). For speed, it is always easiest to copy, paste, and adapt code. However, that is tough to do out of the pdf. Fortunately, all the code and data that accompany this chapter can be accessed as follows: 

        ```{r, eval = FALSE}
        # change the directory to something you want 
        my_zip_path <- "~/Desktop/"
        url <- "http://www.asdar-book.org/bundles2ed/cm_bundle.zip"
        download.file(url, destfile = paste0(my_zip_path, "cm_bundle.zip"))
        unzip(paste0(my_zip_path, "cm_bundle.zip"), 
        exdir = paste0(my_zip_path, "asdar"))
        setwd(paste0(my_zip_path, "asdar"))
        ```

        `cm_mod.R` in the resulting folder contains the code chunks that correspond to what's printed in the chapter, so you can copy and paste out of there, and you will find the datasets that the chapter works with in the unzipped folder. If for some reason the code above does not work for you, then you can manually download and unzip the files from [here](http://www.asdar-book.org/data2ed.php?chapter=1). 
    
+ Do before class (you will need `raster` for the reading also):
    + install the following packages: `rgdal`, `raster`, `rgeos`

    
# Week 8 - Working with vector data
## Class 13 (Oct 16) 
+ Please read and work through the [vector data vignette](vector-data.html), from the start through Section 1. 

## Class 14 (Oct 18)
+ Read and work through Section 2 in the [vector data vignette](vector-data.html). 
+ There are 6 exercises that should be done in this reading. Please come prepared with the code that you used to do the exercises, as I will randomly select a different person for each exercise to demonstrate how to do it, or provide the answer, whichever is most appropriate.  
+ Assignment 4 (to be released by Friday, 20 October)

# Week 9 - Vectors continued, working with rasters
## Class 15 (Oct 23)
+ Read and work through Section 3 in the [vector data vignette](vector-data.html). 
+ This one has 4 exercises which we will go over in Monday's class, following the same approach as in Class 15.     

## Class 16 (Oct 25) 
+ Read and work through Section 1 in the [raster data vignette](raster-data.html)
+ There are 5 exercises to before class. 

# Week 10 - Rasters continued
## Class 17 (Oct 30) 
+ Read and work through Section 2 in the [raster data vignette](raster-data.html)
+ This section also has 5 pre-class exercises (do just the first 3; exercises 6-8)

## Class 18 (Nov 1)
+ Read and work through Section 3 in the [raster data vignette](raster-data.html) 
+ Do exercises 9-10 from Section 2, and 11-13 in Section 3
+ Assignment 5

# Week 10 - End of Unit 2
## Class 19 (Nov 6)
+ Section 4 in the [raster data vignette](raster-data.html)
+ Do exercises 14-17




