---
title: "Appendix"
subtitle: "GEOG246-346"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    fig_caption: yes
    number_sections: yes
    toc_depth: 4
    toc: yes
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Unit 1 Module Appendix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Answers to practice
## Module 1

## Module 2

## Module 3: Data preparation
### Practice 1
#### Questions
1. A `tibble` is an enhanced `data.frame`. Among other features, it provides more information on data types it contains when printing. It can be treated exactly like a `data.frame` though, in terms of indexing and other operations. 

2. Base methods: `tb_a$a`; `tb_a[["a"]]`; `tb_a[, "a"]`. tidyverse, with `tb_a %>% dplyr::select(a)`. Note that that gives back a `tibble` (or `data.frame`, if you have a `data.frame`). To get a vector back, use `tb_a %>% dplyr::pull(a)` 

3. The data are messy, because the "column headers are values, not variable names" (see [here](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)). We would gather the data in the month rows, setting the month as the key and the mean flu cases as the value. 

4. `inner_join` preserves just the values in x and drops non-matching rows from y. `left_join` fills non-matching values in y with NAs. `right_join` preserves values in y dropping non-matching values in x. `full_join` preserves all non-matching rows in both x and y, filling NAs into non-matched rows. 

#### Code
1. 
```{r, eval = FALSE}
set.seed(1)
dat <- tibble(a = sample(1:10), b = rnorm(10))
td <- "/path/where/you/want/to/write" # REPLACE THIS WITH YOUR OWN!!!!
readr::write_csv(dat, path = file.path(td, "dummy.csv"))
```

2. 
```{r, eval = FALSE}
td <- "/path/where/you/want/to/write" # REPLACE THIS WITH YOUR OWN!!!!
readr::read_csv(file.path(td, "dummy.csv"))
```

3. Recreate `tibble`s first
```{r, eval = FALSE}
# Chunk 13
set.seed(1)
t1 <- tibble(v1 = paste0("N", 1:5), v2 = rnorm(5))
t2 <- tibble(v1 = paste0("N", 1:5), v3 = runif(5))
t3 <- tibble(v1 = paste0("N", 1:7), v4 = sample(1:100, 7))
             # v5 = letters[sample(1:26, 7)])
t4 <- tibble(v1 = paste0("N", c(1:2, 4:7, 11)), 
             v5 = letters[sample(1:26, 7)])
```

Then do joins:
```{r, eval = FALSE}
left_join(t1, t2) %>% left_join(., t3) %>% left_join(., t4)
right_join(t1, t2) %>% right_join(., t3) %>% right_join(., t4)
```

4. 
```{r, eval = FALSE}
left_join(t1, t2) %>% left_join(., t3) %>% left_join(., t4) %>% arrange(v5)
right_join(t1, t2) %>% right_join(., t3) %>% right_join(., t4) %>% 
  arrange(desc(v5))
```

5.
```{r, eval=FALSE}
fs <- dir(system.file("extdata/", package = "geospaar"), pattern = "FAOSTAT", 
          full.names = TRUE)
crops <- lapply(fs, readr::read_csv)
crops_df <- do.call(rbind, lapply(crops, function(x) {
  x %>% dplyr::select(Item, Area, Element, Year, Value) %>% 
    spread(key = Element, value = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = `Area harvested`, prod = Production) %>% 
}))
crop_ylds <- crops_df %>% mutate(yields = prod / harv_area)
crop_ylds <- crop_ylds %>%     
  mutate(country = ifelse(country == "South Africa", "ZAF", country)) %>%  
  mutate(country = ifelse(country == "Zambia", "ZMB", country)) %>% 
  mutate(harv_km2 = harv_area / 100)
```

6. 
```{r, eval=FALSE}
crop_ylds %>% rename(harv_area_km2 = harv_km2)
```

7. 
```{r, eval=FALSE}
my_tb <- tibble(v1 = 1:10, v2 = 11:20) %>% 
  rbind(., tibble(v1 = 11:20, v2 = 21:30)) %>% mutate(v3 = v2^2) %>% 
  arrange(-v3)
```

8.
```{r, eval=FALSE}
my_tb %>% slice(1, 10, 17) %>% dplyr::select(v2, v3)
```

### Practice 2: Analysis
#### Questions
1. `dplyr::filter`
2. `group_by(crop, country, y2k)` is doing the splitting, on crop type, then country, and then year. `summarize(...)` is doing the apply using a `mean`. There is no combine line, as it is implicit. 
3. Chunk 30 adds a `filter` for crop type (selecting out maize), and then simply groups on the *y2k* variable. 
4. It doesn't work when the output of the analysis is not tabular/a list, as with `cor.test` and `lm`.  We can overcome this by 1) creating individual functions that reproduce the component outputs of the analysis (e.g. Chunk 33) and these as a list of functions using `funs` to `summarise_all`, 2) doing the splits outside of the pipeline (e.g. Chunk 36), or 3) using functions such as `do` and `broom::tidy` within the pipeline (e.g. Chunk 39). 

#### Code
1. 
```{r, eval = FALSE}
crop_ylds %>% filter(crop == "sorghum" & country == "ZAF" & year >= 2000)
crop_ylds[crop_ylds$crop == "sorghum" & crop_ylds$country == "ZAF" & 
            crop_ylds$year >= 2000, ]
```

2. 
```{r, eval = FALSE}
crop_ylds %>% filter(crop == "sorghum" & country == "ZAF" & year >= 2000) %>% 
  select(prod, harv_area, yields) %>% summarise_all(funs(mean, sd))
```

3. 
```{r, eval = FALSE}
crop_ylds %>% group_by(crop, country) %>% select(prod, harv_area) %>%
  summarise_all(funs(mean, sd))
```

4. 
```{r, eval = FALSE}
crop_ylds %>% filter(crop == "maize" & country == "ZMB") %>% 
  select(yields, harv_area) %>% cor()

dat <- crop_ylds %>% filter(crop == "maize" & country == "ZMB")
cor.test(dat$harv_area, dat$yields)
```

5. 
```{r, eval = FALSE}
crop_ylds %>% filter(crop == "maize") %>% summarize(mu_yld = mean(yields))
# 2.07 t/ha
```

6. South Africa shows larger yields gains (0.061 t/ha/yr versus 0.03 t/ha/yr)
```{r, eval = FALSE}
summary(lm(yields ~ year, 
           data = crop_ylds %>% filter(crop == "maize" & country == "ZMB")))
summary(lm(yields ~ year, 
           data = crop_ylds %>% filter(crop == "maize" & country == "ZAF")))

crop_ylds %>% filter(crop == "maize" & country == "ZMB") %>% 
  lm(yields ~ year, data = .) %>% summary() 
crop_ylds %>% filter(crop == "maize" & country == "ZAF") %>% 
  lm(yields ~ year, data = .) %>% summary()
```

7. 
```{r, eval = FALSE}
crop_ylds %>% filter(crop != "sorghum") %>% group_by(crop, country) %>% 
  do(prod_ha_lm = lm(yields ~ year, data = .)) %>% 
  broom::tidy(., prod_ha_lm)
```

### Practice 3: Visualization
#### Questions
1. `ggplot2` is built from `grid` graphics, and is based on an underlying visualization philosophy. It builds up graphics objects using the `+` operator, and easily does splits within the data using the "color" argument within `aes` and/or the `facet_grid` function. `graphics` plots can be faster to implement for exploratory analysis, `ggplot2` has more attractive, presentation-grade defaults.
2. Because the syntax used in `graphics` plots is used in many of the plotting functions developed for spatial packages, including newer ones such as `sf` and `stars`. 
3. Each of the three plots takes the axis labels exactly as they are specified to the axis arguments (either as they were specified in the formula in the case of Chunk 40 or to the "x" and "y" arguments in Chunk 41). You can change the names using the "xlab" and "ylab" arguments. 
4. Using "col", "pch", and "cex" arguments. 
5. You have to add the `.` to the "data" argument of `plot`, e.g. `dat %>% plot(y ~ x, data = .)`
6. You get just an empty grey background--`ggplot` won't plot anything without a `geom_*` function added to the `ggplot` object. 

#### Code
1. 
```{r, eval=FALSE}
crop_ylds %>% filter(crop == "sorghum") %>%
  ggplot() + geom_histogram(aes(x = yield), bins = 15) +
  ggtitle("Distribution of sorghum yields")
crop_ylds %>% filter(crop == "sorghum") %>%
  ggplot() + geom_histogram(aes(x = yield), bins = 15, fill = "red") +
  ggtitle("Distribution of sorghum yields")
```

2.
```{r, eval = FALSE}
crop_ylds %>% filter(crop == "wheat" & country == "ZAF") %>% 
  plot(harv_area ~ year, data = ., pch = 16, col = "blue", 
       xlab = "", ylab = "Harvested area (ha)", 
       main = "South Africa wheat (1961-2017)")
crop_ylds %>% filter(crop == "wheat" & country == "ZAF") %>% 
  plot(harv_area ~ year, data = ., pch = 16, type = "l", col = "blue", 
       xlab = "", ylab = "Harvested area (ha)", 
       main = "South Africa wheat (1961-2017)")
```

3. 
```{r, eval=FALSE}
crop_ylds %>% filter(crop == "wheat" & country == "ZAF") %>% 
  ggplot() + geom_point(aes(year, harv_area), col = "blue") + 
  xlab("") + ylab("Harvested area (ha)") +  
  ggtitle("South Africa wheat (1961-2017)")
crop_ylds %>% filter(crop == "wheat" & country == "ZAF") %>% 
  ggplot() + geom_line(aes(year, harv_area), col = "blue") + 
  xlab("") + ylab("Harvested area (ha)") +  
  ggtitle("South Africa wheat (1961-2017)")
```

4. 
```{r, eval=FALSE}
crop_ylds %>% filter(crop == "wheat") %>% 
  ggplot() + geom_line(aes(year, harv_area, color = country)) +
  scale_color_manual(values = c("red", "blue")) + 
  xlab("") + ylab("Harvested area (ha)") +  
  ggtitle("Wheat (1961-2017)")

# extra
crop_ylds %>% filter(crop == "wheat") %>% 
  ggplot() + geom_line(aes(year, log10(harv_area), color = country)) +
  scale_color_manual(values = c("red", "blue")) + 
  xlab("") + ylab("Harvested area (ha)") +  
  ggtitle("Wheat (1961-2017)")
```

5. 
```{r, eval = FALSE}
crop_ylds %>% filter(crop == "wheat" & country == "ZAF") %>% 
  ggplot() + geom_point(aes(year, harv_area)) +
  geom_smooth(aes(year, harv_area)) +
  xlab("") + ylab("Harvested area (ha)") +  
  ggtitle("South African wheat (1961-2017)")
```

6. 
```{r}
# ggplot2
crop_ylds %>% filter(crop == "wheat" & country == "ZMB") %>% 
  ggplot() + 
  geom_histogram(aes(x = harv_area), bins = 10, col = "black", fill = "blue") +
  xlab("Harvested area (ha)") + ggtitle("Zambian Wheat (1961-2017)")

# hist
# with dplyr
crop_ylds %>% filter(crop == "wheat" & country == "ZMB") %>% pull(harv_area) %>% 
  hist(., main = "Zambian Wheat (1961-2017)", xlab = "Harvested area (ha)", 
       col = "blue")
# with base subsetting
hist(crop_ylds$harv_area[crop_ylds$crop == "wheat" & 
                           crop_ylds$country == "ZMB"], 
     main = "Zambian Wheat (1961-2017)", xlab = "Harvested area (ha)", 
     col = "blue")
```

7. 
```{r}
crop_ylds %>% filter(country == "ZAF") %>% 
  ggplot() + geom_point(aes(x = year, y = harv_area)) +
  geom_smooth(aes(x = year, y = harv_area)) + 
  facet_grid(cols = vars(crop)) +
  scale_color_manual(values = c("red", "blue")) + 
  ylab("Yield (tons/ha)") + xlab("")
```

