---
title: "Appendix"
subtitle: "GEOG246-346"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    fig_caption: yes
    number_sections: yes
    toc_depth: 4
    toc: yes
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Unit 1 Module Appendix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Answers to practice
## Module 1

## Module 2

## Module 3
### Practice 1
#### Questions
1. A `tibble` is an enhanced `data.frame`. Among other features, it provides more information on data types it contains when printing. It can be treated exactly like a `data.frame` though, in terms of indexing and other operations. 

2. Base methods: `tb_a$a`; `tb_a[["a"]]`; `tb_a[, "a"]`. tidyverse, with `tb_a %>% dplyr::select(a)`. Note that that gives back a `tibble` (or `data.frame`, if you have a `data.frame`). To get a vector back, use `tb_a %>% dplyr::pull(a)` 

3. The data are messy, because the "column headers are values, not variable names" (see [here](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)). We would gather the data in the month rows, setting the month as the key and the mean flu cases as the value. 

4. `inner_join` preserves just the values in x and drops non-matching rows from y. `left_join` fills non-matching values in y with NAs. `right_join` preserves values in y dropping non-matching values in x. `full_join` preserves all non-matching rows in both x and y, filling NAs into non-matched rows. 

#### Code
1. 
```{r, eval = FALSE}
set.seed(1)
dat <- tibble(a = sample(1:10), b = rnorm(10))
td <- "/path/where/you/want/to/write" # REPLACE THIS WITH YOUR OWN!!!!
readr::write_csv(dat, path = file.path(td, "dummy.csv"))
```

2. 
```{r, eval = FALSE}
td <- "/path/where/you/want/to/write" # REPLACE THIS WITH YOUR OWN!!!!
readr::read_csv(file.path(td, "dummy.csv"))
```

3. Recreate `tibble`s first
```{r, eval = FALSE}
# Chunk 13
set.seed(1)
t1 <- tibble(v1 = paste0("N", 1:5), v2 = rnorm(5))
t2 <- tibble(v1 = paste0("N", 1:5), v3 = runif(5))
t3 <- tibble(v1 = paste0("N", 1:7), v4 = sample(1:100, 7))
             # v5 = letters[sample(1:26, 7)])
t4 <- tibble(v1 = paste0("N", c(1:2, 4:7, 11)), 
             v5 = letters[sample(1:26, 7)])
```

Then do joins:
```{r, eval = FALSE}
left_join(t1, t2) %>% left_join(., t3) %>% left_join(., t4)
right_join(t1, t2) %>% right_join(., t3) %>% right_join(., t4)
```

4. 
```{r, eval = FALSE}
left_join(t1, t2) %>% left_join(., t3) %>% left_join(., t4) %>% arrange(v5)
right_join(t1, t2) %>% right_join(., t3) %>% right_join(., t4) %>% 
  arrange(desc(v5))
```

5.
```{r, eval=FALSE}
fs <- dir(system.file("extdata/", package = "geospaar"), pattern = "FAOSTAT", 
          full.names = TRUE)
crops <- lapply(fs, readr::read_csv)
crops_df <- do.call(rbind, lapply(crops, function(x) {
  x %>% dplyr::select(Item, Area, Element, Year, Value) %>% 
    spread(key = Element, value = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = `Area harvested`, prod = Production) %>% 
}))
crop_ylds <- crops_df %>% mutate(yields = prod / harv_area)
crop_ylds <- crop_ylds %>%     
  mutate(country = ifelse(country == "South Africa", "ZAF", country)) %>%  
  mutate(country = ifelse(country == "Zambia", "ZMB", country)) %>% 
  mutate(harv_km2 = harv_area / 100)
```

6. 
```{r}
crop_ylds %>% rename(harv_area_km2 = harv_km2)
```

7. 
```{r, eval=FALSE}
my_tb <- tibble(v1 = 1:10, v2 = 11:20) %>% 
  rbind(., tibble(v1 = 11:20, v2 = 21:30)) %>% mutate(v3 = v2^2) %>% 
  arrange(-v3)
```

8.
```{r, eval=FALSE}
my_tb %>% slice(1, 10, 17) %>% dplyr::select(v2, v3)
```

