---
title: "Appendix"
subtitle: "GEOG246-346"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    fig_caption: yes
    number_sections: yes
    toc_depth: 4
    toc: yes
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Unit 2 Module Appendix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Answers to practice
## Module 1
### Practice 1
#### Questions
1. Answers [here](https://r-spatial.github.io/sf/articles/sf1.html#simple-feature-geometry-types).

2. Assuming the `tibble` has x and y or lat/long coordinates, you apply the function `st_as_sf` with the "coords" argument set to specify which columns contain the x and y coordinates. 

3. `sf::plot` by default plots one panel per variable. You can create a single panel by specifying the variable you want, or by using the `st_geometry` argument to strip out the geometry from object. 

#### Code
1. 
```{r, eval=FALSE}
farmers %>% st_geometry()
st_geometry(farmers)
```

2. 
```{r, eval=FALSE}
st_crs(farmers) <- st_crs(districts)
# p <- "path/to/your/project/notebooks/data"
p <- "~/Desktop"
st_write(farmers, dsn = file.path(p, "farmers.sqlite"))
rm(farmers)
st_read(file.path(p, "farmers.sqlite"))
```

3. 
```{r, eval = FALSE}
class(roads)
str(roads)
class(districts)
str(districts)
```

4. 
```{r, eval=FALSE}
plot(roads %>% st_geometry(), col = "blue")
```

5. 
```{r, eval=FALSE}
plot(districts %>% select(distName), main = "Zambia Districts")
```

6. 
```{r, eval=FALSE}
pts <- st_multipoint(x = cbind(x = c(27, 28, 29), y = c(-13, -14, -15)))
plot(districts %>% st_geometry(), col = "grey")
plot(pts, add = TRUE, col = "orange", pch = 16)
```

## Module 2

## Module 3
