---
title: "Week 2"
output: 
  learnr::tutorial:
    theme: united
    highlight: tango
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
gradethis::gradethis_setup()
tutorial_options(exercise.completion = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

## Class 3

### Quiz

```{r quiz_c1}
quiz(
  question('What does `switch("x", x = , y = 2, z = 3)` return?',
    answer("`NULL`", 
           message = "In `switch`, right hand side = empty and the input will “fall through” to the next value."),
    answer("`2`", 
           correct = T),
    answer("`3`", message = "In `switch`, right hand side = empty and the input will “fall through” to the next value."),
    answer("Syntax error", 
           message = "In `switch`, right hand side = empty and the input will “fall through” to the next value."),
    correct = "In `switch`, right hand side = empty and the input will “fall through” to the next value.",
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question('Which function is used to exist the entire `for` loop?',
    answer("`break`", 
           correct = T),
    answer("`next`", 
           message = "`next` exits the current iteration."),
    answer("`exit`", message = "Not a commonly used function in R."),
    answer("`quit`", 
           message = "`quit` terminate the current R session."),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question('What are the three components of a function?',
    answer("body", 
           correct = T),
    answer("arguments", 
           correct = T),
    answer("environment", 
           correct = T),
    answer("name", 
           message = "Anonymous function have no name."),
    correct = 'The three components of a function are its body, arguments, and environment.',
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question('What you will get?\n```
  a <- 15; b <- 12
  f <- function(a) {a * 10 + b}
  f(20)',
    answer("`212`", 
           correct = T),
    answer("`162`",
           message = "Function `f` will take the input `a` value, and then search level up for b in global environment."),
    answer("`312`",
           message = "Function `f` will take the input `a` value, and then search level up for b in global environment."),
    answer("Syntax error.", 
           message = "Function `f` will take the input `a` value, and then search level up for b in global environment."),
  correct = 'Here uses two rules: name masking and searching level up.',
    random_answer_order = TRUE, allow_retry = TRUE
  )
)
```

## Class 4

### Quiz

```{r quiz_c2}
quiz(
  # reference to https://learn.co/lessons/git-github-learn-quiz
  question("Which function we should use to make all simulation in R reproducible?",
    answer("`set.seed()`", correct = TRUE),
    answer("`seed()`"),
    answer("`random.seed()`"),
    answer("`make.seed()`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to download a repository from GitHub to your local machine freshly?",
    answer("`git clone`", correct = TRUE),
    answer("`git fork`", message = '`git fork` is a not a git command'),
    answer("`git push`"),
    answer("`git pull`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Which command creates an empty Git repository in the specified directory?",
    answer("`git init`", correct = TRUE),
    answer("`git log`"),
    answer("`git reset`"),
    answer("`git config`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to upload your changes to GitHub?",
    answer("`git push`", correct = TRUE),
    answer("`git add`"),
    answer("`git status`"),
    answer("`git clone`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to check the state of local repo since last commit?",
    answer("`git status`", correct = TRUE),
    answer("`git add`"),
    answer("`git commit`"),
    answer("`git diff`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to stage files for a commit?",
    answer("`git add`", correct = TRUE),
    answer("`git commit`"),
    answer("`git status`"),
    answer("`git reset`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to save the current state of your repo into git?",
    answer("`git commit`", correct = TRUE),
    answer("`git add`"),
    answer("`git init`"),
    answer("`git diff`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the shortcut to stage all the changes that you made?",
    answer("`git add .`", correct = TRUE),
    answer("`git add all`"),
    answer("`git commit .`"),
    answer("`git commit all`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("How do you supply a commit message to a commit?",
    answer('`git commit -m "Commit message"`', correct = TRUE),
    answer('`git commit "Commit message"`'),
    answer('`git add -m "Commit message"`'),
    answer('`git add "Commit message"`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("How do you commit for all changes with a message?",
    answer('`git commit -am "Commit message"`', correct = TRUE),
    answer('`git commit -a "Commit message"`'),
    answer('`git add -am "Commit message"`'),
    answer('`git add -a "Commit message"`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("When you commit changes to git, which one comes first: staging with `git add` or committing with `git commit`?",
    answer('Staging with `git add`', correct = TRUE),
    answer('Committing with `git commit`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to unstage a currently staged file called geospaar.txt?",
    answer('`git checkout -- geospaar.txt`', correct = TRUE),
    answer('`git reset geospaar.txt`'),
    answer('`git HEAD^ -- geospaar.txt`'),
    answer('`git unstage -- geospaar.txt`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to make a new branch?",
    answer('`git branch`', correct = TRUE),
    answer('`git -b`'),
    answer('`git new branch`'),
    answer('`git checkout branch`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to switch to another branch?",
    answer('`git branch`'),
    answer('`git -b`'),
    answer('`git new branch`'),
    answer('`git checkout branch`', correct = TRUE),
    random_answer_order = TRUE, allow_retry = TRUE
  )
)
```

