---
title: "Week 3"
output: 
  learnr::tutorial:
    theme: united
    highlight: tango
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.completion = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

## Class 5 challenge
### Introduction

This dataset was build by augmenting datasets of rainfall, climate and fertilizer data available for India ([link](https://www.kaggle.com/atharvaingle/crop-recommendation-dataset)).
We use this dataset to practice how to use `dplyr` to clean, modify, and summarize dataset. The packages or functions that are recommended to use are:

- Package `readr` and its function `read_csv` to read data. Install the package if necessary.
- `group_by` function in `dplyr` to group data and then `summarise` or more flexible `summarise_at` to do something to the groups.
- `select` function in `dplyr`  to subset columns, and `filter` function to subset rows.
- `arrange` function in `dplyr` to order the table based on column(s). `desc` function can be used inside of `arrange` to arrange decreasingly.
- `mutate` function in `dplyr` to add a new column to a dataset. Use with `ifelse` might have some surprising effect.
- `slice` function to index rows.
- `pull` function in `dplyr` to read a column as a vector in pipeline.

Read Unit1-Module4 "Data preparation" and the help documentation of each function to learn how to use them.

### Tasks
Create an R markdown file for these tasks.

#### Task 1

1. Install (if in need) and load necessary packages.
1. Click the link to download the csv file [Crop_recomendation.csv](https://www.kaggle.com/atharvaingle/crop-recommendation-dataset/download) and then read it into working environment as object `crop_rmd` using `readr::read_csv`. Hint: set the right data path. One of the safe way is to use the full path.
2. List all crops growing in this area. Hint: which function we learned to clean duplicated elements?

You are expected to get crop types first and assign to an object `crops`. And then print out them and how many of them using **inline code**. 

*Expected result:*

```{r, warning=F, message=F}
# Task 1
# install.packages('readr')
library(readr)
library(dplyr)

# Load data
crop_rmd <- read_csv('./Crop_recommendation.csv')
crops <- crop_rmd %>% pull(label) %>% unique()
```

There are `r length(crops)` crops in this region: `r crops`.

#### Task 2

Let's focus on maize:

1. Just select out maize.
2. Calculate the mean values for N, P, K, temperature, humidity, ph, and rainfall.
3. Directly print out the result.

There might be multiple methods to do this work. 

*Expected result:*

```{r}
# Task 2
crop_rmd %>% 
  filter(label == 'maize') %>% 
  dplyr::select(-label) %>% 
  colMeans()
```

#### Task 3

Have a look at papaya this time:

1. Only select out papaya.
2. Add a new column `n_level` and sign values based on the rule: 

   - "low", N <= 45, 
   - "median", N > 45 and N <= 55
   - "high", N > 45
  
    Hint: use nested `ifelse`.
  
3. Then just select out the rows that `n_level` is "median", and rainfall >= 150, and temperature < 30. Hint: which operator is for logical and?
4. Just select rainfall, temperature and humidity.
5. Random select one row out using seed 10.
6. Assign the result to object `papaya` and print it out. Hint: use function `sample_n`.

*Expected result:*

```{r}
set.seed(10)
papaya <- crop_rmd %>% 
  filter(label == 'papaya') %>% 
  mutate(n_level = ifelse(N <= 45, 'low', ifelse(N <= 55, 'median', 'high'))) %>% 
  filter(n_level == 'median' & rainfall >= 150 & temperature < 30) %>% 
  dplyr::select(rainfall, temperature, humidity) %>% 
  sample_n(1)
papaya
```

#### Task 4

Let's get some crop that might tolerate heat and drought:

1. Summarize the table to get the mean, max, and min value of temperature, humidity, ph, and rainfall for each crop type.
2. Arrange the table based on mean rainfall increasingly and max temperature decreasingly.
3. Pick up the first 10 crop types.
4. Assign the result to object `crop_hd` and print it out.

*Expected result:*

```{r}
# Task 4
crop_hd <- crop_rmd %>% 
  group_by(label) %>% 
  summarise_at(.vars = c('temperature', 'humidity', 
                         'ph', 'rainfall'),
               .funs = c(mean = "mean",
                         max = "max",
                         min = "min")) %>% 
  arrange(rainfall_mean, desc(temperature_max)) %>% 
  slice(1:10)
crop_hd
```

## Class 6 challenge
### Introduction

We use the same dataset as class 5 challenge to practice how to use `*apply` family:

- `apply()`
- `lapply()`
- `sapply()`
- `tapply()`
- `mapply()`

You have to use `*apply` functions to finish all tasks, except that you might need to use `dplyr` to subset the dataset. In order to let you practice how to use these functions, as well as how to pick up the right function, I won't mention which function you should use. 

### Task

Write an R markdown file for the tasks. Read the dataset (`crop_rmd`) the same as class 5 challenge.

```{r, warning=F, message=F}
# Task 1
# install.packages('readr')
library(readr)
library(dplyr)

# Load data
crop_rmd <- read_csv('./Crop_recommendation.csv')
```

#### Task 1

1. Select watermelon
2. Calculate the `mean` of all items (temperature, rainfall, etc.). Assign the result to `wml_mean` and print out.
4. What if I want to add a random noise from `rnorm(mean = 0, sd = 2)` to these means? Assign the result to `wml_noisy` and print out. Use seed 10 to for the noises. Hint: use anonymous function within a `*apply` function.

*Expected result:*

```{r}
watermelon <- crop_rmd %>% 
  filter(label == 'watermelon') %>% 
  dplyr::select(-label)
wml_mean <- apply(watermelon, 2, mean)
wml_mean
set.seed(10)
wml_noisy <- apply(watermelon, 2, function(x) mean(x) + rnorm(1, 0, 2))
```

#### Task 2

Still for watermelon, I want to calculate $(temperature - 20) / (rainfall - 100)$, what is the easiest way to do that using `*apply`?

*Expected result:*

```{r}
mapply(function(x, y){
  (x - 20) / (y - 100)
}, watermelon$temperature, watermelon$rainfall)
```

#### Task 3

Loop over crop types to calculate max of each type of observations. You are expected to get a matrix back and assign to object `crop_smr`. Hint: You might need function `unique`, and write anonymous function and use `dplyr` within this anonymous function.

*Expected result:*

```{r}
crop_smr <- sapply(unique(crop_rmd$label), function(crop_type){
  crop_rmd %>% 
    filter(label == crop_type) %>% 
    dplyr::select(-label) %>% 
    colMeans()
})
crop_smr
```

#### Task 4

What is the most convenient way to calculate the max value of temperature for each crop types?

*Expected result:*

```{r}
tapply(crop_rmd$temperature, crop_rmd$label, max)
```

