---
title: "Week 2"
output: 
  learnr::tutorial:
    theme: united
    highlight: tango
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
# gradethis::gradethis_setup()
tutorial_options(exercise.completion = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

## Class 3

### Quiz

```{r quiz_c1}
quiz(
  question('What does `switch("x", x = , y = 2, z = 3)` return?',
    answer("`NULL`", 
           message = "In `switch`, right hand side = empty and the input will “fall through” to the next value."),
    answer("`2`", 
           correct = T),
    answer("`3`", message = "In `switch`, right hand side = empty and the input will “fall through” to the next value."),
    answer("Syntax error", 
           message = "In `switch`, right hand side = empty and the input will “fall through” to the next value."),
    correct = "In `switch`, right hand side = empty and the input will “fall through” to the next value.",
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question('Which function is used to exist the entire `for` loop?',
    answer("`break`", 
           correct = T),
    answer("`next`", 
           message = "`next` exits the current iteration."),
    answer("`exit`", message = "Not a commonly used function in R."),
    answer("`quit`", 
           message = "`quit` terminate the current R session."),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question('What are the three components of a function?',
    answer("body", 
           correct = T),
    answer("arguments", 
           correct = T),
    answer("environment", 
           correct = T),
    answer("name", 
           message = "Anonymous function have no name."),
    correct = 'The three components of a function are its body, arguments, and environment.',
    random_answer_order = TRUE, allow_retry = TRUE
  )
)
```

### In-class practice

 - Define a function to convert Fahrenheit to Celsius.

$$Celsius = (Fahrenheit - 32) / 1.8$$

- Exam a sequence of temperature in Fahrenheit `c(70, 34, 25, 10, 89, 92, 15)`, when the temperature in Celsius is lower than -10, give a warning "Stay inside." and print out this temperature in Celsius.

```{r temp, exercise=TRUE, exercise.lines = 18}
# Define the function
c_to_f <- function(fah = 0){
  
}

# Initialize the temperature vector
fahs <- c(70, 34, 25, 10, 89, 92, 15)

# Loop over the temperature to check
for (fah in fahs){
  
  
  
  
  
}
```

The result should be:

```{r temp_result, echo = F, eval = T}
# Define the function
c_to_f <- function(fah = 0){
  # This is a function to convert Fahrenheit to Celsius.
  # The argument is fah with a default value 0.
  (fah - 32) / 1.8
}

# Initialize the temperature vector
fahs <- c(70, 34, 25, 10, 89, 92, 15)

# Loop over the temperature to check
for (fah in fahs){
  cel <- c_to_f(fah)
  if (cel < -10){
    print('Stay inside.')
    print(cel)
  }
}
```

```{r temp-solution, exercise.reveal_solution = T}
# Define the function
c_to_f <- function(fah = 0){
  # This is a function to convert Fahrenheit to Celsius.
  # The argument is fah with a default value 0.
  (fah - 32) / 1.8
}

# Initialize the temperature vector
fahs <- c(70, 34, 25, 10, 89, 92, 15)

# Loop over the temperature to check
for (fah in fahs){
  cel <- c_to_f(fah)
  if (cel < -10){
    print('Stay inside.')
    print(cel)
  }
}
```

### Exercise

Fill up the R code chunk below required to:

1. Make a data.frame `df` with 50 rows by defining columns:
   - Column `a` is a vector of random integers from `1:100`.
   - Column `b` is a vector of random numbers from a uniform distribution.
   - Column `g` is a vector of random elements in `r paste0('group', 1:5)`.
2. Once got the `df`, fo the following:

   - Define a function to mimic function `dimnames`.
   - Define a function to convert column `a` of `df` from number to level based on the rules: A: 100 >= a > 90, B: 90 >= a > 60, C: a 60 >= a > 0.
   - Define a function to check a character is equal to 'group5' or not. Then use for loop function to check each element in column `g` of `df`.
   - Define a function to:
  
     1. Check if column `a` of `df` has any value larger than 50.
     2. And if column `b` of `df` has all values larger than 0.
     3. If both yes, then return the sum of `a * b`.
     
Hints: 

1. The place with "`___`" is the place for you to fill in code. So double click "`___`" and replace it with your code.
1. Use the syntax `df$a` to get an column out from a data.frame as a vector. 
1. If you don't know how to use a function, use `?function_name` to open the help document to learn how to use it.
1. Pay attention to the hints and requirements provided in the code chunk.

```{r df, exercise=TRUE, exercise.lines = 80}
# Define number of row
num_row <- ___ # a number

# Make a
set.seed(10)
a <- sample(___, num_row) # a vector

# Create b
set.seed(10)
b <- runif(___, min = 0, max = 2) # an object

# Create g
g <- ___('group', sample(1:5, 50, replace = T)) # a function

# Combine them to make the df
df <- data.frame(a = ___, b = ___, g = ___) # objects
df

## Define a function to mimic function dimnames
### Could use any functions related to attributes
dimnames_mimic <- function(df){
  # A function to mimic dimnames
  # Example: dimnames_ls(df)
  nms_row <- ___(df) # a function
  nms_col <- ___ # call a function
  
  list(___, ___) # objects
}

dimnames_mimic(df) # should be the same as dimnames(df)

## Define a function to convert column a of df from number to level
## based on the rules: A: 100 >= a > 90, B: 90 >= a > 60, C: a 60 >= a > 0
## Question 1: ifelse or if-else? why? Think about the output shape.
## Hint 2: use df$a to get the a column from df
convert_level <- function(___){ # an object name
  # A function to convert numbers to levels
  # Example: convert_level(1:100)
  ifelse(x > 90 & x <= 100, ___, # a character
         ifelse(___, 'B', # a checking condition
                ifelse(x > 0, 'C', NA)))
}

levels_df <- convert_level(___); levels_df # a vector

## Define a function to check a character is equal to 'group5' or not.
## Then use for loop function to check each element in column g of df.
## Requirement 1: use switch for the check function
check_group <- function(x){
  # A function to check if a character is equal to 'group5'
  # Example: check_group('group3')
  switch(x,
         group1 = ___, # could use "fall through"
         ___ = ___, 
         ___ = ___,
         ___ = 'No',
         group5 = 'Yes', 
         stop('Not valid `x` value.'))
}

# for loop
for (i in ___){ # a vector
  i_check <- ___ # call a function
  print(i_check)
}

## Define a function to:
## 1. Check if column a of df has any value larger than 50
## 2. And if column b of df has all values larger than 0.
## If both yes, then return the sum of a * b
sum_a_b <- function(a, b){
  if (___(a > 50) & (!___(b <= 0))){ # function
    sum(a ___ b) # an operator
  } else{
    # always good to catch other conditions
    print('a or b does not match the condtion.')
  }
}
```

```{r df-solution, exercise.reveal_solution = T}
# Define number of row
num_row <- 50

# Make a
set.seed(10)
a <- sample(1:100, num_row)

# Create b
set.seed(10)
b <- runif(num_row, min = 0, max = 2)

# Create g
g <- paste0('group', sample(1:5, 50, replace = T))

# Combine them to make the df
df <- data.frame(a = a, b = b, g = g)
df

## Define a function to mimic function dimnames
### Could use any functions related to attributes
dimnames_ls <- function(df){
  # A function to mimic dimnames
  # Example: dimnames_ls(df)
  nms_row <- rownames(df) # or row.names(df) or attr(df, 'names')
  nms_col <- colnames(df) # or names(df) or attr(df, 'row.names')
  
  list(nms_row, nms_col)
}

dimnames_ls(df) # should be the same as dimnames(df)

## Define a function to convert column a of df from number to level
## based on the rules: A: 100 >= a > 90, B: 90 >= a > 60, C: a 60 >= a > 0
## Hint 1: ifelse or if-else? why? Think about the output shape
## Hint 2: use df$a to get the a column from df
convert_level <- function(x){
  # A function to convert numbers to levels
  # Exampple: convert_level(1:100)
  ifelse(x > 90 & x <= 100, 'A',
         ifelse(x > 60, 'B',
                ifelse(x > 0, 'C', NA)))
}

levels_df <- convert_level(df$a); levels_df

## Define a function to check a character is equal to 'group5' or not.
## Then use for loop function to check each elemnet in column g of df.
## Requirement 1: use switch for the check function
check_group <- function(x){
  # A function to check if a character is equal to 'group5'
  # Example: check_group('group3')
  switch(x,
         group1 = , # could use "fall through"
         group2 = ,
         group3 = ,
         group4 = 'No',
         group5 = 'Yes',
         stop('Not valid `x` value.'))
}

# for loop
for (i in df$g){
  i_check <- check_group(i)
  print(i_check)
}

## Define a function to:
## 1. Check if column a of df has any value larger than 50
## 2. And if column b of df has all values larger than 0.
## If both yes, then return the sum of a * b
sum_a_b <- function(a, b){
  if (any(a > 50) & (!all(b <= 0))){
    sum(a * b)
  } else{
    # always good to catch other conditions
    print('a or b does not match the condtion.')
  }
}
```

## Class 4

### Quiz

```{r quiz_c2}
quiz(
  # reference to https://learn.co/lessons/git-github-learn-quiz
  question("Which function we should use to make all simulation in R reproducible?",
    answer("`set.seed()`", correct = TRUE),
    answer("`seed()`"),
    answer("`random.seed()`"),
    answer("`make.seed()`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question('What you will get?\n```
  a <- 15; b <- 12
  f <- function(a) {a * 10 + b}
  f(20)',
    answer("`212`", 
           correct = T),
    answer("`162`",
           message = "Function `f` will take the input `a` value, and then search level up for b in global environment."),
    answer("`312`",
           message = "Function `f` will take the input `a` value, and then search level up for b in global environment."),
    answer("Syntax error.", 
           message = "Function `f` will take the input `a` value, and then search level up for b in global environment."),
  correct = 'Here uses two rules: name masking and searching level up.',
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to download a repository from GitHub to your local machine freshly?",
    answer("`git clone`", correct = TRUE),
    answer("`git fork`", message = '`git fork` is a not a git command'),
    answer("`git push`"),
    answer("`git pull`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Which command creates an empty Git repository in the specified directory?",
    answer("`git init`", correct = TRUE),
    answer("`git log`"),
    answer("`git reset`"),
    answer("`git config`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to upload your changes to GitHub?",
    answer("`git push`", correct = TRUE),
    answer("`git add`"),
    answer("`git status`"),
    answer("`git clone`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to check the state of local repo since last commit?",
    answer("`git status`", correct = TRUE),
    answer("`git add`"),
    answer("`git commit`"),
    answer("`git diff`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to stage files for a commit?",
    answer("`git add`", correct = TRUE),
    answer("`git commit`"),
    answer("`git status`"),
    answer("`git reset`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to save the current state of your repo into git?",
    answer("`git commit`", correct = TRUE),
    answer("`git add`"),
    answer("`git init`"),
    answer("`git diff`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the shortcut to stage all the changes that you made?",
    answer("`git add .`", correct = TRUE),
    answer("`git add all`"),
    answer("`git commit .`"),
    answer("`git commit all`"),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("How do you supply a commit message to a commit?",
    answer('`git commit -m "Commit message"`', correct = TRUE),
    answer('`git commit "Commit message"`'),
    answer('`git add -m "Commit message"`'),
    answer('`git add "Commit message"`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("How do you commit for all changes with a message?",
    answer('`git commit -am "Commit message"`', correct = TRUE),
    answer('`git commit -a "Commit message"`'),
    answer('`git add -am "Commit message"`'),
    answer('`git add -a "Commit message"`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("When you commit changes to git, which one comes first: staging with `git add` or committing with `git commit`?",
    answer('Staging with `git add`', correct = TRUE),
    answer('Committing with `git commit`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to unstage a currently staged file called geospaar.txt?",
    answer('`git checkout -- geospaar.txt`', correct = TRUE),
    answer('`git reset geospaar.txt`'),
    answer('`git HEAD^ -- geospaar.txt`'),
    answer('`git unstage -- geospaar.txt`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to make a new branch?",
    answer('`git branch`', correct = TRUE),
    answer('`git -b`'),
    answer('`git new branch`'),
    answer('`git checkout branch`'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("What is the git command to switch to another branch?",
    answer('`git branch`'),
    answer('`git -b`'),
    answer('`git new branch`'),
    answer('`git checkout branch`', correct = TRUE),
    random_answer_order = TRUE, allow_retry = TRUE
  )
)
```

