---
title: "Week 1"
output: 
  learnr::tutorial:
    theme: united
    highlight: tango
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
gradethis::gradethis_setup()
tutorial_options(exercise.completion = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

## Class 1

### Quiz

```{r quiz_c1}
quiz(
  # Took this question from https://github.com/rstudio/learnr/blob/master/inst/tutorials/ex-setup-r/ex-setup-r.Rmd
  question("What is the RStudio IDE?",
    answer("An application that makes it easier to use R.", 
           correct = TRUE, 
           message = "RStudio organizes your workspace and makes it easier to write, use, debug, and save R code."),
    answer("An application that let's you use R without writing any code", 
           message = "Code provides a reproducible record of your work, which is essential for data science."),
    answer("A spreadsheet program like Microsoft Excel."),
    answer("Another name for R", 
           message = "R and RStudio are two separate things. R is a language, like English. Think of RStudio as a program that helps you use the language, kind of like how a word processing program helps you write in English."),
    random_answer_order = TRUE, allow_retry = TRUE
  )
)
```

## Class 2

### Quiz

```{r quiz_c2}
quiz(
  question("What of items listed below are atomic data types?",
    answer("chracter", correct = TRUE),
    answer("vector", message = 'Vector is an object.'),
    answer("integer", correct = TRUE),
    answer("factor", message = 'Factor is a special integer type, but not atomic type.'),
    answer("logical", correct = TRUE),
    answer("raw", correct = TRUE),
    answer("Date", message = 'Date is a special double type, but not atomic type.'),
    correct = "6 atomic data type in R: character, double, integer, logical, complex, and raw.",
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("A matrix is ___dimensional object? ",
    answer("1", message = 'A matrix has dimension attribute, which means it is not one dimensional.'),
    answer("2", correct = TRUE),
    answer("3", message = 'An array is the obejct that could have 3 or higher dimensions.'),
    answer("Higher", message = 'An array is the obejct that could have 3 or higher dimensions.'),
    correct = "A matrix is 2 dimensional object. A vector is 1 dimensional, and array is higher dimensional.",
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Attributes of an object (if any) can be accessed using function ___.",
    answer("`attributes()`", correct = TRUE),
    answer("`str()`", message = '`str()` function is used for compactly displaying the internal structure of a R object.'),
    answer("`attr()`", message = '`attr()` function is used to get or set specific attributes of an object.'),
    answer("`summary()`", message = '`summary()` is a generic function used to produce result summaries of the results'),
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Which one of the following expressions assigns the number 10 to the variable x?",
    answer("`x == 10`", message = '`==` is used to compare two objects.'),
    answer("`x <- 10`", correct = TRUE),
    answer("`x - 10`", message = '`-` is used to minus.'),
    answer("`x = 10`", correct = TRUE, 
           message = 'Even though it works, but it is highly recommended to use `<-` .'),
    correct = '`<-` and `=` both works in R to assign values, but it is highly recommended to use `<-`.', 
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Which one of the following function could split a string?",
    answer("`strsplit`", correct = TRUE),
    answer("`gsub`", message = '`gsub` is a function to replace sub-string within a string.'),
    answer("`paste`", message = '`paste` is a function to put two strings together with a defined sep symbol.'),
    answer("`paste0`",  message = '`paste0` is a function to put two strings together without a sep symbol.'),
    correct = '`strsplit` is a function to split a string. `gsub` is a function to replace sub-string within a string. `paste` is a function to put two strings together with a defined sep symbol. `paste0` could paste two strings without a sep symbol.', 
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Suppose I have a vector x <- c(12, 10, 17, 14, 4, 5, 13) and I want to get the indexes of all elements of this vector that are greater than 10. What R code should I use?",
    answer("`which(x > 10)`", correct = TRUE),
    answer("`any(x > 10)`", message = '`any` is a function to check if at least one of the values bigger than 10.'),
    answer("`all(x > 10`", message = '`all` is a function to check if all values of x bigger than 10.'),
    answer("`10 %in% x`",  message = '`%in%` is a function to check if 10 is in x.'),
    correct = '`which` is a function to get the indexes of elements match a checking condition. `any` is a function to check if at least one of the elements match a checking condition. `all` is a function to check if all elements match a checking condition. `%in%` a function to check if object(s) is in another object.', 
    random_answer_order = TRUE, allow_retry = TRUE
  ),
  question("Suppose I have a vector x <- c(12, 10, 17, 8), a vector y <- c(7, 10, 12), and a vector z <- c(9, 10). What I will get if run `x + y` and `x + z`?",
    answer("`c(19, 20, 29, 15)` and `c(21, 20, 26, 18)`", correct = TRUE),
    answer("`c(19, 20, 29)` and `c(21, 20)`", message = 'When two vectors with different length do calculation, R will repeat the shorter one to match with the longer one.'),
    answer("`c(19, 20, 29, 8)` and `c(21, 20, 17, 8)`", message = 'When two vectors with different length do calculation, R will repeat the shorter one to match with the longer one.'),
    answer("Syntax error", message = 'This is runnable in R.'),
    correct = 'When two vectors with different length do calculation, R will repeat the shorter one to match with the longer one. If the length of longer one is not a multiple of shorter one, it will give you a warning.', 
    random_answer_order = TRUE, allow_retry = TRUE
  )
)
```

### Exercise 

#### R markdown

Write an R markdown file to ensure the knit result is as follow: [expected result](https://agroimpacts.github.io/geospaar/class2_rmarkdown_practice.html). And then share your html in our slack channel.

#### Coding
Write the R code with proper style required to:

1. create `i` as a vector of 1:5.
2. create `f` as a vector of `c(11.1, 18.74, 29.02, 47.45, 76.2)`
3. use `cbind` to bind `i` and `f` as a matrix `mc` of integer. And set the column names as "a" and "b".
4. print out `mc`.
5. use `rbind` to bind `i` and `f` as a matrix `mr` of double. And set the row names as "a" and "b".
6. print out `mr`.
7. create `V` as a vector of a:e.
8. create `b` as a vector of `c(TRUE, TRUE, FALSE, FALSE, TRUE)`.
9. collect `i`, `f`, `V` and `b` into a list `l`, and set their names.
10. add a meta data (attribute) 'objective' to `l` with value "combine objects into list".
11. check the attributes of `l`.
12. print out `l`.
13. combine `i`, `f`, `V` and `b` into columns of a data.frame `df`, and set their names.
14. check the attributes of `df`.
15. print out `df`.

```{r v_int, exercise=TRUE, exercise.lines = 33}
# Create a

# Create b

# Create mc and set names


# Print out mc

# Create mr and set names


# Print out mr

# Create V

# Create b

# Create l

# Add attribute to l

# Check the attributes of l

# print out l

# Create df

# Check attributes of df

# print out df

```

Expected output:

```{r v_int_result, echo = F, eval = T}
# Create a
i <- 1:5
# Create b
f <- c(11.1, 18.74, 29.02, 47.45, 76.2)
# Create mc and set names
mc <- cbind(i, as.integer(f))
colnames(mc) <- c('a', 'b')
# Print out mc
mc
# Create mr and set names
mr <- rbind(as.double(i), f)
rownames(mr) <- c('a', 'b')
# Print out mr
mr
# Create V
V <- c('a', 'b', 'c', 'd', 'e')
# Create b
b <- c(TRUE, TRUE, FALSE, FALSE, TRUE)
# Create l
l <- list(i = i, f = f, V = V, b = b)
# Add attribute to l
attr(l, 'objective') <- "combine objects into list"
# Check the attributes of l
attributes(l)
# print out l
l
# Create df
df <- data.frame(i = i, f = f, V = V, b = b)
# Check attributes of df
attributes(df)
# print out df
df
```

```{r v_int-solution, exercise.reveal_solution = T}
# Create a
i <- 1:5
# Create b
f <- c(11.1, 18.74, 29.02, 47.45, 76.2)
# Create mc and set names
mc <- cbind(i, as.integer(f))
colnames(mc) <- c('a', 'b')
# Print out mc
mc
# Create mr and set names
mr <- rbind(as.double(i), f)
rownames(mr) <- c('a', 'b')
# Print out mr
mr
# Create V
V <- c('a', 'b', 'c', 'd', 'e')
# Create b
b <- c(TRUE, TRUE, FALSE, FALSE, TRUE)
# Create l
l <- list(i = i, f = f, V = V, b = b)
# Add attribute to l
attr(l, 'objective') <- "combine objects into list"
# Check the attributes of l
attributes(l)
# print out l
l
# Create df
df <- data.frame(i = i, f = f, V = V, b = b)
# Check attributes of df
attributes(df)
# print out df
df
```
