---
title: "class 6 challenge"
author: "Lei Song"
date: "3/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We use the same dataset as class 5 challenge to practice how to use `*apply` family:

- `apply()`
- `lapply()`
- `sapply()`
- `tapply()`
- `mapply()`

You have to use `*apply` functions to finish all tasks, except that you might need to use `dplyr` to subset the dataset. In order to let you practice how to use these functions, as well as how to pick up the right function, I won't mention which function you should use. 

## Task

Write an R markdown file for the tasks. Read the dataset (`crop_rmd`) the same as class 5 challenge.

```{r, warning=F, message=F}
# Task 1
# install.packages('readr')
library(readr)
library(dplyr)
library(here)

# Load data
crop_rmd <- read_csv(here('docs/homeworks/Crop_recommendation.csv'))
```

### Task 1

1. Select watermelon
2. Calculate the `mean` of all items (temperature, rainfall, etc.). Assign the result to `wml_mean` and print out.
4. What if I want to add a random noise from `rnorm(mean = 0, sd = 2)` to these means? Assign the result to `wml_noisy` and print out. Use seed 10 to for the noises. Hint: use anonymous function within a `*apply` function.

```{r}
watermelon <- crop_rmd %>% 
  filter(label == 'watermelon') %>% 
  dplyr::select(-label)
wml_mean <- apply(watermelon, 2, mean)
wml_mean
set.seed(10)
wml_noisy <- apply(watermelon, 2, function(x) mean(x) + rnorm(1, 0, 2))
```

### Task 2

Still for watermelon, I want to calculate $(temperature - 20) / (rainfall - 100)$, what is the easiest way to do that using `*apply`?

```{r}
mapply(function(x, y){
  (x - 20) / (y - 100)
}, watermelon$temperature, watermelon$rainfall)
```

#### Task 3

Loop over crop types to calculate max of each type of observations. You are expected to get a matrix back and assign to object `crop_smr`. Hint: You might need function `unique`, and write anonymous function and use `dplyr` within this anonymous function.

```{r}
crop_smr <- sapply(unique(crop_rmd$label), function(crop_type){
  crop_rmd %>% 
    filter(label == crop_type) %>% 
    dplyr::select(-label) %>% 
    colMeans()
})
crop_smr
```

#### Task 4

What is the most convenient way to calculate the max value of temperature for each crop types?

```{r}
tapply(crop_rmd$temperature, crop_rmd$label, max)
```
