---
title: "Geospatial Analysis with R"
subtitle: Class 13
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
## Election Results
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width=8, fig.height=5}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")

cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>% 
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% 
  dplyr::select(-other, -delres) %>% 
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab(NULL) + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

---
```{r, eval = FALSE}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")
cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>%
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% select(-other, -delres) %>%
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab("") + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2))
```
---

## Today
- Spatial
  - Non-spatial to spatial
  - Read and write spatial
  - CRS
  - Make data from scratch
  - Spatial properties
  - joins, subsets, intersections...
  
---




### Non-spatial to spatial

- Read in farmer data
```{r, eval = FALSE}
library(tidyverse)
library(sf)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() 
farmers_sf <- st_as_sf(farmers, coords = c("x", "y"), crs = 4326)
```

- Write and read the spatial data
```{r, eval=FALSE}
write_sf(farmers_sf, dsn = "external/notebooks/data/farmers_spatial.geojson")
farmers_sf <- read_sf("external/notebooks/data/farmers_spatial.geojson")
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
```

---
  
### Wrangle spatial data
```{r, eval = FALSE}
farmers %>% filter(uuid == "009a8424")
farmers_sf %>% filter(uuid == "009a8424")
farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% 
  arrange(n) %>% st_geometry %>% plot(pch = 16, cex = 0.5)
farmers_sf %>% arrange(uuid)
class(farmers_sf)
# intersection with st_join and left = FALSE to remove plots outside of Zambia
farmers_sf_trim <- farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% 
  st_join(., districts, left = FALSE)
plot(st_geometry(districts))
farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% st_geometry %>% 
  plot(pch = 16, add = TRUE, cex = 1)
farmers_sf_trim %>% st_geometry %>% 
  plot(pch = 16, add = TRUE, col = "red", cex = 0.5)
```

---
### Make spatial data

```{r, eval = FALSE}
pts <- st_multipoint(x = cbind(x = c(27.5, 28, 28.5), y = c(-14.5, -15, -15.5)))
pts <- st_geometry(pts) %>% st_as_sf() %>% st_cast("POINT") %>% 
  cbind(ID = 1:nrow(.), .)
pol <- st_polygon(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5), 
                             y = c(-15.5, -16.5, -17, -16, -15.5))))
pol <- st_geometry(pol) %>% st_as_sf() %>% cbind(ID = 1:nrow(.), .)
sline <- st_linestring(cbind(x = c(27, 27.5, 28), y = c(-15, -15.5, -16)))
sline <- st_geometry(pol) %>% st_as_sf() %>% cbind(ID = 1:nrow(.), .)
```

---
### Spatial properties
```{r, eval=FALSE}
pol %>% st_area
pol %>% st_sf(crs = 4326) %>% st_area
sline %>% st_sf(crs = 4326) %>% st_length
```

---
### Spatial transforms
```{r, eval=FALSE}
pol %>% st_sf(crs = 4326) %>% st_transform(., st_crs(roads)) %>% st_area
```

---
### Adding to data

Add some categories
```{r, eval=FALSE}
set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
```
---

## Analyses
```{r, eval=FALSE}
# districts2 %>% dplyr::select(yield) %>% summarize_all(funs(mean, sd))
# districts2 %>% summarize(mean(yield), sd(yield))
districts2 %>% group_by(yld_cat) %>% 
  summarize(yield = mean(yield)) %>% 
  ggplot() + geom_sf(aes(fill = yield))
  # geom_sf(aes(fill = yld_cat))
```

---
### Spatial joins
```{r, eval=FALSE}
p <- ggplot(districts2) + geom_sf(fill = "grey")
# farmer_ct <- farmers %>% group_by(uuid) %>% count()
# st_crs(farmer_ct) <- st_crs(districts)
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))

farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
p + geom_sf(data = farmers_districts, aes(color = n)) + 
  scale_color_viridis_c()
plot(districts2['yld_cat'])

# lapply(farmers_per_dist$distName, function(x) {
#   farmers_per_dist %>% filter(distName == x) %>% st_cast("POINT")
# }) %>% do.call(rbind, .)
# farmers_per_dist %>% st_cast("POINT") %>% plot()
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
p + geom_sf(data = farmers_per_dist) + 
  geom_sf(data = farmers_per_dist %>% st_centroid(), col = "blue")
```

---
### Spatial-non-spatial join
```{r, eval=FALSE}
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))

p + geom_sf(data = districts3, aes(fill = log10(n))) +
  scale_fill_brewer(palette = "RdBu")
  # scale_fill_viridis_c()
# p + geom_sf(data = districts3, aes(fill = n))
```

---
### Subsetting

By count
```{r, eval = FALSE}
p + districts3 %>% filter(n > 15) %>% geom_sf(data = ., fill = "red") + 
  districts3 %>% filter(between(n, 1, 15)) %>% geom_sf(data = ., fill = "blue")
p + geom_sf(data = districts3 %>% filter(n > 15), fill = "red") + 
  geom_sf(data = districts3 %>% filter(between(n, 1, 15)), fill = "blue")
```

By area
```{r, eval = FALSE}
dmed <- districts3 %>% #mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  mutate(area = as.numeric(units::set_units(st_area(.), "km2"))) %>% 
  filter(area == quantile(area, probs = 0.5, type = 1)) 
p2 <- p + geom_sf(data = dmed, fill = "red")

dmax <- districts3 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  # arrange(area) %>% slice(nrow(.)) %>% 
  arrange(desc(area)) %>% slice(1) 
  # slice_max(order_by = area) #
p2 + geom_sf(data = dmax, fill = "blue")

districts3 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  arrange(area) %>% pull(distName)
```

---
## Various

Dataset
```{r, eval=FALSE}
pols <- st_multipolygon(list(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5),
                                        y = c(-15.5, -16.5, -17, -16, -15.5))),
                             list(cbind(x = c(26.5, 27.5, 27, 26, 26.5) + 1,
                                        y = c(-15.5, -16.5, -17, -16, -15.5) -
                                          1))))
pols <- pols %>% st_geometry %>% st_cast("POLYGON") %>% 
  st_as_sf(crs = 4326) %>%   
  mutate(ID = 1:nrow(.)) %>% #as.data.frame() %>%
  dplyr::select(ID)

p + geom_sf(data = pols, aes(fill = as.factor(ID))) + 
  geom_sf(data = st_union(pols), 
          fill = rgb(1, 0, 1, alpha = 0.7), col = "purple")
```

---
### Intersects
```{r, eval=FALSE}
# which(st_intersects(districts2, st_union(pols), sparse = FALSE))
int_dists <- districts2 %>% 
  slice(which(st_intersects(., st_union(pols), sparse = FALSE)))
p + geom_sf(data = int_dists, fill = "red") + 
  geom_sf(data = pols, fill = "transparent", col = "blue")
```

---
### Intersection
```{r, eval=FALSE}
districts_pol <- st_intersection(districts2, st_union(pols))
ggplot(districts_pol) + geom_sf(aes(fill = distName))
p + geom_sf(data = districts_pol, aes(fill = distName))
```

---
### On your own

- Select the smallest district in Zambia from `districts3`, and convert that
to centroid
```{r, eval = FALSE}
smallest_centroid <- districts3 %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter
```

- Create a buffer of width 0.5 around that
```{r, eval = FALSE}
smallest_buffer <- smallest_centroid %>% st_buf...
```

- Use the buffer to select the districts from `districts3` that intersect with it
```{r, eval = FALSE}
dists_in_buff <- districts3 %>% 
  filter(st_intersects(., smallest_buffer...
```


---
- Use the buffer to do an intersection with `districts3`
```{r, eval=FALSE}
buff_int_dists <- st_intersection(...
# p + geom_sf(data = buff_int_dists, aes(fill = distName))
```


- Calculate the area of the intersection



