---
title: "Geospatial Analysis with R"
subtitle: Class 19
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Challenge
- Raster intro, cont.

---
## Challenge

- Create 20 random rasters using lapply
  - resolution of 0.1
  - extent of Zambia
  - convert to brick
- crop and mask to extent of Zambia’s largest district
- plot  


---
## Calculations

- raster algebra
- statistics
- z dimension stats


---
## Calculations

```{r, eval=FALSE}
rf1 <- chirpsz[[1]] + chirpsz[[2]]
rf2 <- chirpsz[[1]] * 5

mx <- cellStats(rf2, max)
zl <- c(0, mx)
par(mfrow = c(2, 2))
plot(chirpsz[[1]], zlim = zl)
plot(chirpsz[[2]], zlim = zl)
plot(rf1, zlim = zl)
plot(rf2, zlim = zl)

# dev.off()
plot(rf1 / 1000)
plot(((log10(rf1 + 1) * 10) / 5)^2)

```

- statistics
```{r, eval=FALSE}
# mean(c(1, 4, 10, NA), na.rm = TRUE)
cellStats(chirpsz[[1]], stat = mean)
cellStats(chirpsz[[1]], stat = range)
cellStats(chirpsz, stat = mean)
cellStats(chirpsz, stat = sum)
cellStats(chirpsz, stat = range)
cellStats(chirpsz, summary)
```

---
- z dimension stats
```{r, eval=FALSE}
plot(calc(chirpsz, mean))
plot(calc(chirpsz, sum))
plot(calc(chirpsz, range))
```
---

- neighborhoods

```{r, eval=FALSE}
raintot <- calc(chirpsz, stat = sum)
w <- matrix(1, nrow = 5, ncol = 5)
rainwindow <- focal(raintot, w = w, fun = mean, na.rm = TRUE)
rainwindow2 <- focal(raintot, w = w, fun = mean)
rainwindow3 <- focal(raintot, w = w, fun = sd)
par(mfrow = c(2, 2), mar = rep(0, 4))
plot_noaxes(raintot, legend = FALSE)
plot_noaxes(rainwindow, legend = FALSE)
plot_noaxes(rainwindow2, legend = FALSE)
plot_noaxes(rainwindow3, legend = FALSE)
```

---

- reclassify

```{r, eval=FALSE}
raincat <- cut(raintot, breaks = c(0, 50, 100, 200))
cols <- c("grey70", "lightblue", "lightgreen")
plot_noaxes(raincat, col = cols,
            legend = FALSE,  main = "Total Rainfall",
            mar = c(0, 0, 1, 0))
legend(x = "bottomright", legend = c("high", "medium", "low"),
       pch = 15, pt.cex = 3, col = rev(cols), bty = "n")
```

```{r, eval=FALSE}
rain <- cellStats(raintot, range)
rain_3 <- cbind(c(floor(rain[1]), 50, 100),
                 c(50, 100, ceiling(rain[2])), 1:3)
raincat2 <- reclassify(raintot, rain_3, include.lowest = TRUE)
plot_noaxes(raincat2, col = cols,
            legend = FALSE,  main = "Total Rainfall",
            mar = c(0, 0, 1, 0))
legend(x = "bottomright", legend = c("high", "medium", "low"),
       pch = 15, pt.cex = 3, col = rev(cols), bty = "n")

```

---
## Exercises

- Create a dummy raster (sampling from 1:100) using district 49 for extent and res of 0.1
- Reproject to Albers
- Disaggregate to 0.02 degrees (bilinear and NGB)
- Calculate the sum of rainfall from `chirps`
- Identify all areas of total rainfall > 10 mm
- Calculate the mean rainfall in Zambia for the 15th day in `chirps`

---
## Terrain analysis with zonal statistics

```{r, eval=FALSE}
districts <- districts %>% mutate(ID = 1:nrow(.))
distsr <- districts %>% rasterize(x = ., y = raintot, field = "ID") %>% print()
plot(distsr)

# Download DEM
data(dem)
# dem <- getData(name = "alt", country = "ZMB", path = tempdir())
dem <- getData(name = "worldclim", var = 'tmean', res = 2.5,  path = tempdir())

# calculate slope
slope <- terrain(x = dem, opt = 'slope', unit = 'degrees')
plot(slope)

# calculate mean by district
distsr_rs <- resample(x = distsr, y = slope, method = "ngb")
zoneslope <- zonal(x = slope, z = distsr_rs, fun = "mean")
hist(zoneslope[, 2])
zoneelevation <- zonal(x = dem, z = distsr_rs, fun = "mean")
hist(zoneelevation[, 2])
```
---

```{r, eval=FALSE}
# map zonal statistics
distr_slopezone <- zoneslope %>% data.frame %>%  
  subs(x = distsr_rs, y = ., by = "zone")
distr_elezone <- zoneelevation %>% data.frame %>% 
  subs(x = distsr_rs, y = ., by = "zone")

# plot
l <- list(distr_elezone, distr_slopezone)
titles <- c("Elevation", "Slope")
par(mfrow = c(1, 2))
for(i in 1:length(l)) {
  plot_noaxes(l[[i]], main = titles[i])
}
```

---
## Climate data summaries and zonal stats

```{r, eval=FALSE}
# rainfall <- getData(name = "worldclim", res = 10, var = 'tmax',
#                     path = tempdir())
# dt <- as.Date(gsub("Y", "", names(chirpsz)), format = "%y%j")
dt <- lubridate::parse_date_time(gsub("Y", "", names(chirpsz)), orders = "yj")
wk <- lubridate::week(dt)
weekly_rainfall <- lapply(unique(wk), function(x) {
  which(wk == x)
  calc(chirpsz[[which(wk == x)]], sum)
}) %>% stack(.)
names(weekly_rainfall) <- paste0("wk", unique(wk))

rflim <- range(cellStats(weekly_rainfall, range))
plot(stack(weekly_rainfall), zlim = rflim)
```





