---
title: "Geospatial Analysis with R"
subtitle: Class 12
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Attribute operations
- Spatial operations
- Geometry operations

---

## Use `osmdata` to query some vectors

```{r, eval=F}
library(dplyr)
library(osmdata)
q <- opq(bbox = st_bbox(clark_campus_ply))

# Building
buildings <- q %>% 
  add_osm_feature(key = 'building') %>%
  osmdata_sf() %>% `[[`('osm_polygons') %>% 
  filter(!is.na(name))

# Streets
streets <- q %>% 
  add_osm_feature(key = 'highway') %>%
  osmdata_sf() %>% `[[`('osm_lines') %>% 
  filter(!is.na(name))

# footways
footways <- q %>% 
  add_osm_feature(key = c('foot')) %>%
  osmdata_sf() %>% `[[`('osm_lines')

# Trees
trees <- q %>% 
  add_osm_feature(key = c('natural')) %>%
  osmdata_sf() %>% `[[`('osm_points')
```

---

## Attribute operations

- `sf` is a special data.frame, so all base R and `dplyr` syntax works on it. E.g.

  - subsetting
  - aggregation (group)
  - joining

- `st_drop_geometry()` to drop geometry column

---

## Attribute operations

Tasks:

- Select out `streets` that have good or fair condition
- Calculate the mean width of `streets` by conditions.
- Drop the geometry from `streets`, and join the table with `buildings`, then select out the buildings that belong to intolerable streets.

---

- Select out `streets` that have good or fair condition
- Calculate the mean width of `streets` by conditions.
- Drop the geometry from `streets`, and join the table with `buildings`, then select out the buildings that belong to intolerable streets.

```{r, eval=F}
# 1
streets %>% filter(condition %in% c('good', 'fair'))

# 2
streets %>% mutate(width = as.numeric(width)) %>% 
  group_by(condition) %>% 
  summarize(width = mean(width, na.rm = T))

# 3
left_join(buildings, 
          streets %>% st_drop_geometry(), 
          by = c(addr.street = 'name')) %>% 
  filter(condition == 'intolerable')
```

---

## Spatial operations

- Topological relations (returns are index):

  - `st_intersects`
  - `st_disjoint`
  - `st_within`
  - `st_touches`
  - `st_is_within_distance` = `st_buffer` + `st_intersects`

- Spatial joining:

  - `st_join`

- `st_distance`: works for all geometries, but only points.

---

Tasks:

- Manipulate `cnts_africa` and `occurrence` to get countries that have more than 100 occurrences. Hint: `st_join`, `group_by`.
- Get the countries that touch with Tanzania. Hint: `st_touch`.
- Get the occurrence inside of Tanzania.
- Get the occurrence outside of Tanzania.
- Calculate the distance between the first point and all other points of occurrence.

```{r, eval=F}
# Prepare data
library(rnaturalearth)
cnts_africa <- ne_countries(continent = 'africa', 
                            returnclass = 'sf') %>% 
  dplyr::select(name)
dt_url <- 'https://agroimpacts.github.io/geospaar/hyenas_occurrence.csv'
occurrence <- read.csv(dt_url, stringsAsFactors = F) %>% 
  dplyr::select(gbifID, decimalLatitude, decimalLongitude) %>% 
  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'),
           crs = 4326)
plot(cnts_africa %>% st_geometry())
plot(occurrence$geometry, add = T, pch = 20, col = 'red')
```


---

- Manipulate `cnts_africa` and `occurrence` to get countries that have more than 100 occurrences. Hint: `st_join`, `group_by`.
- Get the countries that touch with Tanzania. Hint: `st_touch`.

```{r, eval=F}
# 1
cnts_selected <- occurrence %>% 
  st_join(cnts_africa) %>% 
  filter(!is.na(name)) %>% 
  group_by(name) %>% 
  summarise(n = n()) %>% 
  st_drop_geometry() %>% 
  filter(n > 100) %>% 
  pull(name)
cnts_selected 

# 2
tz <- cnts_africa %>% 
  filter(name == 'Tanzania')
tz_touches <- cnts_africa %>% 
  slice(st_touches(tz, cnts_africa) %>% unlist())
```

---

- Get the occurrence inside of Tanzania.
- Get the occurrence outside of Tanzania.
- Calculate the distance between the first point and all other points of occurrence.

```{r, eval=F}
# 3
occ_tz <- occurrence %>% 
  slice(st_intersects(tz, occurrence) %>% unlist())

# 4
occ_out_tz <- occurrence %>% 
  slice(st_disjoint(tz, occurrence) %>% unlist())

# 5
pts_distance <- st_distance(occurrence %>% slice(1), 
                            occurrence)
pts_distance[1]
pts_distance[2]
```

---

## Geometry operations

- Statistics

  - `st_area`
  - `st_length`

- Clipping/overlapping

  - `st_intersection`
  - `st_difference` & `st_sym_difference`
  - `st_union`: return `sfc`

- Others

  - `st_centroid`
  - `st_buffer`
  - `st_coordinates`
  - `st_bbox`
  - `st_simplify`

---

```{r, eval=F}
# 1
africa <- st_union(
  cnts_africa %>% st_buffer(0.0001)) %>% 
  st_sf()
tz_bbox <- st_bbox(tz) %>% st_as_sfc() %>% 
  st_buffer(1, endCapStyle = 'SQUARE', 
            joinStyle = "BEVEL") %>% 
  st_sf()

# 2
st_difference(africa, tz_bbox) %>% 
  st_geometry() %>% plot(col = 'red')
st_intersection(tz_bbox, africa) %>% 
  st_geometry() %>% plot(add = T, col = 'blue')
st_difference(tz_bbox, africa) %>% 
  st_geometry() %>% plot(col = 'green', add = T)

# 3
st_sym_difference(tz_bbox, africa) %>% st_geometry() %>% 
  plot(col = 'yellow', add = T)

# 4
cnts_ctd <- cnts_africa %>% st_centroid()
plot(cnts_africa$geometry)
plot(cnts_ctd$geometry, pch = 20, col = 'red', add = T)
cnts_ctd %>% st_coordinates()
```

---

## Units

`units::set_units()`

```{r, eval=F}
# Examples of usage
10 %>% units::set_units('km2')
st_area(africa) %>% units::set_units('ha')
st_buffer(africa, 10 %>% 
            units::set_units('miles') %>% 
            units::set_units('m'))

# Remove units
set_units(x, NULL)
```

---

## Next class

- Visualization

---

## Homework

Reference to help document and [`sf` cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/sf.pdf), learn these functions by yourself and prepare examples:
- `st_sample`
- `st_contains`
- `st_nearest_feature`
- `st_nearest_points`
- `st_convex_hull`
- `st_crop`

Will ask you to share your examples in next class.
