---
title: "Geospatial Analysis with R"
subtitle: Class 15
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


## Today

- More vector analysis and exercises

---

Data:  read in `farmers` and make into `sf`, `districts`, and `roads`
```{r}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sf(coords = c("x", "y"))
st_crs(farmers) <- 4326
districts <- st_read(
  system.file("extdata/districts.shp", package = "geospaar")
)
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
```


---
## Data
```{r}
library(tidyverse)
library(geospaar)
farmers_alb <- farmers %>% st_transform(st_crs(roads))
long_roads <- roads %>% filter(as.numeric(st_length(.)) > 500000)
districts_alb <- districts %>% st_transform(st_crs(roads))
```

---
## Plotting
- Use same crs
- For `base` plotting, use `st_geometry`
```{r}
plot(st_geometry(districts), col = 'grey')
plot(roads[1], col = 'red', add = TRUE)

```

---
## st_intersects vs. st_intersection
- `st_intersects()` is a "find by location"
- Answer question - which districts intersect each road?
- order matters
```{r}
intersects_obj <- st_intersects(long_roads, districts_alb)
intersects_obj

```


```{r}
## plot districts intersecting first road
first_road <- long_roads[1,]
intersecting_dists <- districts_alb %>% slice(intersects_obj[[1]])
plot(st_geometry(intersecting_dists))
plot(first_road, col = 'red', add = T)
```

---
## st_intersection
- Returns intersecting object
-  st_intersection(lines, polygons) will be "lines"
- Each element is one component of intersection
```{r}
intersection_obj <- st_intersection(long_roads, districts_alb)
#print(intersection_obj)
plot(st_geometry(intersection_obj))
```

---
- Switching order of inputs only affects column order of output
```{r}
intersection_obj2 <- st_intersection( districts_alb, long_roads)
#print(intersection_obj)
plot(st_geometry(intersection_obj2))
```

---
## Dissolving
- Use `summarise` to dissolve
- Let's add a random Group ID for each district
- How do we plot?


```{r}
districts_alb$GROUP <- sample(LETTERS[1:3], NROW(districts_alb), replace = T)
FILLS = c("A" = "red", "B" = "slateblue", "C" = "green4")

ggplot(districts_alb) +
  geom_sf(aes(fill = GROUP)) +
  scale_fill_manual(values = FILLS)
```

---
## Dissolving
- Add an area column
- Dissolve features using `summarise`, and calculate total area
```{r}
## add area column
districts_alb <- districts_alb %>% mutate(area_km2 = st_area(.) %>% units::set_units("km^2") %>% as.numeric())
districts_alb
```

---
```{r}
## Dissolve and plot
dissolved_groups <- districts_alb %>% group_by(GROUP) %>% summarise(group_area = sum(area_km2))
plot(dissolved_groups)
```


---
## st_cast
- Cast sf objects to specific data type
- By default, will cast point--> multipoint, polygon --> multipolygon
- You can also specify output, to cast polygons --> lines
```{r}
district_borders <- districts_alb %>% st_cast("MULTILINESTRING")
plot(st_geometry(district_borders))
```




---
## st_union
- Each element in the output is a combination of one element in each input
- Output size is product of size of each input
```{r}
options(max.print = 999999999)
union_dists_roads <- st_union(districts_alb, long_roads)
print(union_dists_roads$geometry)
```

---
## st_union
- Isolate one district
```{r}
chibombo <- districts_alb %>% filter(distName == "Chibombo")
chibombo_union_roads <- st_union(chibombo, long_roads)
plot(chibombo_union_roads[3,])
#(chibombo_union_roads[3,] %>% pull(geometry))[[1]]
```





---
## split apply combine
- Separate areas into quintiles
- how can we add original district borders (in green)?
```{r}
quintiles <- function(x) quantile(x, probs = seq(0, 1, 0.2))

dist_quintiles <- districts %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  mutate(acls = cut(area, breaks = quintiles(area), include.lowest = TRUE)) %>%  group_by(acls) %>% summarize(mean_area = mean(area))

dist_quintiles


cols <- heat.colors(5)
par(mar = rep(0, 4))
plot(st_geometry(dist_quintiles), col = cols)
#plot(st_geometry(districts), border = "red", add = T)
legend(x = "bottomright", pch = 15, col = cols, bty = "n", 
       legend = paste0(1:5, c("st", "nd", "rd", "th", "th"), " quantile"))
```

---
## Joins
- For table joins, use `dplyr` functions `*_join`, or `merge`
- For spatial joins, use `st_join` 

```{r}
## spatial join of roads to districts
dists_join_roads <- st_join(districts_alb, long_roads)
```

---
## Joins
- How to find which roads intersect districts?
- `st_join` by default returns info for road 
```{r}
dists_join_roads <- st_join(districts_alb, long_roads, largest = F)
```

---
## Plot options
- Monday

---
## Adding to data

Add some categories
```{r, eval=FALSE}
set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
```
---

## Analyses
```{r, eval=FALSE}
# districts2 %>% dplyr::select(yield) %>% summarize_all(funs(mean, sd))
# districts2 %>% summarize(mean(yield), sd(yield))
districts2 %>% group_by(yld_cat) %>% 
  summarize(yield = mean(yield)) %>% 
  ggplot() + geom_sf(aes(fill = yield))
  # geom_sf(aes(fill = yld_cat))
```

---
## Joins
```{r, eval=FALSE}
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))
farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))

smallest_centroid <- districts3 %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter(area == min(area)) %>% st_centroid()
sf_use_s2(use_s2 = FALSE)
smallest_buffer <- smallest_centroid %>% st_buffer(dist = 0.5)
buff_int_dists <- st_intersection(smallest_buffer, districts3)
```

---
- Differencing
```{r, eval=FALSE}
dif <- st_difference(districts3, smallest_buffer)
ggplot(dif) + geom_sf(fill = "grey50")

dif2 <- st_difference(smallest_buffer, districts3)
ggplot(dif2) + geom_sf(fill = "grey50")
```

- Unioning

```{r, eval=FALSE}
uni <- st_union(dif)
ggplot(uni) + geom_sf(fill = "grey50")
uni2 <- st_union(dif, smallest_buffer)
ggplot(uni2) + geom_sf(fill = "grey50")
uni3 <- st_union(dif, buff_int_dists)
ggplot(uni3) + geom_sf(fill = "grey50")

uni3
```

---
- Sampling

```{r, eval=FALSE}
# simple within extent
set.seed(1)
sample100 <- districts3 %>% filter(n > 0) %>% 
  st_sample(size = 100, exact = TRUE)
ggplot(districts3) + geom_sf() + 
  geom_sf(data = districts3 %>% filter(n > 0), fill = "lightblue") + 
  geom_sf(data = sample100, size = 0.5, col = "red")

# stratified by polygon
ndists <- districts3 %>% filter(n > 0) %>% nrow(.)
strat_sample <- districts3 %>% filter(n > 0) %>% 
  st_sample(size = rep(5, ndists), exact = TRUE)

ggplot(districts3) + geom_sf() + 
  geom_sf(data = districts3 %>% filter(n > 0), fill = "lightblue") + 
  geom_sf(data = strat_sample, size = 0.5, col = "red")
```

---
### Exercises

#### Round 1
- Calculate the length of roads in kilometers
- Buffer roads > 100 km by 30 kilometers, save as new object
- Buffer roads > 100 km 10 kilometers, save as new object
- Take the difference between the 30 km and 10 km buffers, i.e. erase/difference the two
- Plot the resulting difference using `ggplot` with the difference in red over Zambia's districts

```{r, eval=FALSE}
# code to adapt
roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
  filter(km > 200) %>% 
  st_buffer(dist = 20000) %>% 
  st_transform(crs = 4326)
st_difference(larger_object, smaller_object)
ggplot() + geom_sf

```

---

#### Round 2
- Randomly sample 100 points within the 30 km road buffer (use a seed of 1)
- Plot the 30 km buffer (red) the points in it (blue) over the districts

```{r, eval=FALSE}
buffered_object %>% st_sample()
```

---
### How would you solve these problems? 

- Calculate the area of Zambia's districts 

    - mutate on districts

- **Accurately** calculate the area of Zambia's districts
- Find the largest and smallest districts in Zambia
- Find the longest road in Zambia
- Find the names of the districts that Zambia's longest road runs through
- Calculate the area of each of those districts along Zambia's longest road that lies within 10 km distance of that road
- Create a 0.5 (E-W) X 0.25 (N-S) degree polygon centered on Zambia's smallest district
- Count how many farmers from our sample live within 10 km of Zambian roads that are longer than 200 km

---
### Solve the problems!


---
## Group work
- Each group does hw # 1
- Groups split on other problems
- First 5 minutes. No coding, draw diagrams and determine what functions you need
- [SF cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/sf.pdf)
- [Examples](http://www.geo.hunter.cuny.edu/~ssun/R-Spatial/spatialops.html#spatial-operations)

--- 
## HW 
- Unit 2 Mod 1 due Tuesday
- Recommended: review links on previous slide
- QUIZ Monday
