---
title: "Geospatial Analysis with R"
subtitle: Class 17
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Homework from last class
- Your questions
- More exercises

---
### Data

```{r, eval=FALSE}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sf(coords = c("x", "y"))
districts <- st_read(system.file("extdata/districts.shp", package = "geospaar"))
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))

set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))
farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))
```

---

### Exercises

#### Round 1
- Calculate the length of roads in kilometers
- Buffer roads > 100 km by 30 kilometers, save as new object
- Buffer roads > 100 km 10 kilometers, save as new object
- Take the difference between the 30 km and 10 km buffers, i.e. erase/difference the two
- Plot the resulting difference using `ggplot` with the difference in red over Zambia's districts

```{r, eval=FALSE}
roads_10 <- roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
    filter(km > 100) %>% 
    st_buffer(dist = 10000) %>% 
    st_transform(crs = 4326) #%>% 
    # st_union(.)
roads_30 <- roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
    filter(km > 100) %>% 
    st_buffer(dist = 30000) %>% 
    st_transform(crs = 4326) #%>% 
    # st_union(.)
buf_diffs <- st_difference(roads_30, roads_10)
ggplot(districts) + geom_sf() + 
  geom_sf(data = buf_diffs, fill = "red")
```

---
```{r, eval=FALSE}
roads_bufs <- lapply(c(10000, 30000), function(x) {
  roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
    filter(km > 100) %>% 
    st_buffer(dist = x) %>% 
    st_transform(crs = 4326) %>% 
    st_union(.)
})
buf_diffs <- st_difference(roads_bufs[[2]], roads_bufs[[1]])
ggplot(districts) + geom_sf() + 
  geom_sf(data = buf_diffs, fill = "red")
```

---

#### Round 2
- Randomly sample 100 points within the 30 km road buffer (use a seed of 1)
- Plot the 30 km buffer (red) the points in it (blue) over the districts

```{r, eval=FALSE}
set.seed(1)
roads_sample <- roads_bufs[[2]] %>% st_sample(size = 100)
ggplot(districts) + geom_sf() + 
  geom_sf(data = roads_bufs[[2]], fill = "red") + 
  geom_sf(data = roads_sample, col = "blue")
```

---
### How would you solve these problems? 

- Calculate the area of Zambia's districts 

    - mutate on districts

- **Accurately** calculate the area of Zambia's districts
- Find the largest and smallest districts in Zambia
- Find the longest road in Zambia
- Find the names of the districts that Zambia's longest road runs through
- Calculate the area of each of those districts along Zambia's longest road that lies within 10 km distance of that road
- Create a 0.5 (E-W) X 0.25 (N-S) degree polygon centered on Zambia's smallest district
- Count how many farmers from our sample live within 10 km of Zambian roads that are longer than 200 km

---
### Solve the problems!






















