---
title: "Geospatial Analysis with R"
subtitle: Class 16
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Raster intro, continued
  - classes
  - dummy rasters
  - reading and writing
  - rasterize, raster to vector
  - pre-processing
  - calculations

---
### Demo

```{r, eval=FALSE}
library(parallel)
system.time(b3 <- lapply(1:250, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.001, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  stack(r, r * runif(n = ncell(r), 0.9, 1.2), r * runif(n = ncell(r), 0.8, 1.3))
})) 

system.time(b4 <- mclapply(1:250, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.001, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  stack(r, r * runif(n = ncell(r), 0.9, 1.2), r * runif(n = ncell(r), 0.8, 1.3))
}, mc.cores = 7)) 

system.time(l1 <- lapply(b3, function(x) calc(x, mean)))
system.time(l2 <- mclapply(b4, function(x) calc(x, mean), mc.cores = 7))

s1 <- stack(l2)
system.time(calc(s1, mean))
b1 <- brick(l2)
system.time(calc(b1, mean))
```

---


### Data
```{r, warning=FALSE, message=FALSE, results='hide'}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>%
  read_csv %>% distinct(uuid, .keep_all = TRUE) %>% select(uuid, x, y) %>% 
  mutate(count = 1) %>% st_as_sf(coords = c("x", "y"))
roads <- system.file("extdata/roads.shp", package = "geospaar") %>% st_read
districts <- system.file("extdata/districts.shp", package = "geospaar") %>%
  st_read
```

---
## Reading/Writing/Manipulating rasters

- create
```{r, eval = FALSE}
r <- raster(extent(30, 31, -14, -13), res = 0.1, 
            crs = "+proj=longlat +datum=WGS84")
values(r) <- sample(1:10, size = ncell(r), replace = TRUE)

par(mar = c(0, 0, 0, 0))
plot(districts %>% st_geometry)
plot(r, add = TRUE)
plot(r)

s <- stack(r, log10(r))
names(s) <- c("dummy", "log10dummy")
plot(s)

b <- brick(s)
plot(b)

b2 <- lapply(1:10, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.1, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  r
}) %>% stack %>% brick
plot(b2)
```

---
### A large brick
```{r, eval = FALSE}
b3 <- lapply(1:10, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.001, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  r
}) %>% stack %>% brick
plot(b3)

```

---

- read and write

```{r, eval = FALSE}
writeRaster(r, filename = file.path(tempdir(), "mydummy.tif"), overwrite = TRUE)
rm(r)
r
r <- raster(file.path(tempdir(), "mydummy.tif"))
plot(r)

brick(s, filename = file.path(tempdir(), "mydumbrick.tif"), overwrite = TRUE)
brick(s, filename = file.path(tempdir(), "mydumbrick.grd"), overwrite = TRUE)
brick(file.path(tempdir(), "mydumbrick.tif"))
brick(file.path(tempdir(), "mydumbrick.grd"))
```

---

- plot
```{r, eval = FALSE}
plot(b)
plot(b2)
plot_noaxes(b)
plot_noaxes(b2, main = paste("Random", 1:10))
plot(b2, main = paste("Random", 1:10))
plot_noaxes(b, legend = FALSE)
plot_noaxes(b[[1]], legend = FALSE)
legend("right", legend = 1:10, fill = terrain.colors(10), border = FALSE, 
       bty = "n")

# plotting with larger datasets
plot_noaxes(b3, main = paste("Random", 1:10))
rasterVis::levelplot(b3, names.attr = paste("Random", 1:10))

rasterVis::levelplot(b2, names.attr = paste("Random", 1:10))
rasterVis::levelplot(b[[1]])
rasterVis::gplot(b[[1]]) + geom_tile(aes(fill = value)) + 
  scale_fill_viridis_c()
ggplot2::ggplot(as.data.frame(b[[1]], xy = TRUE)) + 
  # geom_tile(aes(x = x, y = y, fill = dummy)) + scale_fill_viridis_c()  
  geom_raster(aes(x = x, y = y, fill = dummy)) + scale_fill_viridis_c()
st_as_stars(b) %>% plot(col = viridis::viridis(10))
st_as_stars(b) %>% plot()
```

---
## Raster to other types
```{r, eval=FALSE}
b[1:10]
b[1:ncell(b)]
values(b)
as.matrix(b)
as.data.frame(b) %>% as_tibble()
values(b[[1]])
```
---
## Pre-processing
- Aggregating/disaggregating
```{r, eval = FALSE}
aggregate(b, fact = 2) %>% plot
# aggregate(b, fun = min, fact = 2) %>% plot
# aggregate(b, fun = sd, fact = 2) %>% plot
plot(b)
disaggregate(b, fact = 2) %>% plot
disaggregate(b, fact = 2, method = "bilinear") %>% plot
```

- Masking
```{r, eval = FALSE}
data(chirps)
plot_noaxes(chirps[[1]])
chirpsz <- mask(chirps, districts)
plot_noaxes(chirpsz[[1]])
plot(st_geometry(districts), add = TRUE)
# rasterVis::levelplot(chirpsz[[1:5]])
plot_noaxes(chirpsz, nr = 2, nc = 3, maxnl = 6, mar = c(0, 0, 0, 0), 
            legend = FALSE, zlim = c(0, 50))
plot_noaxes(chirpsz, nr = 3, nc = 3, maxnl = 9, mar = c(0, 0, 0, 0), 
            legend = FALSE, zlim = c(0, 50))
plot_noaxes(chirpsz, nr = 2, nc = 3, maxnl = 6, mar = c(0, 0, 0, 0), 
            legend = FALSE, zlim = c(0, 50))
plot(chirpsz, nr = 2, nc = 3, maxnl = 6, mar = c(0, 0, 0, 0), 
     legend = FALSE, zlim = c(0, 50))
```

---

- Projecting
```{r, eval = FALSE}
chirpsz %>% projectRaster(., crs = crs(roads), res = 5000) -> chirpsz_alb
chirpsz_alb %>% print
plot(chirpsz_alb[[1]])
plot(chirpsz[[1]])
```

- Rasterize/vectorize
```{r, eval = FALSE}
distsr <- districts %>% rasterize(., chirpsz)
distsr %>% plot_noaxes
# stack(distsr, chirpsz[[1:2]]) %>% plotRGB(stretch = "lin")
distsr_pol <- rasterToPolygons(distsr, dissolve = TRUE)
distsr_pol %>% st_as_sf %>% st_geometry %>% plot
distsr_pol %>% st_as_sf %>% slice(49) %>% plot(add = TRUE)
distsr_pol %>% st_as_sf %>% slice(49) %>% plot()
```

---
## Calculations

- raster algebra
```{r}
rf1 <- chirpsz[[1]] + chirpsz[[2]]
plot(rf1)
plot(rf1 / 1000)
```

- statistics
```{r}
cellStats(chirpsz[[1]], stat = mean)
cellStats(chirpsz[[1]], stat = range)
cellStats(chirpsz, stat = mean)
cellStats(chirpsz, stat = sum)
cellStats(chirpsz, stat = range)
```

- z dimension stats
```{r}
plot(calc(chirpsz, mean))
plot(calc(chirpsz, sum))
plot(calc(chirpsz, range))
```

---
## Exercises

- Create a dummy raster (sampling from 1:100) using district 49 for extent and res of 0.1
- Reproject to Albers
- Disaggregate to 0.02 degrees (bilinear and NGB)
- Calculate the sum of rainfall from `chirps`
- Identify all areas of total rainfall > 10 mm
- Calculate the mean rainfall in Zambia for the 15th day in `chirpsz`

