---
title: "Class8 demo"
author: 
- Lei Song^1^
- ^1^ Clark Unversity
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Class8 demo}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

Within this vignette (or document), packages `here` is used to take care of the paths. `dplyr` is used to do data manipulation, and `readr` is used to read csv file. 

In terms of data visualization, `kableExtra` is used to make tables, and `ggplot2` is used to make figures.

```{r, message=FALSE, warning=FALSE}
# Load all packages in need
library(here)
library(dplyr)
library(readr)
library(ggplot2)
library(cowplot)
library(kableExtra)
```

## Combining
### Prepare dummy data

In order to show how `*_join` works, I created dummy data by splitting the data recommendation dataset into weather part and soil part. Then randomly select 80% from soil part to make it shorter than weather part.

```{r}
# Calculate the mean of all observations for each crop type
## Change to your own path
crop_rmd <- read_csv(here(
  file.path('docs/homeworks',
            'Crop_recommendation.csv'))) %>% 
  group_by(label) %>% 
  summarise_at(c('temperature', 'humidity', 'rainfall',
                 'N', 'P', 'K', 'ph'), mean)

# Split the table into two parts: weather and soil
crop_rmd_weather <- crop_rmd %>% 
  select(label, temperature, humidity, rainfall)
crop_rmd_weather

set.seed(10)
crop_rmd_soil <- crop_rmd %>% 
  select(label, N, P, K, ph) %>% 
  sample_frac(0.8) %>% 
  # add a dummy row for soybean
  rbind(data.frame(label = 'soybean', 
                   N = 18.9, P = 78.4, 
                   K = 45.3, ph = 6.45))
crop_rmd_soil
```

### `*_join`

Then I take the dummy data that I created above to test how these `*_join` functions work and what are the differences between them.

```{r}
# Left join, keep all rows of data on left side.
crop_lj <- left_join(crop_rmd_weather, crop_rmd_soil, 
                     by = 'label')
# Right join, keep all rows of data on right side.
crop_rj <- right_join(crop_rmd_weather, crop_rmd_soil, 
                      by = 'label')
# What I will get if I do this?
crop_lj2 <- left_join(crop_rmd_soil, crop_rmd_weather, 
                      by = 'label')

# Inner join, take the common part of data on left and right side.
crop_ij <- inner_join(crop_rmd_weather, crop_rmd_soil, 
                      by = 'label')
# Full join, keep all rows of data on both left and right side.
crop_fj <- full_join(crop_rmd_weather, crop_rmd_soil, 
                     by = 'label')
```

## Make tables

Then, using the data generated above, I create some tables using package `kableExtra`. You could find more details and examples [here](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html).

```{r}
crop_ij %>%
  kbl(caption = 'Mean of observations') %>%
  kable_paper(lightable_options = 'basic', # could be basic, striped, or hover
              html_font = 'helvetica', # could be any font installed on your machine.
              font_size = 12, # font size
              full_width = F, # full width or not
              position = "left", # center, left, right, or float-left and float-right
              fixed_thead = T) %>% # for a very long cross-pages table.
  column_spec(1, bold = T, background = 'pink') %>% # customize a column
  row_spec(which(crop_ij$temperature == max(crop_ij$temperature)),
           italic = T, color = 'red') # customize a row
```

## Make figures

In this section, I will use the data generated above to make some figures using both base R and `ggplot2`.

### Base plot
#### Scatterplot

I make scatterplots to show the relationship between ph value, rainfall and humidity. Then fit a linear model to show their quantitative correlation. The figure below show a highly customized figure.

```{r, fig.align='center', fig.width=6, fig.height=4}
# set the margin, c(bottom, left, top, right)
par(mar = c(3, 4, 3, 4)) 
# Make the main plot
plot(crop_ij$ph, crop_ij$rainfall, # set x and y
     pch = 20, col = 'blue', # set points style
     main = 'Relationship between ph, rainfall and humidity', # add main
     col.main = 'brown', # set main color
     xlab = 'ph', ylab = 'Rainfall') # set x and y axis

# Add a fit line
abline(lm(crop_ij$rainfall ~ crop_ij$ph),
       lty = 3, lwd = 3, col = 'blue')

# Add another scatterplot using points
## Reset plot window
plot.window(xlim = range(crop_ij$ph), 
            ylim = range(crop_ij$humidity))

## Another syntax to use plot: with(dataset, plot(x, y))
with(crop_ij, points(ph, humidity,
                     pch = 18, col = 'red'))
# Add fit line
with(crop_ij, abline(lm(humidity ~ ph),
                     lty = 4, lwd = 2, 
                     col = 'red'))

# Add another axis
axis(4, col.axis = "red")
## Add lab for this axis
mtext(side = 4, line = 2, "Humidity", col = 'red')

# Add legend
legend("topright", pch = c(20, 18), 
       col = c("blue", "red"), 
       legend = c("Rainfall", "Humidity"))
```

#### Histogram and boxplot

I make histogram and boxplot of temperature.

```{r, fig.align='center', fig.width=10, fig.height=5}
# Re-read the full dataset
crop_rmd <- read_csv(here('docs/homeworks/Crop_recommendation.csv'))
# set the layer, c(row, col)
par(mfrow = c(1, 2), mar = c(6, 2, 2, 2)) 
hist(crop_rmd$temperature,
     col = 'lightblue', border = 'purple',
     main = 'The distribution of temperature',
     xlab = 'Temperature')
boxplot(temperature ~ label, crop_rmd, 
        xlab = '', 
        ylab = 'Temperature',
        las = 2)
```

### `ggplot2`

Using `ggplot2` to make figure deserve another full credit course. Here I just list some examples corresponding to the figures made by base plot above.

#### Scatterplot

```{r, fig.align='center', fig.width=6, fig.height=4}
ggplot(crop_ij) +
  geom_point(aes(x = ph, y = rainfall), 
             color = 'blue') +
  geom_smooth(aes(x = ph, y = rainfall),
              method = "lm", se = FALSE, 
              color = 'blue', linetype = 2) + 
  geom_point(aes(x = ph, y = humidity*1.5), 
             color = 'red') +
  geom_smooth(aes(x = ph, y = humidity*1.5),
              method = "lm", se = FALSE, 
              color = 'red', linetype = 3) +
  # Add a second axis and specify its features
  scale_y_continuous(
    sec.axis = sec_axis(~.*(1 / 1.5), name = "Humidity")) + 
  labs(x = "ph",
       title = 'Relationship between ph, rainfall and humidity') +
  theme_bw() +
  theme(axis.text.y.right = element_text(color = "red"),
        axis.title.y.right = element_text(color = "red"),
        plot.title = element_text(color = "brown",
                                  hjust = 0.5))
```

#### Histogram and boxplot

```{r, fig.align='center', fig.width=10, fig.height=5}
# Re-read the full dataset
crop_rmd <- read_csv(here('docs/homeworks/Crop_recommendation.csv'))

p1 <- ggplot(crop_rmd) +
  geom_histogram(aes(x = temperature), 
                 fill = 'lightblue', color = 'purple') +
  labs(x = 'Temperature', title = 'The distribution of temperature') +
  theme_classic()
p2 <- ggplot(crop_rmd) +
  geom_boxplot(aes(x = label, y = temperature)) +
  labs(y = 'Temperature', x = '') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

plot_grid(p1, p2, labels = "AUTO")
```

#### Groups and facets in `ggplot2`

Now let's try to use groups and facets in `ggplot2`.

##### Groups
```{r, fig.align='center', fig.width=6, fig.height=4}
# Subset the dataset
set.seed(12)
crop_rmd <- crop_rmd %>% 
  filter(label %in% sample(unique(crop_rmd$label), 4))

## Groups for shape
ggplot(crop_rmd) +
  geom_point(aes(x = ph, y = rainfall,
                 shape = label)) +
  labs(x = 'ph', 
       y = 'Rainfall',
       title = 'Relationship between ph and rainfall') +
  theme_light()

## Groups for color
ggplot(crop_rmd) +
  geom_point(aes(x = ph, y = rainfall,
                 color = label)) +
  scale_color_brewer(palette = "Dark2") + 
  labs(x = 'ph', 
       y = 'Rainfall',
       title = 'Relationship between ph and rainfall') +
  theme_dark()
```

##### Facets

```{r, fig.align='center', fig.width=6, fig.height=6}
ggplot(crop_rmd) +
  geom_point(aes(x = ph, y = rainfall), color = 'red') +
  facet_wrap(~ label) +
  labs(x = 'ph', 
       y = 'Rainfall',
       title = 'Relationship between ph and rainfall') +
  theme_light() +
  theme(plot.title = element_text(color = "brown",
                                  hjust = 0.5))
```
