---
title: "Geospatial Analysis with R"
subtitle: Class 10
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

Review some concepts of geospatial data:
  - Vector vs. raster. And their common file types.
  - Geodatabase and attributes of vectors.
  - Basic spatial analysis.
  - Coordinate System. Geographic coordinate system and Cartesian coordinate system.
  - Projection and transformation.

---

## A few packages and tools

- GDAL
- `sf` package
- `raster` package
- `geospaar` package
- Others: `rgdal`, `rgeos` and `sp`.

---

## Vector

---

## Vector

Within this class, we will mainly use `sf` package to deal with vectors. For some cases, we need to use `sp` a bit.

**Common feature types**

- `POINT` and `MULTIPOINT`
- `LINESTRING` and `MULTILINESTRING`
- `POLYGON` and `MULTIPOLYGON`
- `GEOMETRYCOLLECTION`

**Common file types**

- ESRI Shapefile
- GeoJSON
- Idrisi
- `st_drivers(what = "vector")` to check all supported types

---

## Geodatabase and attributes of vectors

```{r, warning=F, message=F}
library(sf)
districts <- st_read(system.file('extdata/districts.shp', 
                             package = 'geospaar'),
                 quiet = T)
head(districts)
```

---

## Basic spatial operation/analysis

- Selection and classification
  
  - Non-spatial algebra of attributes
  - Spatial algebra, e.g. contain, adjacent
- Dissolve
- Buffer
- Overlay
  
  - Overlay
  - Clip, intersect, and union

---

## Basic spatial operation/analysis

```{r, warning=F, message=F, fig.align='center', fig.asp = 1, fig.width = 10, out.width = "40%"}
library(dplyr)
chama <- districts %>% filter(distName == 'Chama')
chama_touches <- districts %>% 
  slice(st_touches(chama, .) %>% unlist() %>% unique())
chama_touches_dsv <- chama_touches %>% 
  st_union() %>% st_buffer(0.1)
plot(districts$geometry)
plot(chama_touches_dsv, add = T, col = 'grey')
plot(chama$geometry, add = T, col = 'red')
plot(chama_touches$geometry, add = T, col = 'orange')
```

---

## Raster

We mainly use `raster` package to deal with raster data in this class.

```{r, warning=F, message=F, fig.align='center', fig.asp = 1, fig.width = 10, out.width = "40%"}
library(raster)
library(geospaar)
dem <- raster(system.file('extdata/demalb.tif',
                          package = 'geospaar'))
dem <- aggregate(dem, 20)
plot_noaxes(dem, legend = F)
```

---

## Raster drivers

- GeoTIFF (.tif)
- `st_driver(what = 'raster')`

---

## Basic spatial analysis

- Local functions
- Neighborhood, zonal and global functions
- Terrain analysis
- Spatial interpolation and prediction

---

## Basic spatial analysis

```{r, warning=F, message=F, fig.align='center', fig.asp = 0.5, fig.width = 10, out.width = "50%"}
districts <- districts %>% st_transform(crs = crs(dem))
districts_rst <- rasterize(as(districts, 'Spatial'), dem)
dem_mean_dst <- zonal(dem, districts_rst)
par(mfrow = c(1, 2))
plot_noaxes(dem, legend = FALSE)
plot_noaxes(districts_rst, legend = FALSE)
head(dem_mean_dst, 3)
```

---

## Coordinate Systems

- Geographic coordinate systems, e.g. WGS84

```{r, warning=F, message=F, echo=F, fig.align='center', fig.asp = 1, fig.width = 10, out.width = "50%"}
library(sf)
library(ggplot2)
library(dplyr)
library(mapproj)
library(rnaturalearth)

crs <- "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +datum=WGS84 +units=m +no_defs"
districts_sphere <- st_transform(districts, crs = crs)
sphere <- st_graticule(ndiscr = 10000, margin = 10e-6) %>%
  st_transform(crs = crs) %>%
  st_convex_hull() %>%
  summarise(geometry = st_union(geometry))
spdf_world <- ne_countries() %>% st_as_sf()
spdf_world %>% ggplot() +
  geom_sf(data = sphere, color = NA, fill = NA, size = 1.1) +
  geom_sf(fill = 'aliceblue', size = 1.1) +
  geom_sf(data = districts_sphere, col = NA, fill = 'red') +
  coord_sf(crs = crs) +
  theme_minimal()
```

---

## Coordinate Systems

- Projected coordinate reference system, e.g. UTM

```{r, warning=F, message=F, echo=F, fig.align='center', fig.asp = 0.6, fig.width = 14, out.width = "70%"}
spdf_world %>% ggplot() +
  geom_sf(fill = 'aliceblue', size = 1.1) +
  geom_sf(data = districts_sphere, col = NA, fill = 'red') +
  theme_minimal()
```

---

## Projection and transformation

- `st_crs` and `st_transform`
- `projectRaster`

---

## Next class

- Read and write vector data.
- Create geometries from scratch: points, linestrings, and polygons.
- Convert geometry types.
- Transform CRS.

---

## Homework

- Read Unit2 - Module 1
