---
title: "Geospatial Analysis with R"
subtitle: Class 18
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Exercise
- Raster intro, cont.


---
## Exercise

- Create a raster `r` that has the extent xmin = 25, xmax = 30, ymin = -12, ymax = -10, and a resolution of 0.1

```{r, eval=FALSE}
r <- raster(extent(....

```

- Populate it with randomly sampled values from between 1:100

```{r, eval=FALSE}
values(r) <- ....

```
- Create a stack `s` using r and r * 10
- Convert `s` to a brick `b`


---
## Raster to other types
```{r, eval=FALSE}
plot(b)
b[1:10]
b[1:ncell(b)]
as.matrix(b)
as.data.frame(b) %>% as_tibble()
```


---
## Pre-processing
- Aggregating/disaggregating
```{r, eval = FALSE}
aggregate(b, fact = 2) %>% plot
# aggregate(b, fun = min, fact = 2) %>% plot
# aggregate(b, fun = sd, fact = 2) %>% plot
plot(b)
disaggregate(b, fact = 2) %>% plot
disaggregate(b, fact = 2, bilinear = TRUE) %>% plot
```

- Masking
```{r, eval = FALSE}
data(chirps)
plot_noaxes(chirps[[1]])
chirpsz <- mask(chirps, districts)
plot_noaxes(chirpsz[[1]])
plot(st_geometry(districts), add = TRUE)
# rasterVis::levelplot(chirpsz[[1:5]])
plot_noaxes(chirpsz, nr = 3, nc = 4, maxnl = 12, mar = c(0, 0, 0, 0), 
            legend = FALSE, zlim = c(0, 50))
plot(chirpsz, nr = 3, nc = 4, maxnl = 12, mar = c(0, 0, 0, 0), 
     legend = FALSE, zlim = c(0, 50))
```

---

- Projecting
```{r, eval = FALSE}
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
chirpsz %>% projectRaster(., crs = crs(roads), res = 5000) -> chirpsz_alb
chirpsz_alb %>% print
```

- Rasterize/vectorize
```{r, eval = FALSE}
distsr <- districts %>% rasterize(., chirpsz)
distsr %>% plot_noaxes
# stack(distsr, chirpsz[[1:2]]) %>% plotRGB(stretch = "lin")
distsr_pol <- rasterToPolygons(distsr, dissolve = TRUE)
distsr_pol %>% st_as_sf %>% st_geometry %>% plot
distsr_pol %>% st_as_sf %>% slice(49) %>% plot(add = TRUE)
```

---
## Calculations

- raster algebra
- statistics
- z dimension stats


---
## Calculations

```{r, eval=FALSE}
rf1 <- chirpsz[[1]] + chirpsz[[2]]
rf2 <- chirpsz[[1]] * 5

mx <- cellStats(rf2, max)
zl <- c(0, mx)
par(mfrow = c(2, 2))
plot(chirpsz[[1]], zlim = zl)
plot(chirpsz[[2]], zlim = zl)
plot(rf1, zlim = zl)
plot(rf2, zlim = zl)

# dev.off()
plot(rf1 / 1000)
plot(((log10(rf1 + 1) * 10) / 5)^2)

```

- statistics
```{r, eval=FALSE}
# mean(c(1, 4, 10, NA), na.rm = TRUE)
cellStats(chirpsz[[1]], stat = mean)
cellStats(chirpsz[[1]], stat = range)
cellStats(chirpsz, stat = mean)
cellStats(chirpsz, stat = sum)
cellStats(chirpsz, stat = range)
cellStats(chirpsz, summary)
```

- z dimension stats
```{r, eval=FALSE}
plot(calc(chirpsz, mean))
plot(calc(chirpsz, sum))
plot(calc(chirpsz, range))
```
---

- neighborhoods

```{r, eval=FALSE}
raintot <- calc(chirpsz, stat = sum)
w <- matrix(1, nrow = 5, ncol = 5)
rainwindow <- focal(raintot, w = w, fun = mean, na.rm = TRUE)
rainwindow2 <- focal(raintot, w = w, fun = mean)
rainwindow3 <- focal(raintot, w = w, fun = sd)
par(mfrow = c(2, 2), mar = rep(0, 4))
plot_noaxes(raintot, legend = FALSE)
plot_noaxes(rainwindow, legend = FALSE)
plot_noaxes(rainwindow2, legend = FALSE)
plot_noaxes(rainwindow3, legend = FALSE)
```

---

- reclassify

```{r, eval=FALSE}
raincat <- cut(raintot, breaks = c(0, 50, 100, 200))
cols <- c("grey70", "lightblue", "lightgreen")
plot_noaxes(raincat, col = cols,
            legend = FALSE,  main = "Total Rainfall",
            mar = c(0, 0, 1, 0))
legend(x = "bottomright", legend = c("high", "medium", "low"),
       pch = 15, pt.cex = 3, col = rev(cols), bty = "n")
```

```{r, eval=FALSE}
rain <- cellStats(raintot, range)
rain_3 <- cbind(c(floor(rain[1]), 50, 100),
                 c(50, 100, ceiling(rain[2])), 1:3)
raincat2 <- reclassify(raintot, rain_3, include.lowest = TRUE)
plot_noaxes(raincat2, col = cols,
            legend = FALSE,  main = "Total Rainfall",
            mar = c(0, 0, 1, 0))
legend(x = "bottomright", legend = c("high", "medium", "low"),
       pch = 15, pt.cex = 3, col = rev(cols), bty = "n")

```



---
## Exercises

- Create a dummy raster (sampling from 1:100) using district 49 for extent and res of 0.1
- Reproject to Albers
- Disaggregate to 0.02 degrees (bilinear and NGB)
- Calculate the sum of rainfall from `chirps`
- Identify all areas of total rainfall > 10 mm
- Calculate the mean rainfall in Zambia for the 15th day in `chirps`



