---
title: "Geospatial Analysis with R"
subtitle: Class 12
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

### Today

- Homework
- More practice with reshaping/analysis/plotting 


---
### Data 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
set.seed(1)
price_weight_long <- tibble(
  year = 1951:2000, 
  price = runif(n = length(year), 20, 50),
  weight = (price * 10) * runif(n = length(year), 0.8, 1.2)
) %>% pivot_longer(
  cols = price:weight, names_to = "element", values_to = "value"
) %>% arrange(element)

```


---
### Homework

```{r, eval=FALSE}
price_weight_long %>% 
  group_by(element) %>% 
  summarize(mean = mean(value), sd = sd(value))

do.call(rbind, lapply(unique(price_weight_long$element), function(x){
  dat <- price_weight_long[price_weight_long$element == x, ]
  data.frame(element = x, mean = mean(dat$value), stdev = sd(dat$value))
}))

```

---
### A few more plots

- Let's do a scatter plot, with Price on the x axis and weight on the y axis

```{r, out.width = "60%", fig.align='center'}
price_weight_long %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = weight / price) %>%
  ggplot() + geom_point(aes(x = price, y = weight))
```

---

- Let's plot a line graph of Price (y axis) against year (x axis)

```{r, out.width = "55%", fig.align='center', fig.cap="Demo line plot"}
price_weight_long %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = weight / price) %>%
  ggplot() + geom_line(aes(x = year, y = price))
```

---

- Let's calculate a histogram of wt_price
```{r, out.width = "60%", echo=FALSE, fig.align='center'}
price_weight_long %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = weight / price) %>%
  ggplot() +
  geom_histogram(aes(x = wt_price), color = "black", fill = "blue", bins = 20)
```
---
### Dataset

```{r, message=FALSE, warning=FALSE}
set.seed(1)
price_weight2 <- tibble(
  year = rep(1951:2000, 2), 
  group = rep(sample(letters[1:5], 50, replace = TRUE), 2), 
  value = runif(n = 100, min = 50, max = 100),
  element = c(rep("Price", 50), rep("Weight", 50))
)
```
- Let's do a scatter plot, with Price on the x axis and weight on the y axis, with one series for each group

```{r, out.width = "60%", echo=FALSE, fig.align='center'}
price_weight2 %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = Weight / Price) %>%
  ggplot() + geom_point(aes(x = Price, y = Weight, color = group))
```

---
- Let's plot a line graph of Price (y axis) against year (x axis)

```{r, out.width = "70%", echo=FALSE, fig.align='center'}
price_weight2 %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = Weight / Price) %>%
  ggplot() + geom_line(aes(x = year, y = wt_price, color = group))
```

---

- Let's calculate a histogram of wt_price
```{r, out.width = "70%", echo=FALSE, fig.align='center'}
price_weight2 %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = Weight / Price) %>%
  ggplot() +
  geom_histogram(aes(x = wt_price, color = group), bins = 20)
```
---
- Do the same, but let's make a histogram for each group.
```{r, out.width = "70%", echo=FALSE, fig.align='center'}
price_weight2 %>%
  pivot_wider(names_from = element, values_from = value) %>%
  mutate(wt_price = Weight / Price) %>%
  ggplot() +
  geom_histogram(aes(x = wt_price), color = "black", fill = "blue", bins = 20) +
  facet_wrap(facets = ~group)
```


---

## More practice
### Part 1
- read in the three `geospaar` FAO datasets into a list using `lapply`
- Reshape each of the three datasets by using `pivot_wider` to put harvested area into a new variable **harv_area** and production into **prod**, with the other variables being **crop**, **country**, and **year**. Necessary functions: `%>%`, `select`, `pivot_wider`, `rename`. 
- Combine all these into a single data.frame `crops_df`
- Best: combine the previous two points using `lapply` and `do.call(rbind)`
- Okay: Do this separately for each dataset and then combine with `rbind`

---

- read in the three `geospaar` FAO datasets into a list using `lapply`

```{r, eval=FALSE}
library(geospaar)
fs <- dir(system.file("extdat", package = "geospaar"), 
          pattern = "FAOSTAT", full.names = TRUE)
crops <- lapply(fs, readr::read_csv)
```

Should see...
```
[[1]]
# A tibble: 228 x 14
   `Domain Code` Domain `Area Code` Area  `Element Code` Element
   <chr>         <chr>        <dbl> <chr>          <dbl> <chr>  
 1 QC            Crops          202 Sout…           5312 Area h…
 2 QC            Crops          202 Sout…           5312 Area h…
# … with 218 more rows, and 8 more variables: `Item Code` <dbl>,
...

[[2]]
# A tibble: 228 x 14
   `Domain Code` Domain `Area Code` Area  `Element Code` Element
   <chr>         <chr>        <dbl> <chr>          <dbl> <chr>  
 1 QC            Crops          202 Sout…           5312 Area h…
 2 QC            Crops          202 Sout…           5312 Area h…

...

[[3]]
# A tibble: 228 x 14
   `Domain Code` Domain `Area Code` Area  `Element Code` Element
   <chr>         <chr>        <dbl> <chr>          <dbl> <chr>  
 1 QC            Crops          202 Sout…           5312 Area h…
 2 QC            Crops          202 Sout…           5312 Area h…
...
```
---

- Reshape each of the three datasets in `crops` by using `pivot_wider` to put harvested area into a new variable **harv_area** and production into **prod**, with the other variables being **crop**, **country**, and **year**. 
- Combine all these into a single data.frame `crops_df`


```{r, eval=FALSE}
crops_df <- do.call(rbind, lapply(crops, function(x) {
  x %>% select(Item, Area, Element, Year, Value) %>% 
    pivot_wider(names = Element, values = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = Area harvested, prod = Production)
}))
crops_df
```

Should see:

```
> crops_df
# A tibble: 342 x 5
   crop  country       year harv_area    prod
   <chr> <chr>        <dbl>     <dbl>   <dbl>
 1 Maize South Africa  1961   4118000 5293000
 2 Maize South Africa  1962   4341000 6024000
 3 Maize South Africa  1963   4339000 6127000
 4 Maize South Africa  1964   4433000 4310000
 5 Maize South Africa  1965   4290000 4608000
 6 Maize South Africa  1966   4241000 5161000
 7 Maize South Africa  1967   4589000 9802000
 8 Maize South Africa  1968   4728000 5358000
 9 Maize South Africa  1969   4387000 5378000
10 Maize South Africa  1970   4418000 6179000
# … with 332 more rows
```

---
#### Pipeline example
```{r, eval = FALSE}
fs <- dir(system.file("extdata", package = "geospaar"), 
          pattern = "maize", full.names = TRUE)
dat <- readr::read_csv(fs) %>% 
  dplyr::select(Item, Area, Element, Year, Value) %>% 
  pivot_wider(names_from = Element, values_from = Value) %>% 
  rename(harv_area = `Area harvested`, prod = Production, country = Area) %>% 
  mutate(yield = prod / harv_area) %>% 
  dplyr::select(country, harv_area, prod, yield) %>% 
  group_by(country) %>% 
  summarise_all(list(mu = mean, sdev = sd))
```

---
### Part 2 - plotting
Using `crops_df`:
- Using `ggplot2`, make a line plot of the annual maize production by country, using red for Zambia and blue for South Africa
- Use `ggplot2` to make three side-by-side line plots of production per year for all three crops by country (red for Zambia, blue South Africa). You need `facet_grid` for this. Do the same for yield (you will need to a new column *yield*)
- Plot the linear regression fit between year wheat yields, using `ggplot` and `geom_smooth`, with `method = "lm"`. Make a separate series and fit for South African and Zambian yields.

---
- Use `ggplot2` to make three side-by-side line plots of production for all three crops by country (red for Zambia, blue South Africa). Do the same but for yield. You need `facet_grid` for this.

```{r, eval=FALSE}
ggplot(crops_df) + 
  geom_line(aes(x = year, y = prod, color = country)) +
  facet_grid(crop)
```

