---
title: "Geospatial Analysis with R"
subtitle: Class 13
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
## Election Results
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width=8, fig.height=5}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")

cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>%
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% 
  dplyr::select(-other, -delres) %>%
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab(NULL) + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

---
```{r, eval = FALSE}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")
cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>%
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% select(-other, -delres) %>%
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab("") + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2))
```
---

## Today
- Pipeline and plotting exercises (in small groups)

---

## Data reading and pipelines

```{r, eval=FALSE}
library(geospaar)
fs <- dir(system.file("extdata", package = "geospaar"), 
          pattern = "FAOSTAT", full.names = TRUE)
crops <- lapply(fs, readr::read_csv)
```

---

```{r, eval=FALSE}
crops_df <- do.call(rbind, lapply(crops, function(x) {
  x %>% select(Item, Area, Element, Year, Value) %>% 
    pivot_wider(names_from = Element, values_from = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = `Area harvested`, prod = Production)
}))
crops_df
```

---

#### Analysis example 

Average maize yield by country
```{r, eval=FALSE}
crops_df %>% 
  filter(crop == "Maize") %>% 
  mutate(yield = prod / harv_area) %>% 
  group_by(country) %>% 
  summarize(yield = mean(yield))
```

---

#### Pipeline example 1

Mean and standard deviation of maize yield by country
```{r, eval = FALSE}
fs <- dir(system.file("extdata", package = "geospaar"), 
          pattern = "maize", full.names = TRUE)
dat <- readr::read_csv(fs) %>% 
  dplyr::select(Item, Area, Element, Year, Value) %>% 
  pivot_wider(names_from = Element, values_from = Value) %>% 
  rename(harv_area = `Area harvested`, prod = Production, country = Area) %>% 
  mutate(yield = prod / harv_area) %>% 
  dplyr::select(country, harv_area, prod, yield) %>% 
  group_by(country) %>% 
  summarise_all(list(mu = mean, sdev = sd))
```

---
#### Pipeline example 2

Creating `crops_df` in one step
```{r, eval=FALSE}
library(geospaar)
fs <- dir(system.file("extdata", package = "geospaar"), 
          pattern = "FAOSTAT", full.names = TRUE)
crops_df <- do.call(rbind, lapply(fs, function(x) {
  readr::read_csv(x) %>% 
    select(Item, Area, Element, Year, Value) %>% 
    pivot_wider(names_from = Element, values_from = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = `Area harvested`, prod = Production)
}))

```

---
#### Exercises 

##### 1. Adapt the code from the Analysis example to calculate mean wheat yields for Zambia and South Africa. 

##### 2. Adapt the code from the Analysis example to calculate mean yields of each of the three crops for each country. 

##### 3. Adapt the code from Pipeline example 2 so that the resulting `crops_df` includes a yield variable. The relevant code you need to do this is found in Pipeline example 1. 


---
### Plotting exercises

Using `crops_df`:
- Using `ggplot2`, make a line plot of the annual maize production by country. Challenge: make Zambia's line red and South Africa's blue

Starter code:
```{r, eval=FALSE}
ggplot(crops_df) + 
  geom_line(aes(x = , y = , ))
```

---

- Use `ggplot2` to make three side-by-side line plots of yield per year for all three crops by country (red for Zambia, blue South Africa). You need `facet_wrap` or `facet_grid` for this. 

Starter code
```{r, eval=FALSE}
crops_df %>% 
  mutate() %>% 
ggplot() + 
  geom_line(aes(x = ??, y = ??, color = ??)) + ??
```

---

- Plot the linear regression fit between year wheat yields, using `ggplot` and `geom_smooth`, with `method = "lm"`. Make a separate point series and fit for South African and Zambian yields.

Starter code

```{r, eval=FALSE}
crops_df %>%
  mutate(?) %>% 
  ? %>% 
ggplot() + 
  geom_??(aes(x = , y = , ?)) + 
  geom_smooth(aes(x = , y = , ?), ?)

```



