---
title: "Geospatial Analysis with R"
subtitle: Class 13
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Units
- Some operation practice
- Visualization

---

## Units

`units::set_units()`

```{r, eval=F}
# Examples of usage
10 %>% units::set_units('km2')
st_area(africa) %>% units::set_units('ha')
st_buffer(africa, 10 %>% 
            units::set_units('miles') %>% 
            units::set_units('m'))

# Remove units
set_units(x, NULL)
```

---

## Operation practice 1

Recreate your building, and select out the same building from `osmdata` (name it `building_osm`). Then:
- Get the intersection (`bld_itst`) of your building and `building_osm`.
- Get the difference (`dif_bld_osm`) between your building and `building_osm`.
- Get the difference (`dif_osm_bld`) between `building_osm` and your building.
- Get the union (`bld_union`) of your building and `building_osm`.
- Get the bounding box (`bld_bbox`) of the union and convert it to `sf`.
- Plot `bld_bbox`, `dif_bld_osm`, `dif_osm_bld`, and `bld_itst` one by one using different colors. Hint: use `add = T` to plot one layer on top of another.
- Calculate the area of `bld_union` in ha, and print it out.
- Calculate the perimeter of the building that you drawn in km and print it out.

---

```{r, echo=F, warning=F, message=F}
# Get buildings from `osmdata`
library(dplyr)
library(sf)
library(osmdata)
q <- opq(bbox = c(-71.82587, 42.24956,
                  -71.82118, 42.25368))
buildings <- q %>% 
  add_osm_feature(key = 'building') %>%
  osmdata_sf() %>% `[[`('osm_polygons') %>% 
  filter(!is.na(name))

building_osm <- buildings %>% filter(name == 'Jonas Clark Hall')

jonas <- rbind(c(42.25073971066906, -71.82325961386029),
                c(42.2507777725669, -71.82318350005451),
                c(42.25078223970596, -71.82318551171117),
                c(42.25084875040589, -71.82305810678872),
                c(42.25083237091165, -71.8230406724309),
                c(42.25091029755825, -71.8228891276284),
                c(42.25099269123274, -71.82278452148155),
                c(42.250988720455716, -71.82277915706375),
                c(42.251034384379025, -71.82269466747479),
                c(42.25113415000695, -71.8227925681005),
                c(42.25115301115209, -71.82275904048932),
                c(42.25120463109945, -71.82281000245831),
                c(42.25118676266091, -71.82284621227836),
                c(42.251274119423435, -71.82293539572409),
                c(42.251233101414996, -71.82301271143761),
                c(42.251168576503304, -71.82312268200226),
                c(42.251181481490924, -71.82313743415118),
                c(42.251096606328346, -71.8232997078066),
                c(42.25108370132338, -71.82329099062768),
                c(42.250977483105125, -71.82349819126475),
                c(42.25087821357723, -71.82340431395349),
                c(42.250858856001734, -71.8234391826691),
                c(42.25081666639184, -71.8233969378754),
                c(42.25083503128357, -71.82335938695088)) %>%
  `[`(, ncol(.):1) %>% st_linestring() %>% # create sfg
  st_cast('POLYGON') %>%
  st_sfc() %>% st_set_crs(4326) %>% # create sfc
  st_sf() %>% # build sf
  mutate(name = "Jonas Clark Hall")

# operation
bld_itst <- st_intersection(jonas, building_osm)
dif_bld_osm <- st_difference(jonas, building_osm)
dif_osm_bld <- st_difference(building_osm, jonas)
bld_union <- st_union(jonas, building_osm)
bld_bbox <- st_bbox(bld_union) %>% 
  st_as_sfc() %>% st_sf()

# plot
plot(bld_bbox$geometry, col = 'lightblue')
plot(dif_osm_bld$geometry, col = 'red', add = T)
plot(dif_bld_osm$geometry, col = 'blue', add = T)
plot(bld_itst$geometry, col = 'yellow', add = T)
```

The area of the union is `r st_area(bld_union) %>% units::set_units('ha')` ha. The perimeter of the building is `r st_length(jonas) %>% units::set_units('km')` km.

---

- `st_sample`
- `st_contains`
- `st_nearest_feature`
- `st_nearest_points`
- `st_convex_hull`
- `st_crop`

Who wants to explain how to use them with an example?

---

## Operation practice 2

- Prepare the buildings and Clark campus polygon based on shared code.
- Use Clark campus polygon to get buildings that are inside of the campus polygon (`blds_inside`).
- Get the nearest building (`bld_nst_jch`) to Jonas Clark Hall in `blds_inside`.
- Get 10 samples (`bld_samples`) from each polygon of `blds_inside`.

```{r, eval=F, echo=F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396),
                # repeat the first point
                c(-71.82117553759312, 42.25084734394512))
clark_campus_ply <- st_polygon(list(coords)) %>% # create sfc
  st_sfc() %>% st_set_crs(4326) %>%  # create sfc
  st_sf() %>% mutate(name = 'Clark main campus')

buildings <- opq(bbox = st_bbox(clark_campus_ply)) %>% 
  add_osm_feature(key = 'building') %>%
  osmdata_sf() %>% `[[`('osm_polygons') %>% 
  filter(!is.na(name))

# Operations
blds_inside <- buildings %>% 
  slice(st_contains(clark_campus_ply, buildings) %>% 
          unlist())

bld_nst_jch <- blds_inside %>% 
  filter(name != 'Jonas Clark Hall') %>% 
  slice(st_nearest_feature(
    blds_inside %>% 
      filter(name == 'Jonas Clark Hall'), .))

bld_samples <- blds_inside %>% 
  st_sample(seed = 10,
            size = rep(10, nrow(.)),
            exact = T) %>% 
  st_sf()
```

---

## Plot them using `ggplot2`

The main function in `ggplot2` to plot `sf` is `geom_sf`. All `ggplot2` syntax that we learned before still work.

```{r, eval = F}
library(ggplot2)
ggplot(clark_campus_ply) +
  geom_sf(fill = 'pink') +
  geom_sf(data = blds_inside, aes(fill = name)) + 
  scale_fill_manual(values = rainbow(nrow(blds_inside))) +
  geom_sf(data = blds_inside %>% 
      filter(name == 'Jonas Clark Hall'),
      fill = 'red') +
  # Show label in figure
  geom_sf_label(data = blds_inside %>% 
      filter(name == 'Jonas Clark Hall'), 
      aes(label = name)) +
  geom_sf(data = bld_nst_jch, fill = 'blue') +
  geom_sf(data = bld_samples, pch = 20, color = 'grey') +
  theme_bw() + ggtitle('Clark campus') + 
  theme(legend.position = 'none')
```

---

## Interactive maps

- `leaflet`
- `mapview`

Go to class homepage and download `class13_clark_blds_teamwork.R`.

---

## Next class

- `sp` and convert between `sp` and `sf`.
- Read and write raster data.

---

## Homework

- Assignment 4 due this week. 
- Make an interactive map of whatever that you are interested in. You will introduce your map to us in next class. Here is an example:

  [MA map](https://agroimpacts.github.io/geospaar/class13_ma_map.html)
