---
title: "Geospatial Analysis with R"
subtitle: Class 14
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message=FALSE, eval=FALSE, warning=FALSE}
library(leaflet)
library(sf)
library(dplyr)
farmersn <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  readr::read_csv() %>% group_by(uuid) %>% 
  summarize(x = mean(x), y = mean(y), n = n()) %>% 
  st_as_sf(coords = c("x", "y"), crs = 4326)
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
farmersn <- st_join(farmersn, districts, left = FALSE)
bb <- unname(st_bbox(districts))
xy <- st_centroid(districts) %>% st_coordinates() %>% 
  bind_cols(name = districts$distName, .)
slist <- list("color" = "white")
label_opts <- labelOptions(noHide = TRUE, style = slist, direction = 'center',
                           textOnly = TRUE, textsize = "5px")
qpal <- colorQuantile("RdYlBu", farmersn$n, n = 5)

m <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  fitBounds(bb[1], bb[2], bb[3], bb[4]) %>% 
  addPolygons(data = districts, fill = FALSE, color = "white", 
              group = "Districts", weight = 1) %>% 
  addCircles(data = farmersn, popup = as.character(farmersn$n), 
             color = ~qpal(n), group = "Farmers") %>% 
  addLabelOnlyMarkers(xy$X, xy$Y, label = xy$name, group = "Names",
                      labelOptions = label_opts) %>%
  addLayersControl(overlayGroups = c("Districts", "Names", "Farmers"),
                   options = layersControlOptions(collapsed = FALSE,
                                                  autoZIndex = FALSE))
```

```{r, eval=FALSE, echo=FALSE}
library(htmlwidgets) 
saveWidget(m, "docs/figures/zambialeaflet2.html")
```

---

<iframe seamless src="figures/zambialeaflet2.html" width="100%" 
height="500"></iframe>

---
### Today
- Projects
- Review of `sf` 
- Spatial Operations


---
### Projects
- [List](https://agroimpacts.github.io/geospaar/projects.html#project-list)
- [USF Nightlights](https://github.com/agroimpacts/USFlite/blob/main/external/notebooks/nightlights-part2.md)

"That is looking at public housing in New York. Students looked at it for Boston and San Fran last semester.  It would be great to add other cities, but maybe public housing isn’t available so other variables could be looked at instead of public housing, like demographic characteristics. Basically looking to see whether there is more brightness in similar neighborhoods that might indicate more floodlighting or other factors that make it brighter (perhaps less tree cover, etc)."


---
### Review
-read/write
-projections
-plot
-area/length


---
### Reading
- Create spatial points from csv
```{r, message=FALSE, }
library(tidyverse)
library(sf)
## read in csv
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() 

## convert tibble to 'sf'
farmers_sf <- st_as_sf(farmers, coords = c("x", "y"), crs = 4326)
```




---

### Reading
- Reading shapefiles
- Both `st_read` and `read_sf` work

```{r, message = F}
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
roads <- st_read(system.file("extdata/roads.shp", package = "geospaar"))
```

---
### Writing
- `st_write` works as expected
- Use '.geojson' (only one file)
```{r, eval = F}
#st_write(districts, "districts_out.shp")
st_write(districts, "districts.geojson")
```


---
### Projections (crs)
- Geographic coordinates are: `4326`
```{r, message = F}
st_crs(farmers_sf) <- 4326
farmers_sf %>% st_crs()
```


---
## projections (use existing)
- you can set projections based on another layer
- but the two layers MUST have the same crs (e.g. geographic)
```{r, message = F}
farmers <- read_csv(system.file("extdata/farmer_spatial.csv", 
                                package = "geospaar"))
farmers <- st_as_sf(farmers, coords = c("x", "y"))

st_crs(farmers) <- st_crs(districts) ## set crs

farmers %>% st_crs() ## print new crs
```

---
## projections (transform)
- Use `st_transform()` to project from one crs to another
```{r, message = F}
roads <- st_transform(x = roads, crs = st_crs(districts))
st_crs(roads)
```


---
## features from scratch
```{r, message = F}
pt <- st_point(x = c(28, -14))
pt
#> POINT (28 -14)
#
# #2
pts <- st_multipoint(x = cbind(x = c(27.5, 28, 28.5), y = c(-14.5, -15, -15.5)))
pts
#> MULTIPOINT ((27.5 -14.5), (28 -15), (28.5 -15.5))
#
# #3
sline <- st_linestring(cbind(x = c(27, 27.5, 28), y = c(-15, -15.5, -16)))
sline
#> LINESTRING (27 -15, 27.5 -15.5, 28 -16)
#
# #4
pol <- st_polygon(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5), 
                             y = c(-15.5, -16.5, -17, -16, -15.5))))
pol
```

---
### Plotting
- Base `plot()` works with `sf`, use `add = TRUE`
```{r, message = F}
par(mar = c(0, 0, 0, 0))
plot(districts %>% st_geometry(), col = "grey")
plot(pt, add = TRUE, col = "red", pch = 20)
plot(pts, add = TRUE, col = "blue", pch = 20)
plot(sline, add = TRUE, col = "cyan", lwd = 2)
plot(pol, add = TRUE, col = "orange", lwd = 2)
```


---
### Plotting (ggplot)
- Use `geom_sf()`
```{r, message = F}
ggplot(districts) + geom_sf() +
  geom_sf(data = roads, col = "red") + 
  geom_sf(data = farmers_sf, col = "blue")
```

---
### Area and Length
- `st_area()`, `st_length`. Default units are in meters
- How many square m in a square km?
```{r, message = F}
dist_areas <- districts %>% st_area()
dist_areas
dist_areas %>% units::set_units("km^2")
```

---
## add area as column
```{r, message = F}
districts <- districts %>% 
  mutate(area = st_area(.) %>% units::set_units("km^2") %>% as.numeric()) 
districts
```

---
## Plot districts (exercise)
- Use `geom_sf()` to plot area
- Fill districts based on area column.
- Use `scale_fill_continuous` , with `type = 'viridis'`
```{r, message = F}

```

---
## ggplot steps
- Start with `ggplot()`. You can include data if you're only using one data set.
- Add plot type (`geom_point`, `geom_line`, `geom_sf`, `geom_histogram`)
- Add aesthetics in `aes()`. These are columns used for plotting. 
- For `geom_point`, `geom_line` you need 'x' and 'y' aesthetics. 
- For `geom_sf`, spatial context used for 'x' and 'y'. But you might want 'fill' or 'color'



```{r, message = F}

```

---
## Answer
```{r, message = F}
ggplot(districts) + 
  geom_sf(aes(fill = area )) +
  scale_fill_continuous(type = 'viridis')
```


---
## Calculate length (exercise)
- Use `st_length()` to add colummn 'road_length' to `roads`
- 'road_length' should be in km
- Use ggplot to plot districts with black outline (color), and empty fill (fill = NA)
- Then plot roads based on length (use `scale_color_continuous()` )
- Plot roads with line width 2 (`lwd = 2`)
- Use `theme_bw()`

```{r, message = F}

```


---
## Answer
```{r, message = F}
roads <- roads %>% 
  mutate(road_length = st_length(.) %>% units::set_units("km") %>% as.numeric()) 
ggplot() +
  geom_sf(data = districts, col = 'black', fill = NA)  +
  geom_sf(data = roads, aes(col = road_length), lwd = 2) + 
  scale_color_continuous(type = 'viridis') + theme_bw()

```

---

### Spatial Operations
Data setup
```{r, message = F}
coords <- cbind("x" = c(25, 25.1, 26.4, 26.4, 26.4, 25), 
                "y" = c(-15.5, -14.5, -14.6, -14.8, -15.6, -15.5))
#
# #2
pol <- st_polygon(x = list(coords)) %>% st_sfc %>% st_sf(ID = 1, crs = 4326)
roads_rpj <- st_transform(roads, crs = st_crs(districts))

```

---
### Plotting
```{r, message = F}
par(mar = rep(0, 4))
plot(st_geometry(districts), col = "grey")
plot(pol, col = "purple", add = TRUE)
plot(roads_rpj, col = 'red', add = TRUE)
```


---
## `st_intersects`
- Similar to "Select by Spatial Location"
- Finds indices of second object that intersect each element of first object
```{r, message = F}
st_intersects(x = roads_rpj, y = districts)
```

---
## `st_intersects`
```{r, message = F}
highway <- roads_rpj[401,]
dists_int <- districts %>% slice(st_intersects(x = highway, y = districts)[[1]])
#> although coordinates are longitude/latitude, st_intersects assumes that they are planar
dists_int
plot(st_geometry(dists_int), col = 'grey')
plot(highway, col = 'red', add = TRUE)
```


---
## `st_intersection()`
- finds intersecting pairs between two objects
- Each row in result is one "intersection object" (i.e. an object from x and y that intersect)
```{r, message = F}
pol_int_dists <- st_intersection(x = pol, y = districts)

par(mar = rep(0, 4))
plot(st_geometry(districts), col = "grey")
plot(st_geometry(pol_int_dists), col = rainbow(n = nrow(pol_int_dists)), 
     add = TRUE)
```

---
## `st_difference`
- Similar to "erase" in ArcGIS
- Returns first object without second object. 
```{r, message = F}
d1 <- st_difference(x = districts, y = pol)
d2 <- st_difference(x = pol, y = districts)

par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
d1 %>% st_geometry %>% plot(col = "grey")
d2 %>% st_geometry %>% plot(col = "grey")
```


---
## `st_union`
- Combines objects without resolving borders. 
- Really takes all possible combinations of first and 2nd object
- Switching order of inputs changes column order
```{r, message = F}
uni1 <- st_union(x = pol, y = districts)
uni1

uni2 <- st_union(x = districts, y = pol)
uni2

par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
plot(uni1[2], reset = FALSE, main = "Union of pol with districts")
plot(uni2[1], reset = FALSE, main = "Union of districts with pol")
```

---
## `st_buffer()`
- Best to use projected coordinates (e.g. Albers)
```{r, message = F}
## project everything to Albers
farmers_alb <- farmers_sf %>% st_transform(st_crs(roads))
long_roads <- roads %>% filter(as.numeric(st_length(.)) > 500000)
districts_alb <- districts %>% st_transform(st_crs(roads))

roads_buff_20km = st_buffer(long_roads, 20000)
plot(st_geometry(districts_alb), col = "grey")
plot(st_geometry(roads_buff_20km), col = "blue", add = TRUE)
plot(st_geometry(long_roads), col = "red", add = TRUE)
```

---
## `st_sample`

```{r, message = F}
par(mar = rep(0, 4))
districts %>% st_union %>% plot(col = "grey")
set.seed(1)
districts %>% st_union %>% st_sample(size = 100, exact = TRUE) %>% 
  plot(pch = 20, add = TRUE)
```


---
## Exercise
Find all districts within 50 km of sampled points. 

```{r, message = F}

```

---
## Exercise
Of the sampled farmers, how many are within 100km of one of the "long roads"?
```{r, message = F}

```


---
### Further Exercises

#### Round 1
- Calculate the length of roads in kilometers
- Buffer roads > 100 km by 30 kilometers, save as new object
- Buffer roads > 100 km 10 kilometers, save as new object
- Take the difference between the 30 km and 10 km buffers, i.e. erase/difference the two
- Plot the resulting difference using `ggplot` with the difference in red over Zambia's districts

```{r, eval=FALSE}
# code to adapt
roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
  filter(km > 200) %>% 
  st_buffer(dist = 20000) %>% 
  st_transform(crs = 4326)
st_difference(larger_object, smaller_object)
ggplot() + geom_sf

```

---

#### Round 2
- Randomly sample 100 points within the 30 km road buffer (use a seed of 1)
- Plot the 30 km buffer (red) the points in it (blue) over the districts

```{r, eval=FALSE}
buffered_object %>% st_sample()
```

---
##HOMEWORK
- Review all of Unit 2, Module 1
- Assignment 4 due next Tuesday (3/21) 
- It is long, don't wait!
