---
title: "Geospatial Analysis with R"
subtitle: Class 16
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Homework from last class
- More spatial operations
- Exercises


---
### Data

```{r, eval=FALSE}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sf(coords = c("x", "y"))
districts <- st_read(system.file("extdata/districts.shp", package = "geospaar"))
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))

set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))
farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))
```

---


### Homework

- Select the smallest district in Zambia from `districts3`, and convert that
to centroid
```{r, eval = FALSE}
smallest_centroid <- districts3 %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter(area == min(area)) %>% st_centroid()
ggplot() + geom_sf(data = districts3) + 
  geom_sf(data = smallest_centroid, col = "red", cex = 0.1)
```

- Create a buffer of width 0.5 around that
```{r, eval = FALSE}
# sf_use_s2(use_s2 = TRUE)
sf_use_s2(use_s2 = FALSE)
smallest_buffer <- smallest_centroid %>% st_buffer(dist = 0.5)
ggplot() + geom_sf(data = smallest_centroid, col = "red") + 
  geom_sf(data = smallest_buffer, col = "blue", fill = "transparent") 
```

---
- Use the buffer to select the districts from `districts3` that intersect with it
```{r, eval = FALSE}
dists_in_buff <- districts3 %>% 
  filter(st_intersects(., smallest_buffer, sparse = FALSE))
ggplot(districts3) + geom_sf() +
  geom_sf(data = dists_in_buff, fill = "red") + 
  geom_sf(data = smallest_buffer, fill = "blue") 
  
```

---
- Use the buffer to do an intersection with `districts3`
```{r, eval=FALSE}
buff_int_dists <- st_intersection(smallest_buffer, districts3)
ggplot(districts3) + geom_sf() +
  geom_sf(data = buff_int_dists, fill = "red")
ggplot() +
  geom_sf(data = buff_int_dists, aes(fill = distName.1))
```

---
### More operations

- Fancy buffering
```{r, eval=FALSE}
roads20 <- roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
  filter(km > 200) %>% 
  st_buffer(dist = 20000) %>% 
  st_transform(crs = 4326)

ggplot(districts3) + geom_sf() + 
  # geom_sf(data = roads20, fill = "red")
  geom_sf(data = st_union(roads20), fill = "red")

st_union(roads20) %>% st_cast("POLYGON") %>% st_as_sf() %>% 
  mutate(ID = as.character(1:nrow(.))) %>% 
  # ggplot() + geom_sf(aes(fill = as.factor(ID)))
  ggplot() + geom_sf(aes(fill = ID))

ggplot(districts3) + geom_sf(aes(fill = distName))
districts3 %>% mutate(ID = 1:nrow(.)) %>% 
  ggplot() + geom_sf(aes(fill = ID))
```

- Differencing
```{r, eval=FALSE}
dif <- st_difference(districts3, smallest_buffer)
ggplot(dif) + geom_sf(fill = "grey50")

dif2 <- st_difference(smallest_buffer, districts3)
ggplot(dif2) + geom_sf(fill = "grey50")
```

- Unioning

```{r, eval=FALSE}
uni <- st_union(dif)
ggplot(uni) + geom_sf(fill = "grey50")
uni2 <- st_union(dif, smallest_buffer)
ggplot(uni2) + geom_sf(fill = "grey50")
uni3 <- st_union(dif, buff_int_dists)
ggplot(uni3) + geom_sf(fill = "grey50")

uni3
```

---
- Sampling

```{r, eval=FALSE}
# simple within extent
set.seed(1)
sample100 <- districts3 %>% filter(n > 0) %>% 
  st_sample(size = 100, exact = TRUE)
ggplot(districts3) + geom_sf() + 
  geom_sf(data = districts3 %>% filter(n > 0), fill = "lightblue") + 
  geom_sf(data = sample100, size = 0.5, col = "red")

# stratified by polygon
ndists <- districts3 %>% filter(n > 0) %>% nrow(.)
strat_sample <- districts3 %>% filter(n > 0) %>% 
  st_sample(size = rep(5, ndists), exact = TRUE)

ggplot(districts3) + geom_sf() + 
  geom_sf(data = districts3 %>% filter(n > 0), fill = "lightblue") + 
  geom_sf(data = strat_sample, size = 0.5, col = "red")
```

---
### Exercises

#### Round 1
- Calculate the length of roads in kilometers
- Buffer roads > 100 km by 30 kilometers, save as new object
- Buffer roads > 100 km 10 kilometers, save as new object
- Take the difference between the 30 km and 10 km buffers, i.e. erase/difference the two
- Plot the resulting difference using `ggplot` with the difference in red over Zambia's districts

```{r, eval=FALSE}
# code to adapt
roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
  filter(km > 200) %>% 
  st_buffer(dist = 20000) %>% 
  st_transform(crs = 4326)
st_difference(larger_object, smaller_object)
ggplot() + geom_sf

```

---

#### Round 2
- Randomly sample 100 points within the 30 km road buffer (use a seed of 1)
- Plot the 30 km buffer (red) the points in it (blue) over the districts

```{r, eval=FALSE}
buffered_object %>% st_sample()
```














