---
title: "Geospatial Analysis with R"
subtitle: Class 18
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Parallelization
- Exercises from last class
- New exercises (teams)

---
## Data

```{r, warning=FALSE, message=FALSE, results='hide'}
library(geospaar)
districts <- system.file("extdata/districts.shp", package = "geospaar") %>%
  st_read
roads <- system.file("extdata/roads.shp", package = "geospaar") %>% st_read

data(chirps)
chirpsz <- mask(chirps, districts)
```

---


### Results from homework
- From `chirpsz`, create a categorical raster (`raincat`) containing legend with "low", "medium", "high" rainfall from `raintot`, wherein the low is areas with rainfall < 50 mm, medium is 50-100 m, and high is rainfall > 100 mm

```{r, eval=FALSE}
raintot <- calc(chirpsz, sum)
raincat <- cut(raintot, breaks = c(0, 50, 200))
plot_noaxes(raincat, col = c("grey70", "lightblue"))
```

- Calculate the local standard deviation of `raintot` in a 5X5 moving window. Remove NAs from the calculation so the boundaries are not shrunk

```{r, eval=FALSE}
# w <- matrix(1, nrow = 3, ncol = 3)
w <- matrix(1, nrow = 9, ncol = 9)
plot(raintot)
plot(focal(raintot, w = w, fun = mean, na.rm = TRUE))
plot(focal(raintot, w = w, fun = mean))
```
- Calculate the average elevation and slope in each district of Zambia
---


---
### mapView
```{r, eval = FALSE}
library(mapview)
mapview(districts)
viewRGB(x = chirpsz[[1:3]])
mapView(districts, alpha.regions = 0, legend = FALSE) + mapview(raintot) 

kili_data <- system.file("extdata", "kiliNDVI.tif", package = "mapview") %>%
  raster::stack(.)
mapview(kili_data[[1:3]])
```

---

### tmap
```{r, eval = FALSE}
library(tmap)
data(World)
# tmaptools::palette_explorer()
tm_shape(raintot) + tm_raster(palette = "magma", breaks = seq(0, 200, 25)) +
  tm_shape(districts) + tm_borders() + 
  tm_layout(bg.color = "grey", inner.margins = 0.1) # +
  # tm_shape(World) + tm_borders()
```
---

### leaflet
```{r, eval = FALSE}
library(leaflet)
m <- leaflet() %>% addTiles()
m  # a map with the default OSM tile layer


# set bounds
m %>% fitBounds(0, 40, 10, 50)

# move the center to Snedecor Hall
m <- m %>% setView(-93.65, 42.0285, zoom = 17)
m

# popup
m %>% addPopups(-93.65, 42.0285, 
                "Here is the <b>Department of Statistics</b>, ISU")
rand_lng <- function(n = 10) rnorm(n, -93.65, .01)
rand_lat <- function(n = 10) rnorm(n, 42.0285, .01)

# use automatic bounds derived from lng/lat data
m <- m %>% clearBounds()

# popup
m %>% addPopups(rand_lng(), rand_lat(), "Random popups")
```
---


### RStoolbox
```{r, eval = FALSE}
library(RStoolbox)
data(lsat)
ggRGB(lsat, stretch = "lin")
ggRGB(lsat[[2:4]], stretch = "lin")
ggR(lsat[[1]])
ggR(lsat[[1]], stretch = "hist")
ggR(lsat[[1]], stretch = "lin")
ggR(raintot)
```

---

### RStoolbox::spectralIndices
```{r, eval = FALSE}
ndvi <- spectralIndices(lsat, red = "B3_dn", nir = "B4_dn", indices = "NDVI")
ggR(ndvi, geom_raster = TRUE) +
  scale_fill_gradientn(colours = c("black", "white"))
  # scale_fill_viridis_c()

mtlFile  <- system.file("external/landsat/LT52240631988227CUB02_MTL.txt", 
                        package="RStoolbox")
lsat_ref <- radCor(lsat, mtlFile, method = "apref")

SI <- spectralIndices(lsat_ref, red = "B3_tre", nir = "B4_tre")
plot(SI)
# rasterVis::levelplot(SI)
```

---
### Plotly

Many examples in code demos section of `plotly` help

```{r, eval=FALSE}
library(plotly)
library(sf)

nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
p <- ggplot(nc) + geom_sf(aes(fill = AREA))
ggplotly(p)
ggplot() + geom_sf(data = nc, aes(fill = AREA))

# If not supplied, coord_sf() will take the CRS from the first layer
# and automatically transform all other layers to use that CRS. This
# ensures that all data will correctly line up
nc_3857 <- sf::st_transform(nc, "+init=epsg:3857")
p2 <- ggplot() +
  geom_sf(data = nc) +
  geom_sf(data = nc_3857, colour = "red", fill = NA)
ggplotly(p2)
```

---

### Plotly cont.
```{r, eval=FALSE}
# Unfortunately if you plot other types of feature you'll need to use
# show.legend to tell ggplot2 what type of legend to use
nc_3857$mid <- sf::st_centroid(nc_3857$geometry)
p3 <- ggplot(nc_3857) +
  geom_sf(colour = "white") +
  geom_sf(aes(geometry = mid, size = AREA), show.legend = "point")
ggplotly(p3)

# You can also use layers with x and y aesthetics: these are
# assumed to already be in the common CRS.
p4 <- ggplot(nc) +
  geom_sf() +
  annotate("point", x = -80, y = 35, colour = "red", size = 4)
ggplotly(p4)
```

---
## Modelling demo
### Random Forests
```{r, eval=FALSE}
# Train model
require(caTools)
library(randomForest)
library(dplyr)

# read in train set
train_ref <- readr::read_csv(
  system.file("extdata/train_reference.csv", package = "ecaascrops")
) %>% mutate(crop = factor(case_when(
  class == "Maize" ~ 1,
  class == "Rice" ~ 2,
  class == "Other" ~ 3,
  class == "noncrop" ~ 4
))) %>% select(id, class, crop, !!names(.))
```

---
### Model 

```{r, eval=FALSE}
# prepare dataset
dset <- train_ref %>% filter(type == "mean") %>%
  select(-id, -class, -type, -seed)
set.seed(10)
samp <- sample.split(dset$crop, SplitRatio = 0.8)
train <- subset(dset, samp == TRUE)
test <- subset(dset, samp == FALSE)
# train %>% tidyr::drop_na()

# fit model
set.seed(123)
mod <- randomForest(crop ~ ., data = train, ntree = 1000,
                    importance = TRUE)

# Evaluation
pred <- predict(mod, newdata = test %>% dplyr::select(-c(crop)))
cm <- table(test$crop, pred)

rf_results <- list("train" = train, "test" = test, "mod" = mod, "pred" = pred,
                   "error_mat" = cm)
```

---
```{r, eval=FALSE}
imptab <- mod$importance
nms <- rownames(imptab)
imp_tbl <- as_tibble(imptab) %>% 
  mutate(Variable = nms) %>% 
  dplyr::select(Variable, MeanDecreaseAccuracy) %>% 
  arrange(MeanDecreaseAccuracy) %>% 
  mutate(order = 1:n()) %>% 
  mutate(Variable = gsub("\\.", "_", Variable))

p <- ggplot(imp_tbl) + 
  geom_point(aes(MeanDecreaseAccuracy, factor(order))) + 
  scale_y_discrete(labels = toupper(imp_tbl$Variable)) + 
  ylab("Variable") + xlab("Mean Decrease in Accuracy") + 
  theme_linedraw() + 
  theme(axis.text = element_text(size = 7), 
        axis.title = element_text(size = 7))
```














