---
title: "Geospatial Analysis with R"
subtitle: Class 11
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Read and write vector data.
- Create geometries from scratch: points, linestrings, and polygons.
- Convert geometry types.
- Set and transform CRS.

---

## Read vector data

```{r, warning=F, message=F}
library(sf)
library(here)
districts_path <- system.file('extdata/districts.shp', package = 'geospaar')
districts <- st_read(dsn = districts_path)
```

Or `read_sf`, the same as `st_read` just with different default values

```{r}
districts <- read_sf(dsn = districts_path)
```

`dsn` could be a file path, a database connection, or even a geojson text.

```{r}
geojson_txt <- paste("{\"type\":\"MultiPoint\",\"coordinates\":",
                     "[[3.2,4],[3,4.6],[3.8,4.4],[3.5,3.8],",
                     "[3.4,3.6],[3.9,4.5]]}")
sf_emp <- read_sf(dsn = geojson_txt)
```

---

## Write vector data

```{r, eval=F}
st_write(districts, dsn = here('inst/extdata/districts.shp'),
         driver = "ESRI Shapefile")
```

Or `write_sf`.  Similarly, `dsn` could be a path or database connection.

---

## CRS

- EPSG code, e.g. 4326 for WGS84
- WKT text, e.g. result of `st_crs(districts)`

[List of ESPG codes on spatialreference.org](https://spatialreference.org/ref/epsg/)

---

## Creation - point

```{r, eval=F}
grace_sfg <- st_point(c(-71.82291048471912, 42.250181963699134))
grace_sfg; class(grace_sfg) # sfg, simple feature geometry
grace_sfc <- st_sfc(grace_sfg)
grace_sfc; class(grace_sfc) # sfc, simple feature column, geospatial already
st_crs(grace_sfc) <- 4326 # set CRS for geopsatial, EPSG code
grace_sfc
grace_pt <- st_sf(grace_sfc) # sf, simple feature collection
grace_pt; class(grace_pt)
attr(grace_pt, "sf_column") # check sfc
grace_geom <- st_geometry(grace_pt) # get sfc

identical(grace_sfc, grace_geom) # st_geometry gets sfc from a sf data.frame
```

It is your turn to put them together within one pipeline.

---

## Creation - multipoint

`multipoint` is only one geometry for multiple points.

```{r, eval=F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396))
clark_campus_mpt <- st_multipoint(coords) %>% # create sfg
  st_sfc() %>% st_set_crs(4326) %>% # create sfc
  st_sf() # build up sf
clark_campus_mpt
plot(clark_campus_mpt %>% st_geometry()) # plot
```

---

## Creation - points

```{r, eval=F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396))
clark_campus_pts <- do.call(st_sfc, 
                        lapply(1:nrow(coords), 
                               function(n){
  st_point(coords[n, ]) # create sfg for each point
})) %>% st_set_crs(4326) %>% # create sfc
  st_sf() # build up sf
clark_campus_pts
plot(clark_campus_pts %>% st_geometry()) # plot
```

---

## `st_as_sf` function

```{r, eval=F}
dt_url <- 'https://agroimpacts.github.io/geospaar/hyenas_occurrence.csv'
occurrence <- read.csv(dt_url, stringsAsFactors = F) %>% 
  dplyr::select(gbifID, decimalLatitude, decimalLongitude) %>% 
  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'),
           crs = 4326)
```

---

## Creation - linestring

```{r, eval=F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396))
clark_campus_l <- st_linestring(coords) %>% # create sfg
  st_sfc() %>% st_set_crs(4326) %>% # create sfc
  st_sf() # build sf
clark_campus_l # Just one geometry
plot(clark_campus_l %>% st_geometry())
```

---

## Creation - linestrings

```{r, eval=F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396))
clark_campus_ls <- do.call(st_sfc, 
                        lapply(1:(nrow(coords) - 1), 
                               function(n){
  # create sfg for each straight line
  st_linestring(coords[n:(n + 1), ])
})) %>% st_set_crs(4326) %>% # create sfc
  st_sf() # build up sf
clark_campus_ls
plot(clark_campus_ls %>% st_geometry()) # plot
```

---

## Creation - multilinestring

`multilinestring` is only one geometry for multiple linestrings.

```{r, eval=F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396))
clark_campus_ml <- st_multilinestring(list(coords)) %>% 
  st_sfc() %>% st_set_crs(4326) %>% # create sfc
  st_sf() # build up sf
clark_campus_ml
plot(clark_campus_ml %>% st_geometry()) # plot
```

---

## Creation - polygon

```{r, eval = F}
coords <- rbind(c(-71.82117553759312, 42.25084734394512),
                c(-71.82212771135572, 42.25194565868353),
                c(-71.82156280512835, 42.252417357356265),
                c(-71.82316986959349, 42.25367904720517),
                c(-71.82545037620447, 42.2518579577067),
                c(-71.82587265873052, 42.25131095436722),
                c(-71.8236126925188, 42.249563032541396),
                c(-71.82117553759312, 42.25084734394512)) # repeat the first point
clark_campus_ply <- st_polygon(list(coords)) %>% # create sfc
  st_sfc() %>% st_set_crs(4326) %>%  # create sfc
  st_sf()
clark_campus_ply
plot(clark_campus_ply$geometry)
```

---

## `st_as_sfc` function

```{r, eval=F}
bbox_ply <- st_bbox(clark_campus_ply) %>% 
  st_as_sfc() %>% st_sf()
```

---

## Convert types: `st_cast` function

Geometry types: `POINT, MULTIPOINT, LINESTRING, MULTILISTRING, POLYGON, MULTIPOLYGON`. 

Literally we could convert from any type to any type. But be careful of the order. E.g. `POINT` -> `MULTIPOINT` (good) and the opposite (bad).

```{r, eval = F}
st_cast(clark_campus_ply, 'POINT')
```

---

## Set and convert CRS

```{r, eval=F}
st_crs(clark_campus_ply) # check
```

If NA, it means there is not CRS set, so we need to set one.

```{r, eval=F}
st_set_crs(clark_campus_ply, 4326) # set 
st_crs(clark_campus_ply) <- 4326
st_crs(clark_campus_ply) <- st_crs(clark_campus_ls) # or a crs object
```

If a geospatial data already have CRS, you might want to convert to other system. It is WRONG to just set another CRS. The right way is to use `st_transform` to convert.

```{r, eval=F}
st_transform(clark_campus_ply, 2202)
```

---

## Next class

- Geospatial statistics
- Buffer
- Union
- Logical relationships
- Others

---

## Homework

Each student creates a geospatial polygon for a building at Clark.

```{r, warning=F, message=F, echo=F}
library(kableExtra)
set.seed(10)
assignments <- data.frame(student = c('Ashna', 'Danielle', 'Lia', 'Lara', 
                                       'Morgan', 'Rachel', 'Ruhua', 'Sarah', 
                                       'Sophie', 'Tom'),
                          building = c('Atwood Hall + Geography', 'Jonas Clark Hall',
                                       'Higgins + Sackler Science Center', 
                                       'Math and Physics', 'Carlson Hall', 
                                       'Robert Hutchings Goddard Library', 
                                       'Kneller Athletic Center',
                                       'Dana Commons', 'Little Center',
                                       'Estabrook Hall') %>% sample(10))
assignments %>% 
  kbl() %>% 
  kable_paper(full_width = F) %>% 
  row_spec(1:10, color = "black", background = "white")
```

