---
title: "Geospatial Analysis with R"
subtitle: Class 18
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Challenge from last class
- Raster intro
  - classes
  - dummy rasters
  - reading and writing
  - rasterize, raster to vector


---
```{r, echo=FALSE, out.width="50%", results='hold'}
dem1 <- "figures/zam_dem_raster.png"
dem2 <- "figures/zam_dem_stars.png"
knitr::include_graphics(c(dem1, dem2))
```


---
### A quick demo of multi-panel plots
```{r, eval = FALSE}
p <- ggplot(districts) + geom_sf() + 
  geom_sf(data = smallest_centroid, color = "red")
p
p1 <- p + geom_sf(data = small_pol, aes(color = as.factor(ID)),
                  fill = "transparent") + labs(color = "ID")
small_pol2 <- pol %>% st_sfc() %>% st_sf(crs = 4326) %>% mutate(ID = 1)
p2 <- p + geom_sf(data = small_pol2, aes(color = as.factor(ID)),  
                  fill = "transparent") + labs(color = "ID")

gridExtra::grid.arrange(p1, p2, ncol = 2)

g1 <- cowplot::plot_grid(p1 + theme(legend.position = "none"), 
                         p2 + theme(legend.position = "none"))
cowplot::plot_grid(g1, cowplot::get_legend(p1), rel_widths = c(2, 0.3))
```

---
### `patchwork`
```{r, eval=FALSE}
library(patchwork)

p1 + p2 + plot_layout(guides = "collect")

p1 + p1 / p
(p1 / p) + (p2 / p)
(p1 + p2) / (p + p)

```


---
## Reading/Writing/Manipulating rasters

- create
```{r, eval = FALSE}
r <- raster(extent(30, 31, -14, -13), res = 0.1, crs = 4326)
values(r) <- sample(1:10, size = ncell(r), replace = TRUE)

par(mar = c(0, 0, 0, 0))
plot(districts %>% st_geometry)
plot(r, add = TRUE)

plot(r)

s <- stack(r, log10(r))
names(s) <- c("dummy", "log10dummy")
plot(s)

b <- brick(s)
brick(list(r, log10(r)))
plot(b)

b2 <- lapply(1:10, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.1, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  r
}) %>% stack %>% brick
plot(b2)
```

---
### A large brick
```{r, eval = FALSE}
b3 <- lapply(1:10, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.001, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  r
}) %>% stack %>% brick
plot(b3)
plot(b3[[1]])
plot(b3[[10]])
plot(b3[[c(1, 10)]])

```

---

- read and write

```{r, eval = FALSE}
writeRaster(r, filename = file.path(tempdir(), "mydummy.tif"), overwrite = TRUE)
dir(tempdir())
# rm(r)
r <- raster(file.path(tempdir(), "mydummy.tif"))
plot(r)

brick(s, filename = file.path(tempdir(), "mydumbrick.tif"), overwrite = TRUE)
# stack(s, filename = file.path(tempdir(), "mydumbstack.tif"), overwrite = TRUE)
brick(s, filename = file.path(tempdir(), "mydumbrick.grd"), overwrite = TRUE)
brick(file.path(tempdir(), "mydumbrick.tif"))
brick(file.path(tempdir(), "mydumbrick.grd"))
```

---

- plot
```{r, eval = FALSE}
plot(b)
plot(b2)
plot_noaxes(b)
plot_noaxes(b2, main = paste("Random", 1:10))
plot(b2, main = paste("Random", 1:10))
plot_noaxes(b, legend = FALSE)

plot_noaxes(b[[1]], legend = FALSE)
legend("right", legend = 1:10, fill = terrain.colors(10), border = FALSE, 
       bty = "n")

# plotting with larger datasets
plot_noaxes(b3, main = paste("Random", 1:10))
plot_noaxes(b3, nc = 3, nr = 4, main = paste("Random", 1:10))

rasterVis::levelplot(b3, names.attr = paste("Random", 1:10))

rasterVis::levelplot(b2, names.attr = paste("Random", 1:10))
rasterVis::levelplot(b[[1]])
rasterVis::gplot(b[[1]]) + geom_tile(aes(fill = value)) + 
  scale_fill_viridis_c()
ggplot2::ggplot(as.data.frame(b[[1]], xy = TRUE)) + 
  # geom_tile(aes(x = x, y = y, fill = dummy)) + scale_fill_viridis_c()  
  geom_raster(aes(x = x, y = y, fill = dummy)) + scale_fill_viridis_c()
stars::st_as_stars(b) %>% plot(col = viridis::viridis(10))
stars::st_as_stars(b) %>% plot()
```

---
## Raster to other types
```{r, eval=FALSE}
plot(b)
b[1:10]
b[1:ncell(b)]
as.matrix(b)
as.data.frame(b) %>% as_tibble()
```


---
## Pre-processing
- Aggregating/disaggregating
```{r, eval = FALSE}
aggregate(b, fact = 2) %>% plot
# aggregate(b, fun = min, fact = 2) %>% plot
# aggregate(b, fun = sd, fact = 2) %>% plot
plot(b)
disaggregate(b, fact = 2) %>% plot
disaggregate(b, fact = 2, bilinear = TRUE) %>% plot
```

- Masking
```{r, eval = FALSE}
library(geospaar)
data(chirps)
plot_noaxes(chirps[[1]])
chirpsz <- mask(chirps, districts)
plot_noaxes(chirpsz[[1]])
plot(st_geometry(districts), add = TRUE)
# rasterVis::levelplot(chirpsz[[1:5]])
plot_noaxes(chirpsz, nr = 3, nc = 4, maxnl = 12, mar = c(0, 0, 0, 0), 
            legend = FALSE, zlim = c(0, 50))
plot(chirpsz, nr = 3, nc = 4, maxnl = 12, mar = c(0, 0, 0, 0), 
     legend = FALSE, zlim = c(0, 50))
```

---

- Projecting
```{r, eval = FALSE}
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
chirpsz %>% projectRaster(., crs = crs(roads), res = 5000) -> chirpsz_alb
chirpsz_alb %>% print
```

- Rasterize/vectorize
```{r, eval = FALSE}
distsr <- districts %>% rasterize(., chirpsz)
distsr %>% plot_noaxes
# stack(distsr, chirpsz[[1:2]]) %>% plotRGB(stretch = "lin")
distsr_pol <- rasterToPolygons(distsr, dissolve = TRUE)
distsr_pol %>% st_as_sf %>% st_geometry %>% plot
distsr_pol %>% st_as_sf %>% slice(49) %>% plot(add = TRUE)
```

---
## Calculations

- raster algebra
- statistics
- z dimension stats

---
## Exercises

- Create a dummy raster (sampling from 1:100) using district 49 for extent and res of 0.1
- Reproject to Albers
- Disaggregate to 0.02 degrees (bilinear and NGB)
- Calculate the sum of rainfall from `chirps`
- Identify all areas of total rainfall > 10 mm
- Calculate the mean rainfall in Zambia for the 15th day in `chirps`



