---
title: "Geospatial Analysis with R"
subtitle: Class 8
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class8.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r, eval=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
data(world.cities, package = "maps")

world <- ne_countries(scale = "medium", returnclass = "sf")
afr_capitals <- world.cities %>% filter(capital == 1) %>% 
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>% 
  st_intersection(., world %>% filter(continent == "Africa"))
p <- world %>% filter(continent == "Africa") %>% 
  ggplot() + geom_sf(aes(fill = name), lwd = 0.2) + 
  geom_sf(data = afr_capitals, col = "blue", size = 0.5) + 
  scale_fill_grey(guide = FALSE) + theme_minimal()
ggsave(here::here("external/slides/figures/africa_capitals.png"), 
       width = 5, height = 4, dpi = 300, bg = "transparent")
```

---
# Programming tip

- Whenever possible, let the computer do your counting for you:
```{r}
a <- 1:1001
a[length(a)]
```
- Rather than 
```{r}
a[1001]
```
- Imagine this case:
```{r}
set.seed(1)
b <- 1:sample(1:10000, 1)
b[length(b)]
```
---


# Today

- Next module
- 2-d indexing
- A little bit of summarizing
- Control structures (maybe)

---

## Next module
- Reading and writing non-spatial data
- Data preparation/shaping
- Basic analyses
- Visualization (non-spatial)

---
## Create your own data

- Create the following 1-D structures
  - `a`: a random vector of integers with 10 elements drawn from 1-20:
    - Use the `sample` function with `set.seed(10)`
    - Name the elements of `a` with a vector of names starting with "V1" and ending with "V10". 
      - Use the `paste0` function to create those names.
      - Create the identical vector of names using the `paste` function.
  - `b`: Using `a` as an index to select from `letters`
  - `d`: Use `rnorm` with a mean = 100 and an sd of 20
    - Why did I skip `c`? 
  - Create a list `l` from `a`, `b`, `d`. 
    - Assign the names of the vectors in `l` to the `l`'s elements

```{r, echo=FALSE}
set.seed(10)
a <- sample(1:20, 10, replace = TRUE)
names(a) <- paste0("V1", 1:10)
names(a) <- paste("V1", 1:10, sep = "")
b <- letters[a]
d <- rnorm(n = 10, mean = 100, sd = 20)
l <- list("a" = a, "b" = b, "d" = d)

```
---
## Create your own data

- Create the following 2-d structures
  - `m`: a matrix with three integer columns named "V1", "V2", "V3"
    - Create each column first as its own vector, then combine
    - `V1` = 1:10
    - `V2` is a random sample between 1:100
    - `V3` is drawn from a random uniform distribution between 0 and 50     
    - Use `set.seed(10)`
    - Inspect the `str` and `class` of `m`
  - `dat`, a data.frame built from `V1`, `V2`, `V3`, and `V4`
    - `V4` is a random selection of the letters A-E
    
```{r, echo=FALSE}
set.seed(10)
m <- cbind(V1 = 1:10, 
           V2 = sample(1:100, size = 10, replace = TRUE), 
           V3 = runif(n = 10, min = 0, max = 50))
str(m)
dat <- data.frame(m, V4 = sample(letters[1:5], size = 10, replace = TRUE))
dat
```
---

## 2-d Indexing/subsetting/replacing
- Select the first 10 values from `m`, using a single vector and no row or column information
- Use a single vector to select the last row, column value from `m`
- Replace the value selected in 2 above with -99

```{r, eval=FALSE}
m[length(m), 1] <- -99
```

- Now select row 3, columns 1:2 from `m`, and replace them with their values multiplied by 10 

```{r, eval=FALSE}
m[3: 1:2] <- m[3, 12] * 10
```

- Do the same, but select the columns by their name, and reset the new values by dividing by 10

```{r, eval=FALSE}
m[3, c("V1", "V2")] <-  # finish this
```

---

- Extract V3 from `dat` and square it. Using the column name and both `[]` (there are three ways) and `$` to do the subsetting.

```{r, eval=FALSE}
# The 1 provided, two missing
dat[["V3"]]^2
```

- Subset the first two rows and columns of `dat` into a new data.frame `datss`.
- Replace `dat` rows 1:2, column 1:2 with the values -1:-4
- Reset the part of `dat` you just changed with the values in `datss`

```{r, eval=FALSE}
datss <- dat[1:2]  # 1
dat[1:2][1:2] <- c(-1, -4)  # 2
# ??? # 3
```


```{r, echo=FALSE, eval = FALSE}
m[1:10, ]
m[length(m)]
m[length(m)] <- -99
m[3, 1:2] <- m[3, 1:2] * 10
m[3, c("V1", "V2")] <- m[3, c("V1", "V2")] / 10
dat$V3^2
dat[, 3]^2
dat[["V3"]]^2
dat["V3"]^2
dat[, "V3"]^2
datss <- dat[1:2, 1:2]
dat[1:2, 1:2] <- -1:-4
dat[1:2, 1:2] <- datss
```

---
## Control structures
### Branching
```{r}
a <- 5
if(a > 10) {
  print("Greater than 10!")
} else {
  print("Less than or equal to 10")
}
```

### Looping
```{r}
b <- 1:3
for(i in b) print(i)
```

---

### *apply
- A special form of looping
- Intended for *applying* a function to data

```{r}
l <- list(dat$V1, dat$V2, dat$V3)
l
lapply(l, mean)
```

---
### *apply

- Key uses: 
  - Return results of loop directly into object
  - Use with anonymous functions to pass an iterator, often into more complex procedures
  
```{r}
# Simple
o <- lapply(1:2, function(x) l[[x]])
o
```

---
```{r}
# More complex
o2 <- lapply(1:5, function(x) {
  l[[1]][x] - l[[2]][x]
})
o2
```

---
## Looping practice

- Write a `for` loop that iterates through the vector 1:10 and prints the iterator `i` multiplied by 10

```{r, eval=FALSE}
for(i in 10) i
```

- Do the same, but instead of print `i * 10`, catch the result in a predefined empty list `o`
```{r, eval=FALSE}
for(i in c(1, 10)) {
  o <- i * 10
}
```

- Do the same as above, but use an `lapply` that assigns output to `o`
- Do the same as above, but use `sapply` instead of `lapply`
- Let's use `sapply` to find which elements of `l` are `matrix`
- Let's use `lapply` to calculate the `colMeans` of matrices and `data.frame`s in `l`
