---
title: "Geospatial Analysis with R"
subtitle: Class 13
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
## Election Results
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width=8, fig.height=5}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")

cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>%
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% 
  dplyr::select(-other, -delres) %>%
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab(NULL) + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

---
```{r, eval = FALSE}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")
cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>%
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% select(-other, -delres) %>%
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab("") + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2))
```
---
## Today

- Project
- Spatial data
  - Non-spatial to spatial
  - Read and write spatial
  - CRS
  - Make data from scratch
  - Spatial properties

---
## Projects
- On [projects page](https://agroimpacts.github.io/geospaar/projects.html)
- USF-projects:
  - Mapping the sources/destinations/flows of budget revenues for police departments, from federal/state/municipal sources. 
  - Changes in property valuation versus police activity and/or budgets (linked to #1)
  - Spatio-temporal patterns and associations of police misconduct
  - Spatio-temporal patterns of surveillance (e.g. CCTV)
  - Mapping #blacklivesmatter social media

---

### Datasets for today
```{r, eval=FALSE}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sfc(coords = c("X", "y"))
districts <- st_reader(
  system.file("extdata/districts.shp", package = "geospaar")
)
roads <- read_sfc(system.file("extdata/roads.shp", package = "geospaar"))
```

---
### Non-spatial to spatial

- Read in farmer data
```{r, eval = FALSE}
library(sf)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() 
farmers_sf <- st_as_sf(farmers, coords = c("x", "y"), crs = 4326)
```

- Write and read the spatial data
```{r, eval=FALSE}
write_sf(farmers_sf, dsn = "~/Desktop/demo/farmers_spatial.geojson")
farmers_sf <- read_sf("~/Desktop/demo/farmers_spatial.geojson")
```

---
  
### Wrangle spatial data
```{r, eval = FALSE}
farmers %>% filter(uuid == "009a8424")
# farmers %>% distinct(uuid)
farmers_sf %>% filter(uuid == "05a95f44")
farmers_sf %>% filter(uuid == "064beba0")
farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% 
  arrange(desc(n)) %>% 
  st_geometry %>% 
  plot(pch = 16, cex = 0.5)

farmers_sf %>% arrange(uuid)
farmers_sf %>% arrange(uuid, desc(date))

class(farmers_sf)
str(farmers_sf)

# intersection with st_join and left = FALSE to remove plots outside of Zambia
farmers_sf_trim <- farmers_sf %>% 
  group_by(uuid) %>% 
  summarize(n = n()) %>% 
  st_join(., districts, left = FALSE)

par(mar = rep(0, 4))
plot(st_geometry(districts))
farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% st_geometry %>% 
  plot(pch = 16, add = TRUE, cex = 1)
farmers_sf_trim %>% st_geometry %>% 
  plot(pch = 16, add = TRUE, col = "red", cex = 0.5)
```

---
### Make spatial data

Points and lines
```{r, eval = FALSE}
pts <- st_multipoint(x = cbind(x = c(27.5, 28, 28.5), y = c(-14.5, -15, -15.5)))
pts <- st_geometry(pts) %>% st_as_sf() %>% st_cast("POINT") %>% 
  cbind(ID = 1:nrow(.), .) %>% rename(geometry = x)

# polyline
sline <- st_linestring(cbind(x = c(27, 27.5, 28), y = c(-15, -15.5, -16)))
sline <- st_geometry(sline) %>% st_as_sf() %>% cbind(ID = 1:nrow(.), .)
plot(sline)
```
---

Polygons
```{r, eval = FALSE}
# Single polygon
pol <- st_polygon(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5), 
                             y = c(-15.5, -16.5, -17, -16, -15.5))))
pol <- st_geometry(pol) %>% st_as_sf() %>% cbind(ID = 1:nrow(.), .)
plot(pol)

# multipolygon
pols <- st_multipolygon(list(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5),
                                        y = c(-15.5, -16.5, -17, -16, -15.5))),
                             list(cbind(x = c(26.5, 27.5, 27, 26, 26.5) + 1,
                                        y = c(-15.5, -16.5, -17, -16, -15.5) -
                                          1))))
pols <- pols %>% st_geometry %>% st_cast("POLYGON") %>% st_as_sf() %>% 
  mutate(ID = 1:nrow(.)) %>% #as.data.frame() %>%
  dplyr::select(ID)
# plot(pols)

```
---
### Spatial properties
```{r, eval=FALSE}
pol %>% st_area
pol %>% st_sf(crs = 4326) %>% st_area
sline %>% st_sf(crs = 4326) %>% st_length
```

---
### Spatial transforms
```{r, eval=FALSE}
pol %>% 
  st_sf(crs = 4326) %>% 
  st_transform(., st_crs(roads)) %>% 
  st_area
```

---

### Tidy analysis `sf` data

- Let's make some data

```{r, eval=FALSE}
set.seed(1)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  dplyr::select(distName, X, Y, yield)
ggplot(districts2) + geom_sf(aes(fill = yield))
districts2 <- districts2 %>% 
  mutate(
    grp = case_when(
      grepl("Cha", distName) ~ "A",
      grepl("Chi", distName) ~ "B", 
      grepl("^L", distName) ~ "C", 
      TRUE ~ "D"
    )
  ) %>% dplyr::select(distName, grp, X, Y, yield)
```

---
- And analyze it

```{r, eval=FALSE}
districts2 %>% 
  dplyr::select(yield) %>% 
  summarize_all(list(mean, sd))
districts2 %>% 
  summarize(mean(yield), sd(yield))
```

---
```{r, eval=FALSE}
districts2 %>% dplyr::select(yield) %>% summarize_all(funs(mean, sd))
districts2 %>% summarize(mean(yield), sd(yield))
districts_mu <- districts2 %>% 
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(1:5, 7) %>% 
  group_by(yld_cat) %>% summarize(yield = mean(yield))
ggplot(districts_mu) + geom_sf(aes(fill = yield))
  
```

---
### Subsetting

By count
```{r, eval = FALSE}
as.data.frame(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n)) %>% 
  ggplot() + geom_sf(aes(fill = log10(n)))
```

By area
```{r, eval = FALSE}
p <- ggplot(districts2) + geom_sf()
districts2 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter(area == quantile(area, probs = 0.5, type = 1))
districts2 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  arrange(area) %>% slice(nrow(.))
districts2 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  arrange(area) %>% pull(distName)
```

---
### Spatial joins
```{r, eval=FALSE}
farmer_ct <- farmers_sf %>% group_by(uuid) %>% count()
farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
```

---
### Spatial-non-spatial join
```{r, eval=FALSE}
as.data.frame(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n)) %>% 
  ggplot() + geom_sf(aes(fill = log10(n)))
```

---
### Intersects/Intersection

```{r, eval=FALSE}
districts2 %>% 
  slice(which(st_intersects(districts2, st_union(pols), sparse = FALSE))
        
districts_pol <- st_intersection(districts2, st_union(pols))
ggplot(districts_pol) + geom_sf(aes(fill = distName))
```



