---
title: "Geospatial Analysis with R"
subtitle: Class 14
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

### Today

- More on vectors and analytics
- Maybe some visualization

---
### Assignment 3 recap
```{r, eval=FALSE}
set.seed(1)
t1 <- tibble(V1 = paste0("G", 1:10), V2 = runif(n = 10, 75, 125))
t2 <- tibble(v1 = paste0("G", 1:15), 
             v3 = sample(LETTERS[1:6], size = 15, replace = TRUE))
t3 <- tibble(v1 = paste0("G", 1:20), 
             v4 = rnorm(n = 20, mean = 100, sd = 20))
t1 %>% left_join(., t2, by = c("V1" = "v1")) %>% 
  left_join(., t3, by = c("V1" = "v1"))
t1 %>% right_join(., t2, by = c("V1" = "v1")) %>% 
  right_join(., t3, by = c("V1" = "v1"))
t1 %>% full_join(., t2, by = c("V1" = "v1")) %>% 
  full_join(., t3, by = c("V1" = "v1"))
t1 %>% inner_join(., t2, by = c("V1" = "v1")) %>% 
  inner_join(., t3, by = c("V1" = "v1"))
```


---
### Read in `geospaar` data
Use code from previous classes to read in `farmers` (make into `sf`), `districts`, and `roads`
```{r, eval=FALSE, echo = FALSE}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sf(coords = c("x", "y"))
districts <- st_read(
  system.file("extdata/districts.shp", package = "geospaar")
)
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
```

---
### Adding to data

Add some categories
```{r, eval=FALSE}
set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
```
---

## Analyses
```{r, eval=FALSE}
# districts2 %>% dplyr::select(yield) %>% summarize_all(funs(mean, sd))
# districts2 %>% summarize(mean(yield), sd(yield))
districts2 %>% group_by(yld_cat) %>% 
  summarize(yield = mean(yield)) %>% 
  ggplot() + geom_sf(aes(fill = yield))
  # geom_sf(aes(fill = yld_cat))
  
```

---
### Spatial joins
```{r, eval=FALSE}
p <- ggplot(districts2) + geom_sf(fill = "grey")
# farmer_ct <- farmers %>% group_by(uuid) %>% count()
# st_crs(farmer_ct) <- st_crs(districts)
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))
farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
# p + geom_sf(data = farmers_districts)

farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
p + geom_sf(data = farmers_per_dist) + 
  geom_sf(data = farmers_per_dist %>% st_centroid(), col = "blue")
```

---
### Spatial-non-spatial join
```{r, eval=FALSE}
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))

p + geom_sf(data = districts3, aes(fill = log10(n)))
# p + geom_sf(data = districts3, aes(fill = n))
```

---
### Subsetting

By count
```{r, eval = FALSE}
p + districts3 %>% filter(n > 15) %>% geom_sf(data = ., fill = "red") + 
  districts3 %>% filter(between(n, 1, 15)) %>% geom_sf(data = ., fill = "blue")
p + geom_sf(data = districts3 %>% filter(n > 15), fill = "red") + 
  geom_sf(data = districts3 %>% filter(between(n, 1, 15)), fill = "blue")
```

By area
```{r, eval = FALSE}
dmed <- districts3 %>% #mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  mutate(area = as.numeric(units::set_units(st_area(.), "km2"))) %>% 
  filter(area == quantile(area, probs = 0.5, type = 1)) 
p2 <- p + geom_sf(data = dmed, fill = "red")

dmax <- districts3 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  # arrange(area) %>% slice(nrow(.)) %>% 
  arrange(desc(area)) %>% slice(1) 
  # slice_max(order_by = area) #
p2 + geom_sf(data = dmax, fill = "blue")

districts3 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  arrange(area) %>% pull(distName)
```

---
## Various

Dataset
```{r, eval=FALSE}
pols <- st_multipolygon(list(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5),
                                        y = c(-15.5, -16.5, -17, -16, -15.5))),
                             list(cbind(x = c(26.5, 27.5, 27, 26, 26.5) + 1,
                                        y = c(-15.5, -16.5, -17, -16, -15.5) -
                                          1))))
pols <- pols %>% st_geometry %>% st_cast("POLYGON") %>% st_as_sf(crs = 4326) %>%   
  mutate(ID = 1:nrow(.)) %>% #as.data.frame() %>%
  dplyr::select(ID)

p + geom_sf(data = pols, aes(fill = as.factor(ID))) + 
  geom_sf(data = st_union(pols), 
          fill = rgb(1, 0, 1, alpha = 0.7), col = "purple")
```

---
### Intersects
```{r, eval=FALSE}
int_dists <- districts2 %>% 
  slice(which(st_intersects(districts2, st_union(pols), sparse = FALSE)))
p + geom_sf(data = int_dists, fill = "red") + 
  geom_sf(data = pols, fill = "transparent", col = "blue")
```

---
### Intersection
```{r, eval=FALSE}
districts_pol <- st_intersection(districts2, st_union(pols))
ggplot(districts_pol) + geom_sf(aes(fill = distName))
p + geom_sf(data = districts_pol, aes(fill = distName))
```

---
### On your own

- Select the smallest district in Zambia from `districts3`, and convert that
to centroid
```{r, eval = FALSE}
smallest_centroid <- districts3 %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter
```

- Create a buffer of width 0.5 around that
```{r, eval = FALSE}
smallest_buffer <- smallest_centroid %>% st_buf...
```

- Use the buffer to select the districts from `districts3` that intersect with it
```{r, eval = FALSE}
dists_in_buff <- districts3 %>% 
  filter(st_intersects(., smallest_buffer...
```


---
- Use the buffer to do an intersection with `districts3`
```{r, eval=FALSE}
buff_int_dists <- st_intersection(...
# p + geom_sf(data = buff_int_dists, aes(fill = distName))
```


- Calculate the area of the intersection

---


## Visualization

- `graphics` versus `ggplot`

```{r, eval = FALSE}
library(tidyverse)
fs <- dir(system.file("extdata/", package = "geospaar"), pattern = "FAOSTAT", 
          full.names = TRUE)
crops <- lapply(fs, readr::read_csv)
crops_df <- do.call(rbind, lapply(crops, function(x) {
  x %>% dplyr::select(Item, Area, Element, Year, Value) %>% 
    spread(key = Element, value = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = `Area harvested`, prod = Production) %>% 
    mutate(harv_km2 = harv_area / 100)
}))
crop_ylds <- crops_df %>% mutate(yields = prod / harv_area) %>%     
  mutate(country = ifelse(country == "South Africa", "ZAF", country)) %>%  
  mutate(country = ifelse(country == "Zambia", "ZMB", country)) %>% 
  mutate(harv_km2 = harv_area / 100)
```

---
### `graphics`

- Need this because we'll use a lot with `sf` and `raster`
- Key functions: 
  - `plot`, `barplot`, `hist`, `boxplot`, `par`
- Other helpers:
  - `axis`, `text`, `mtext`, `legend`

---
### `graphics` examples
- Scatter
- Histogram
- multi-panel

---
#### Scatter
```{r, eval = FALSE}
# single line
cols <- c("red", "orange", "blue", "purple")
par(mfrow = c(2, 2), mar = rep(0, 4))
for(i in 1:4) {
  crop_ylds %>% filter(country == "ZAF" & crop == "Maize") %>% 
    plot(yields ~ year, data = ., pch = i, col = cols[i], axes = FALSE)
}
# multiple lines
par(mar = rep(3, 4))
# crop_ylds[crop_ylds$country == "ZAF" & crop_ylds$crop == "Maize", ]
crop_ylds %>% filter(country == "ZAF" & crop == "Maize") %>%
  plot(yields ~ year, data = ., pch = i, col = cols[i], axes = FALSE)
# dat <- crop_ylds[crop_ylds$country == "ZAF" & crop_ylds$crop == "Maize", ]
# plot(yields ~ year, data = dat, pch = i, col = cols[i], axes = FALSE)
axis(side = 1, las = 2)
axis(side = 2, las = 2)

crop_ylds %>% filter(country == "ZAF" & crop == "Maize") %>%
  plot(yields ~ year, data = ., col = "blue", type = "l")
crop_ylds %>% filter(country == "ZMB" & crop == "Maize") %>%
  lines(yields ~ year, data = ., col = "red")
```

---
```{r, eval=FALSE}
# change axes
dat <- crop_ylds[crop_ylds$country == "ZAF" & crop_ylds$crop == "Maize", ]
plot(yields ~ year, data = dat, pch = i, col = cols[i], axes = FALSE)
axis(side = 1, las = 2)
axis(side = 2, las = 2)

# multiple panels
par(mfrow = c(2, 1))
crop_ylds %>% filter(country == "ZAF" & crop == "Maize") %>%
  plot(yields ~ year, data = ., col = "blue", type = "l", ylim = c(0, 7))
crop_ylds %>% filter(country == "ZMB" & crop == "Maize") %>%
  plot(yields ~ year, data = ., col = "red", type = "l", ylim = c(0, 7))
```

---
#### Histogram
```{r, eval = FALSE}
# basic
par(mar = c(3, 3, 1, 1))
crop_ylds %>% filter(crop == "Maize") %>% 
  pull(yields) %>% hist(., breaks = seq(0, 7, 0.5), main = "Maize yields")

# side-by-side
par(mfrow = c(1, 2))
crop_ylds %>% filter(crop == "Maize" & country == "ZAF") %>%
  pull(yields) %>% 
  hist(., breaks = seq(0, 7, 0.5), main = "Maize yields", xlim = c(0, 10))
crop_ylds %>% filter(crop == "Maize" & country == "ZMB") %>%
  pull(yields) %>% 
  hist(., breaks = seq(0, 7, 0.5), main = "Maize yields", xlim = c(0, 10))
```

---
### ggplot examples
#### scatters
```{r, eval = FALSE}
# basic points
crop_ylds %>% filter(crop == "Maize") %>% 
  ggplot() + geom_point(aes(x = year, y = yields))

# basic lines
crop_ylds %>% filter(crop == "Maize") %>% 
  ggplot() + geom_line(mapping = aes(x = year, y = yields))

# multiple lines
crop_ylds %>% filter(crop == "Maize") %>% 
  ggplot() + geom_line(aes(x = year, y = yields, color = country)) +
  scale_color_manual(values = c("green", "blue"))

# change axes/theme
crop_ylds %>% filter(crop == "Maize") %>%   
  ggplot() + geom_line(aes(x = year, y = yields, color = country)) +
  scale_color_manual(values = c("green", "blue")) + 
  scale_x_continuous(breaks = seq(1960, 2020, 5), expand = c(0, 0)) +
  theme(axis.text.x = element_text(angle = 90), 
        panel.background = element_blank())
```

---
#### multi-panel
```{r, eval = FALSE}
# from a single variable, facet_grid, facet_wrap
crop_ylds %>% 
  ggplot() + geom_line(aes(x = year, y = yields, color = country)) +
  scale_color_manual(values = c("green", "blue")) + 
  facet_grid(cols = vars(crop))

# multiple variables
p1 <- crop_ylds %>% filter(crop == "Maize") %>% 
  ggplot() + geom_line(aes(x = year, y = yields, color = country)) +
  scale_color_manual(values = c("green", "blue")) 
p2 <- crop_ylds %>% filter(crop == "Maize") %>% 
  ggplot() + geom_line(aes(x = year, y = harv_area, color = country)) +
  scale_color_manual(values = c("green", "blue")) 

gridExtra::grid.arrange(p1, p2, ncol = 2)
g1 <- cowplot::plot_grid(p1 + theme(legend.position = "none"), 
                         p2 + theme(legend.position = "none"))
cowplot::plot_grid(g1, cowplot::get_legend(p1), rel_widths = c(2, 0.2))
```

