---
title: "Geospatial Analysis with R"
subtitle: Class 13
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
## Election Results
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width=8, fig.height=5}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")

cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>% 
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% 
  dplyr::select(-other, -delres) %>% 
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab(NULL) + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

---
```{r, eval = FALSE}
library(tidyverse)
library(rvest)

URL <- "https://history.house.gov/Institution/Party-Divisions/Party-Divisions/"
webpage <- read_html(URL)
cong <- as_tibble(html_table(webpage)[[1]])
onms <- colnames(cong)
newnms <- c("congress", "seats", "D", "R", "other", "delres")
cong <- cong %>% 
  slice((which(grepl("^Republican", `Anti-Administration`))[1] + 1):nrow(.)) %>%
  filter(`Congress (Years)` != colnames(cong)[1]) %>% 
  mutate(year = gsub("(*.*-)|(*.*–)|)|)2", "", `Congress (Years)`)) %>%
  mutate(year = as.numeric(year) - 2) %>% 
  rename_at(vars(onms), ~newnms) %>% select(-other, -delres) %>%
  mutate(seats = substr(seats, 1, 3)) %>% 
  mutate_at(.vars = vars(seats, D, R), as.numeric) %>% 
  mutate(swing = (D - R) / (D + R)) 
cong %>%  
  ggplot() + geom_line(aes(year, swing)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + 
  geom_hline(yintercept = 0, lty = 2, col = "red") + 
  xlab("") + ylab("(D - R) / (D + R)") + 
  ggtitle("Normalized Party Control Index for US House") + theme_linedraw() + 
  scale_x_continuous(breaks = seq(1859, 2019, 10), expand = c(0, 2))
```
---
## Exercises


```{r, message=F}
library(dplyr)
folder_path <- "../inst/extdata/"
files <- list.files(folder_path, pattern = "whittier")
for(file in files){
  load(paste0(folder_path, file))
}

up_data <- whittier_up_d_h %>% dplyr::select(date, 
                           lat, 
                           long,
                           device,
                           ndvi_hourly, 
                           gcvi_hourly, 
                           savi_hourly ,
                           evi2_hourly)

mid_data <- whittier_mid_d_h %>% dplyr::select(date, 
                           lat, 
                           long,
                           device,
                           ndvi_hourly, 
                           gcvi_hourly, 
                           savi_hourly ,
                           evi2_hourly)

down_data <- whittier_down_d_h %>% dplyr::select(date, 
                           lat, 
                           long,
                           device,
                           ndvi_hourly, 
                           gcvi_hourly, 
                           savi_hourly ,
                           evi2_hourly)

## add site_name column
up_data$site_name <- "up"
mid_data$site_name <- "mid"
down_data$site_name <- "down"

## combine using rbind

all_sites_data <- do.call('rbind', list(up_data, mid_data, down_data))
all_sites_data$year <- lubridate::year(all_sites_data$date)

```

---

## Exercises
- (1) Create SAVI time-series for 2019 with different color for each site.

```{r}
ggplot(all_sites_data %>% filter(year == 2019)) +
  geom_line(aes(x = date, y = savi_hourly, col = site_name))
```



---
- (2) Same as above, use facet_grid() to create a separate plot for each year.
```{r, eval = FALSE}
ggplot(all_sites_data) +
  geom_line(aes(x = date, y = savi_hourly, col = site_name)) + 
  facet_grid(cols = vars(year) )
```

---
- (3) Create scatter plot of EVI vs SAVI. Use a different shape for each site, and a different color for each year.
```{r, eval = FALSE}
shapes = c("up" = 15, 
            "mid" = 16,
            "down" = 17)
  
colors = c("2018" = "orange", 
            "2019" = "slateblue",
            "2020" = "#AAAAAA",
           "2021" = 'grey')

ggplot(all_sites_data) +
  geom_point(aes(x = evi2_hourly,
                 y = savi_hourly,
                 shape = site_name,
                 col = as.character(year))) +
  scale_shape_manual(values = shapes) + 
  scale_color_manual(values = colors)
```

---
## Challenge
- Create a scatter plot that compares NDVI at the up site (x) and the mid site (y).
- Use a different color for each year
- Hint: you will need to reshape data
```{r, eval = FALSE}

#> # A tibble: 3 × 2
```



---
## Projects
-[Project list](https://agroimpacts.github.io/geospaar/projects.html#project-list)


---
## Today
- Spatial
  - Non-spatial to spatial
  - Read and write spatial
  - CRS
  - Make data from scratch
  - Spatial properties (area, length)

---
### Non-spatial to spatial

- Read in farmer data
```{r}
library(tidyverse)
library(sf)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() 
farmers_sf <- st_as_sf(farmers, coords = c("x", "y"), crs = 4326)
```


---
## `sf` reads/writes vector data
- format is similar to data.frame/tibble
- `geometry` column makes it "spatial"
- all `dplyr` indexing works, plus spatial functions (buffer, etc.)
```{r}
class(farmers)
farmers %>% slice(1:5)
```

---
## file formats
- .shp has many files, not convenient
- .geojson has 1 file, preferred



---
## reading from csv (with x, y columns)
-Nonspatial

```{r}
farmers <- read_csv(system.file("extdata/farmer_spatial.csv", 
                                package = "geospaar"))
farmers
```

---
## reading from csv (with x, y columns)
-Nonspatial to spatial
```{r}
farmers <- st_as_sf(farmers, coords = c("x", "y"))
farmers
str(farmers)
```

---
## reading from shp (with x, y columns)
- Use `st_read()`
- `dsn` refers to file name
```{r}
fnm <- system.file("extdata/roads.shp", package = "geospaar")
roads <- st_read(dsn = fnm)

fnm <- system.file("extdata/districts.shp", package = "geospaar")
districts <- st_read(dsn = fnm)
```

---
## projections (crs)
- use `st_crs()` to get/set crs
- You need to set crs for features from scratch
```{r}
farmers %>% st_crs()
```

---
## projections (crs)
- Geographic coordinates are: `4326`
```{r}
st_crs(farmers) <- 4326
farmers %>% st_crs()
```


---
## projections (use existing)
- you can set projections based on another layer
- but the two layers MUST have the same crs (e.g. geographic)
```{r}
farmers <- read_csv(system.file("extdata/farmer_spatial.csv", 
                                package = "geospaar"))
farmers <- st_as_sf(farmers, coords = c("x", "y"))

st_crs(farmers) <- st_crs(districts) ## set crs

farmers %>% st_crs() ## print new crs
```

---
## bounding box
- use `st_bbox()`
```{r}
sfdat <- list("districts" = districts,  "roads" = roads, "farmers" = farmers)
sapply(sfdat, function(x) st_bbox(x))
```


---
## projections (transform)
- Use `st_transform()` to project from one crs to another
```{r}
roads <- st_transform(x = roads, crs = st_crs(districts))
st_crs(roads)
#sfdat$roads <- st_transform(x = sfdat$roads, crs = st_crs(districts))
```



---
## plotting (base)
- Use `st_geometry` to plot
- Use `add = TRUE` to add additional plots
```{r}
par(mar = c(0, 0, 0, 0))
plot(st_geometry(districts), col = "grey")
plot(st_geometry(roads), col = "red", add = TRUE)
plot(st_geometry(farmers), pch = 20, reset = FALSE, col = "blue", add = TRUE)
```


---
## plotting (ggplot)
- Plot multiple objects in one line
```{r}
ggplot() + 
  geom_sf(data = sfdat$districts) +
  geom_sf(data = sfdat$roads, col = "red") + 
  geom_sf(data = sfdat$farmers, col = "blue")
```

---
## plotting (ggplot)
- Good for plotting attributes
```{r}
ggplot(sfdat$districts) + geom_sf() +
  geom_sf(data = sfdat$roads, col = "red") + 
  geom_sf(data = sfdat$farmers, col = "blue")
```


---
## features from scratch
```{r}
pt <- st_point(x = c(28, -14))
pt
#> POINT (28 -14)
#
# #2
pts <- st_multipoint(x = cbind(x = c(27.5, 28, 28.5), y = c(-14.5, -15, -15.5)))
pts
#> MULTIPOINT ((27.5 -14.5), (28 -15), (28.5 -15.5))
#
# #3
sline <- st_linestring(cbind(x = c(27, 27.5, 28), y = c(-15, -15.5, -16)))
sline
#> LINESTRING (27 -15, 27.5 -15.5, 28 -16)
#
# #4
pol <- st_polygon(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5), 
                             y = c(-15.5, -16.5, -17, -16, -15.5))))
pol
```

---
## Plotting multiple objects

```{r}
par(mar = c(0, 0, 0, 0))
plot(districts %>% st_geometry(), col = "grey")
plot(pt, add = TRUE, col = "red", pch = 20)
plot(pts, add = TRUE, col = "blue", pch = 20)
plot(sline, add = TRUE, col = "cyan", lwd = 2)
plot(pol, add = TRUE, col = "orange", lwd = 2)
```

---
## Area and length
```{r}
dist_areas <- districts %>% st_area()
dist_areas
dist_areas %>% units::set_units("km^2")
```
---

## add area as column
```{r}
districts <- districts %>% 
  mutate(area = st_area(.) %>% units::set_units("km^2") %>% as.numeric()) 
districts
```

---
## Plot districts (exercise)
- Use `geom_sf()` to plot area
- Fill districts based on area column.
- Use `scale_fill_continuous` , with `type = 'viridis'`
```{r}

```

---
## Calculate length (exercise)
- Use `st_length()` to add colummn 'road_length' to `roads`
- 'road_length' should be in km
- Use ggplot to plot districts with green outline (color), and empty fill (fill = NA)
- Then plot roads based on length
- Use `scale_color_continuous()` 

```{r}

```





---
## Homework
- Read Unit 2 Module 1, parts 4-5
- Review `sf` vignettes 
  - [overview](https://r-spatial.github.io/sf/articles/sf1.html)
  - [reading/writing/converting](https://r-spatial.github.io/sf/articles/sf2.html)
  - [geometry operations](https://r-spatial.github.io/sf/articles/sf3.html)

