---
title: "Geospatial Analysis with R"
subtitle: Class 19
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Homework review
- Raster intro, continued
  - Terrain/climate analysis
- Exercises

---

## Data
```{r, eval=FALSE}
library(geospaar)
districts <- read_sf(
  system.file("extdata/districts.shp", package = "geospaar")
) %>% st_transform(4326)
data(chirps)
chirpsz <- mask(chirps, districts)
# chirpsz %>% projectRaster(., crs = "ESRI:102022", res = 5000)
```

---
## Homework
- Create a dummy raster (sampling from 1:100) using district 49 for extent and res of 0.1

```{r, eval=FALSE}
r <- raster(extent(districts[49, ]), res = 0.1, crs = 4326)
set.seed(1)
values(r) <- sample(1:100, size = ncell(r), replace = TRUE)
```

- Reproject to Albers
```{r, eval=FALSE}
ralb <- projectRaster(r, crs = "ESRI:102022", res = 11000)
```
- Disaggregate to 0.02 degrees
```{r, eval=FALSE}
r02 <- disaggregate(r, fact = 5)
r02b <- disaggregate(r, fact = 5, method = "bilinear")
plot(stack(r02, r02b))
```

---

- Calculate the sum of rainfall from `chirps`
```{r, eval=FALSE}
raintot <- calc(chirps, sum)
raintot2 <- calc(chirpsz, sum)
plot(stack(raintot, raintot2))
```
- Identify all areas of total rainfall > 10 mm
```{r, eval=FALSE}
rain10 <- raintot2 > 10
rain100 <- raintot2 > 100
plot(stack(rain10, rain100))
```
- Calculate the mean rainfall in Zambia for the 15th day in `chirps`
```{r, eval=FALSE}
rainmean <- calc(chirpsz[[15]], mean)
plot(rainmean)
```

---
## Terrain analysis with zonal statistics

```{r, eval=FALSE}
districts <- districts %>% mutate(ID = 1:nrow(.))
distsr <- districts %>% rasterize(x = ., y = raintot, field = "ID") %>% print()
plot(distsr)

# Download DEM
dem <- getData(name = "alt", country = "ZMB", path = tempdir())

# calculate slope
slope <- terrain(x = dem, opt = 'slope', unit = 'degrees')
plot(terrain(x = dem, opt = 'aspect', unit = 'degrees'))

# calculate mean by district
distsr_rs <- resample(x = distsr, y = slope, method = "ngb")
zoneslope <- zonal(x = slope, z = distsr_rs, fun = "mean")
zoneelevation <- zonal(x = dem, z = distsr_rs, fun = "mean")

# map zonal statistics
distr_slopezone <- zoneslope %>% data.frame %>%  
  subs(x = distsr_rs, y = ., by = "zone")
distr_elezone <- zoneelevation %>% data.frame %>% 
  subs(x = distsr_rs, y = ., by = "zone")

# plot
l <- list(distr_elezone, distr_slopezone)
titles <- c("Elevation", "Slope")
par(mfrow = c(1, 2))
for(i in 1:length(l)) {
  plot_noaxes(l[[i]], main = titles[i])
}
```

---
## Climate data summaries and zonal stats

```{r, eval=FALSE}
# rainfall <- getData(name = "worldclim", res = 10, var = 'tmax',
#                     path = tempdir())
dt <- as.Date(gsub("Y", "", names(chirpsz)), format = "%y%j")
dt <- lubridate::parse_date_time(gsub("Y", "", names(chirpsz)), orders = "yj")
wk <- lubridate::week(dt)
weekly_rainfall <- lapply(unique(wk), function(x) {
  calc(chirpsz[[which(wk == x)]], sum)
}) %>% stack(.)
rflim <- range(cellStats(weekly_rainfall, range))
plot(stack(weekly_rainfall), zlim = rflim)
```

---
## Exercises

- Use the dem function to calculate the terrain aspect in Zambia. 
- Reclassify the aspect into North, South, East, West (classes 1:4)
- Calculate the most frequently occurring aspect in each district (use the `modal`) function
- Calculate the average total rainfall (from `raintot`) in each aspect class







