---
title: "Geospatial Analysis with R"
subtitle: Class 15
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Manipulation between raster and vector
- Raster calculation
  - Local calculation
  - Aggregate, focal, and zonal calculation (Neighborhood)
  - Global calculation

---

## Reproject

```{r, eval=F}
# Use a defined CRS object
rst_lonlat <- projectRaster(rst, crs = CRS("+init=epsg:4326"))

# Or get crs from an existing object
rst <- projectRaster(rst_lonlat, crs = crs(terrvars))

# Or directly use an existing object
rst <- projectRaster(rst, rst_lonlat)
```

---

## Get data (`getData`)

`getData` is a function in `raster` to query supported raster datasets.

```{r, eval=F}
# Another powerful package to visualize geospatial data.
# Use ggplot2 syntax style
library(tmap)
srtm <- getData('alt', country = 'TZA', path = tempdir())

# tmap_mode("view") # convert to interactive viewing
# tmap_mode("plot") # convert to static plotting
tm_shape(srtm, raster.downsample = F) + 
  tm_raster(midpoint = NA, 
            n = 10, style = 'pretty') +
  tm_legend(legend.position = c('left', 'bottom'))
```

---

## Practice

- Get boundary vector of countries of Africa as a `sf` using function `ne_countries` in package `rnaturalearth`, name it `cnts`.
- Get the centroid of these countries, and name it `cnts_ctds`.
- Use `getData` to download precipitation from worldclim with resolution 10 second, name it `rainfall`. Should have 12 layers for each month.
- Plot them out together.

---

```{r, eval=F}
# Practice
rainfall <- getData('worldclim', var = 'prec', 
                res = 10, path = tempdir())
cnts <- ne_countries(continent = 'africa', 
                     returnclass = 'sf')
cnts_ctds <- st_centroid(cnts)
```

---

## Raster and vector

- `rasterize`: vector to raster
- `rasterToPolygons`, `rasterToPoints`, etc.: raster to vector
- Overlay
- `extract`

```{r, eval=F}
# Overlay: crop and mask
rainfall_africa <- mask(rainfall, cnts)
rainfall_africa <- mask(crop(rainfall, cnts), cnts)

# Raster to vector
cnts_rst <- rasterize(cnts, rainfall_africa$prec1)

# vector to raster, sp object
cnts_vct <- rasterToPolygons(cnts_rst, dissolve = T)

# Extract
vals_ctds <- extract(rainfall_africa, cnts_ctds)
vals_ctds <- extract(rainfall_africa, cnts_ctds, cellnumbers = TRUE)
vals_ctds <- extract(rainfall_africa, cnts_ctds, sp = TRUE)
```

---

## Raster calculation

- Local

  Local operations do operations **cell-by-cell** in one or several layers.
  
  Let's do a practice:
  - Download sentinel-2 bands in 20m (R-G-B-NIR) from [here](https://www.dropbox.com/s/xv37lzch4sgsfnm/S2A_MSIL2A_20201014T154231_N0214_R011_T19TBG_20201014T200156_20m.tif?dl=0). 
  - Read the raster stack into R.
  - Make a simple water mask using NDVI. (What value of NDVI indicate water?)

---

```{r, eval=F}
fname <- 'S2A_MSIL2A_20201014T154231_N0214_R011_T19TBG_20201014T200156_20m.tif'
s2 <- stack(
  file.path('/Users/leisong/Downloads', 
            fname))
names(s2) <- paste0('B', c(2:4, 8))
ndvi <- (s2$B8 - s2$B4) / (s2$B8 + s2$B4)
water_mask <- ndvi < 0
```

---

## Raster calculation

- Local
  - `reclassify` takes a matrix of values to reclassify a raster.
  
```{r, eval=F}
# A clear way to define the reclassify matrix
rcl <- matrix(
  c(-1, 0, 1, # class 1
    0, 0.2, 2, # class 2
    0.2, 0.5, 3, # class 3
    0.5, 1, 4), # class 4
  ncol = 3, byrow = T)

lc <- reclassify(ndvi, rcl, include.lowest = T)
lc <- ratify(lc)
levels(lc)[[1]]$landcover <- c(
  "water", "Urban",
  "sparse vegetation",
  "dense vegetation"
)
tm_shape(lc) + tm_raster()
```

---

## Raster calculation

- "Neighborhood"
  
  - Function `aggregate`: `aggregate(x, fact, fun)`
  - Function `zonal`: `zonal(x, z, fun)`
  - Function `focal` - image filters: `focal(x, w, fun)`
  
```{r, eval=F}
# Aggregate, result is a raster with coarse resolution
rainfall_af_mean <- mean(rainfall_africa)
rainfall_agg <- aggregate(rainfall_af_mean, fact = 8, fun = mean)

# Zonal, result is a matrix 
rainfall_zonal <- zonal(rainfall_af_mean, cnts_rst, fun = max)
```

---

```{r, eval=F}
# Focal, result is a raster as well.
## Use s2 NIR band as an example
fname <- 'S2A_MSIL2A_20201014T154231_N0214_R011_T19TBG_20201014T200156_20m.tif'
nir_band <- raster(
  file.path('/Users/leisong/Downloads', 
            fname), 4)
# weight matrix could not the square, 
# but ncol and nrow must be uneven numbers
## Reproduce Sobel edge detection
wx <- matrix(c(-1, -2, -1, 0, 0, 0, 1, 2, 1), nrow = 3)
wy <- matrix(c(1, 0, -1, 2, 0, -2, 1, 0, -1), nrow = 3)
nir_x <- focal(nir_band, w = wx)
nir_y <- focal(nir_band, w = wy)
nir_edge <- sqrt(nir_x^2 + nir_y^2)
writeRaster(nir_edge, '/Users/leisong/downloads/edge.tif')
```

---

## Raster calculation

- Global
  - Statistics: e.g. `cellStats(rainfall_af_mean, sd)`, `hist(rainfall_af_mean)`
  - distance

```{r, eval=F}
# Calculate the distance to the farmers
library(sf)
rst_template <- raster(system.file('extdata/demalb.tif', package = 'geospaar'))
farmers <- read.csv(system.file('extdata/farmer_spatial.csv', package = 'geospaar'),
                    stringsAsFactors = F) %>% 
  st_as_sf(coords = c('x', 'y'), crs = 4326) %>%
  st_transform(crs(rst_template)) %>% 
  as_Spatial()
farmers_rst <- rasterize(farmers, field = 1, rst_template)
dis_farmers <- raster::distance(farmers_rst)
```

---

## Next class

- Terrain analysis
- Interpolation and prediction
- Fit a model (a case study)

---

## Homework

Use operations we learned so far to make a mini case analysis on raster that can get by `getData` or the provided sentinel-2 bands or whatever images that you can get. Then tell us the story in next class.
