---
title: "Geospatial Analysis with R"
subtitle: Class 15
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


## Today

- More vector analysis and exercises

---

Data:  read in `farmers` and make into `sf`, `districts`, and `roads`
```{r, eval=FALSE}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sf(coords = c("x", "y"))
districts <- st_read(
  system.file("extdata/districts.shp", package = "geospaar")
)
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
```

---
## Adding to data

Add some categories
```{r, eval=FALSE}
set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
```
---

## Analyses
```{r, eval=FALSE}
# districts2 %>% dplyr::select(yield) %>% summarize_all(funs(mean, sd))
# districts2 %>% summarize(mean(yield), sd(yield))
districts2 %>% group_by(yld_cat) %>% 
  summarize(yield = mean(yield)) %>% 
  ggplot() + geom_sf(aes(fill = yield))
  # geom_sf(aes(fill = yld_cat))
```

---
## Joins
```{r, eval=FALSE}
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))
farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))

smallest_centroid <- districts3 %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter(area == min(area)) %>% st_centroid()
sf_use_s2(use_s2 = FALSE)
smallest_buffer <- smallest_centroid %>% st_buffer(dist = 0.5)
buff_int_dists <- st_intersection(smallest_buffer, districts3)
```

---
- Differencing
```{r, eval=FALSE}
dif <- st_difference(districts3, smallest_buffer)
ggplot(dif) + geom_sf(fill = "grey50")

dif2 <- st_difference(smallest_buffer, districts3)
ggplot(dif2) + geom_sf(fill = "grey50")
```

- Unioning

```{r, eval=FALSE}
uni <- st_union(dif)
ggplot(uni) + geom_sf(fill = "grey50")
uni2 <- st_union(dif, smallest_buffer)
ggplot(uni2) + geom_sf(fill = "grey50")
uni3 <- st_union(dif, buff_int_dists)
ggplot(uni3) + geom_sf(fill = "grey50")

uni3
```

---
- Sampling

```{r, eval=FALSE}
# simple within extent
set.seed(1)
sample100 <- districts3 %>% filter(n > 0) %>% 
  st_sample(size = 100, exact = TRUE)
ggplot(districts3) + geom_sf() + 
  geom_sf(data = districts3 %>% filter(n > 0), fill = "lightblue") + 
  geom_sf(data = sample100, size = 0.5, col = "red")

# stratified by polygon
ndists <- districts3 %>% filter(n > 0) %>% nrow(.)
strat_sample <- districts3 %>% filter(n > 0) %>% 
  st_sample(size = rep(5, ndists), exact = TRUE)

ggplot(districts3) + geom_sf() + 
  geom_sf(data = districts3 %>% filter(n > 0), fill = "lightblue") + 
  geom_sf(data = strat_sample, size = 0.5, col = "red")
```

---
### Exercises

#### Round 1
- Calculate the length of roads in kilometers
- Buffer roads > 100 km by 30 kilometers, save as new object
- Buffer roads > 100 km 10 kilometers, save as new object
- Take the difference between the 30 km and 10 km buffers, i.e. erase/difference the two
- Plot the resulting difference using `ggplot` with the difference in red over Zambia's districts

```{r, eval=FALSE}
# code to adapt
roads %>% mutate(km = as.numeric(st_length(.)) / 1000) %>% 
  filter(km > 200) %>% 
  st_buffer(dist = 20000) %>% 
  st_transform(crs = 4326)
st_difference(larger_object, smaller_object)
ggplot() + geom_sf

```

---

#### Round 2
- Randomly sample 100 points within the 30 km road buffer (use a seed of 1)
- Plot the 30 km buffer (red) the points in it (blue) over the districts

```{r, eval=FALSE}
buffered_object %>% st_sample()
```

---
### How would you solve these problems? 

- Calculate the area of Zambia's districts 

    - mutate on districts

- **Accurately** calculate the area of Zambia's districts
- Find the largest and smallest districts in Zambia
- Find the longest road in Zambia
- Find the names of the districts that Zambia's longest road runs through
- Calculate the area of each of those districts along Zambia's longest road that lies within 10 km distance of that road
- Create a 0.5 (E-W) X 0.25 (N-S) degree polygon centered on Zambia's smallest district
- Count how many farmers from our sample live within 10 km of Zambian roads that are longer than 200 km

---
### Solve the problems!
