---
title: "Geospatial Analysis with R"
subtitle: Class 14
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- `sp` and convert between `sp` and `sf`.
- Read and write raster data.
- Raster and its attributes.

---

## `sp`

Package `sp` and `rgdal`.

- `readOGR()` and `writeOGR()` to read and write `sp` vectors.

```{r, eval=F}
library(rgdal)
# Read
dsn <- system.file('extdata/districts.shp', package = 'geospaar')
ogrInfo(dsn) # A list, can check info without opening the file, which is great.
districts <- readOGR(dsn)
# Check
districts
class(districts)
plot(districts)
# Base R syntax to do operation
## An example: subset
zambezi <- districts[districts$distName == 'Zambezi', ]
plot(zambezi, add = T, col = 'red')
# Write, must set driver, 
# use ogrDrivers() to get the list of supported drivers.
# Also, must set layer.
writeOGR(districts, '/Users/leisong/downloads/districts.geojson',
         layer = 'districts', driver = 'GeoJSON')
```

---

## Convert between `sf` and `sp`

```{r, eval=F}
library(sf)
# Convert from `sp` to `sf`
districts_sf <- districts %>% 
  st_as_sf()

# Convert from `sf` to `sp`
districts <- as_Spatial(districts_sf)
# Or
districts <- as(districts_sf, 'Spatial')
```

---

## Practice

- Read `roads.shp` from package `geospaar` as a `sf`.
- Convert it to `sp`.
- Save it out to a path out of your package in shapefile format with name `roads_sp.shp`.
- Read `roads_sp.shp` back in R as a `sp`.
- Convert it to `sf`.
- Plot it out using `ggplot2`.

```{r, eval=F, echo=F}
# Part 1
library(sf)
library(ggplot2)
roads <- read_sf(system.file('extdata/roads.shp', package = 'geospaar'))
roads_sp <- as_Spatial(roads)
writeOGR(roads_sp, '/Users/leisong/downloads/roads_sp.shp',
         layer = 'roads', driver = 'ESRI Shapefile')

# Part 2
roads_sp <- readOGR('/Users/leisong/downloads/roads_sp.shp')
st_as_sf(roads_sp) %>% ggplot() +
  geom_sf() +
  theme_bw()
```

---

## Read and write raster (`raster`)

`raster`, `stack`, or `brick` to read raster.

- `raster` just reads one band.
- `stack` or `brick` could read multiple bands.

```{r, eval=F}
# be careful of function select in raster
# it conflicts with select function in dplyr
library(raster)
dsn <- system.file('extdata/terrvars.tif', package = 'geospaar')
# read one band
terrvars <- raster(dsn, band = 2)
terrvars
# Read raster stack
terrvars <- stack(dsn)
# Or read raster brick
terrvars <- brick(dsn)
terrvars
```

- `writeRaster` to save out a raster. Check help document for arguments.

```{r, eval=F}
writeRaster(terrvars, '/Users/leisong/downloads/terrvars.tif',
            overwrite = TRUE)
```

---

## Raster attributes

```{r, eval=F}
# Check info
res(terrvars)
crs(terrvars)
extent(terrvars)
nlayers(terrvars)
names(terrvars)

# Get pixel values
vals <- getValues(terrvars)
# or 
values(terrvars)

# Get one layer using $
terrvars$terrvars.1
```

---

## Create a raster

- Function `raster`

```{r, eval=F}
# From extent
e <- extent(terrvars)
rst <- raster(e, nrows = 10, ncols = 10, 
              crs = crs(terrvars), # important
              vals = 1:100)
plot(rst)

# From scratch
rst <- raster(nrows = 10, ncols = 10, 
              xmn = -314760.4, xmx = 903239.6,
              ymn = -2122500, ymx = -970500.1)
crs(rst) <- crs(terrvars)
values(rst) <- 1:100

# From another raster
rst <- raster(terrvars) # Just a blank raster, no values
# Choose layer
rst <- raster(terrvars, 4) # Then have the same values with the layer
```

---

## Reproject

```{r, eval=F}
# Use a defined CRS object
rst_lonlat <- projectRaster(rst, crs = CRS("+init=epsg:4326"))

# Or get crs from an existing object
rst <- projectRaster(rst_lonlat, crs = crs(terrvars))

# Or directly use an existing object
rst <- projectRaster(rst, rst_lonlat)
```

---

## Get data (`getData`)

`getData` is a function in `raster` to query supported raster datasets.

```{r, eval=F}
srtm <- getData('alt', country = 'TZA', path = tempdir())
plot(srtm)
```

*NOTE:* their server seems down at this moment.

---

## Practice

- Get boundary vector of a country that you are interested in.
- Download alt using `getData` for this country.
- Project boundary vector and alt to albert.
- Plot them out together.

---

## Next class

- Raster calculation
- Reclassification
- Manipulation between raster and vector

---

## Homework

- Finish assignment 4. 
- Read Unit2 - Module 2.
