---
title: "Geospatial Analysis with R"
subtitle: Class 16
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class4plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Terrain analysis
- Interpolation and prediction
- Fit a model (a case study)

---

## Terrain analysis

- Function `terrain`
- Function `hillshade`

These functions are based on focal operation fundamentally.

```{r, eval=F}
library(raster)
# terrain
dem <- getData(name = 'alt', country = 'TZA', path = tempdir())
slope <- terrain(dem, opt = "slope", unit = "radians", neighbors = 8)
aspect <- terrain(dem, opt = "aspect", neighbors = 4)
flowdir <- terrain(dem, opt = "flowdir", neighbors = 4)

# hillshade
# have to define an angle and direction
hillshade <- hillShade(slope, aspect, 40, 270)
```

---

## Interpolation and prediction

**What is spatial interpolation?**

*NOTE: we could do interpolation for any numeric variables, but it does not mean the interpolation makes sense.*

- Nearest neighbor interpolation
- Inverse distance weighted interpolation (IDW)
- Kriging (Ordinary Kriging, Universal Kriging, etc.)

---

```{r, eval=F}
library(raster)
library(dplyr)
library(rnaturalearth)
gha <- ne_countries(continent = 'africa', returnclass = 'sf') %>%
  filter(name == 'Ghana')
rainfall <- getData(name = 'worldclim', var = 'prec', res = 2.5) %>% 
  calc(., mean)
rainfall <- mask(crop(rainfall, gha), gha)

# Get some pseudo samples
# sampleRandom, sampleStratified
set.seed(10)
pts_rainfall <- sampleRandom(rainfall, size = 200,
                             sp = T) %>% st_as_sf() %>% 
  rename(rainfall = layer) %>% 
  mutate(id = row_number())
plot(rainfall, breaks = seq(60, 180, 10),
     col = terrain.colors(length(seq(60, 180, 10))))
plot(pts_rainfall$geometry, add = T, 
     pch = 20, col = 'grey', size = 1)

# Split points to train and test (a MUST for any model)
set.seed(11)
pts_train <- pts_rainfall %>% 
  sample_frac(0.75)
pts_test <- pts_rainfall %>% 
  filter(!id %in% pts_train$id)
```

---

## Neareast neighbour interpolation

```{r, eval=F}
library(gstat)
library(dplyr)
library(magrittr)

# step 1 - Define a raster template for the result
rst_template <- raster(rainfall)

# Step 2 - Define gstat object that hold all necessary info for interpolation
nn_gs <- gstat(formula = rainfall ~ 1, 
               locations = pts_train, 
               nmax = 5, set = list(idp = 0))

# Step 3 - Do the prediction
rainfall_nn <- interpolate(rst_template, nn_gs) %>% 
  mask(., gha)

# Step 4 - Evaluation
# Also could use function predict(nn_gs, pts_test) to
# get the evaluation table
pred_test <- raster::extract(rainfall_nn, 
                     pts_test, sp = T) %>% 
  st_as_sf() %>% st_drop_geometry() %>% 
  dplyr::select(-id) %>% 
  magrittr::set_names(c('observation', 'prediction'))
```

---

```{r, eval=F}
# visualization
library(plotly)

# Evaluation
f <- list(family = "Courier New, monospace",
          size = 18, color = "#7f7f7f")
plot_ly(data = pred_test, x = ~observation, y = ~prediction,
        color = 'red') %>% 
  layout(xaxis = list(title = 'Real max temperature',
                      titlefont = f),
         yaxis = list(title = 'Nearest neighbor interpolation',
                      titlefont = f))
plot_ly(data = pred_test, y = ~observation,
        color = 'red', type = 'box', name = 'Real max temperature') %>% 
  add_boxplot(data = pred_test, y = ~observation,
        color = 'blue', name = 'NN interpolation')

# Maps
par(mfrow = c(1, 2))
plot(rainfall, breaks = seq(60, 180, 10),
     col = terrain.colors(length(seq(60, 180, 10))),
     main = 'Real max temperature')
plot(rainfall_nn, breaks = seq(60, 180, 10),
     col = terrain.colors(length(seq(60, 180, 10))),
     main = 'Nearest neighbor interpolation')
```

---

## Inverse distance weighted interpolation (IDW)

Nearest neighbor interpolation is a special case of IDW, which not use weights at all. In function `gstat`, we set `set = list(idp = 0)`.

Now, it is your turn to create a map using IDW. And compare it with the real max temperature and nearest neighbor interpolation.

- Use function `gstat`
- Or use function `idw`, use `raster` to convert result to a raster.

---

```{r, eval=F}
# Step 2 - Define gstat
idw_gs <- gstat(formula = rainfall ~ 1, 
               locations = pts_train, 
               nmax = 5)

# Step 3 - Do the prediction
rainfall_idw <- interpolate(rst_template, idw_gs) %>% 
  mask(., gha)

imgs <- stack(rainfall,
              rainfall_nn,
              rainfall_idw)
names(imgs) <- c('Real max temperature',
                 'NN interpolation',
                 'IDW interpolation')
tm_shape(imgs) + 
  tm_raster(title = 'Rainfall',
            breaks = seq(60, 180, 10),
            palette = "-RdYlGn", 
            n = length(seq(60, 180, 10))) 
```

---

## Kriging

NN and IDW don't use variogram. But Kriging methods use variogram. So besides the formula, we need to fit a variogram as well.

```{r, eval=F}
# Ordinary kriging
# Step 1 - define raster template (done before)
# Step 2 - initially define gstat object
ok_gs <- gstat(formula = rainfall ~ 1, 
               locations = pts_train)
# Step 3 - fit variogram
v <- variogram(krg_gs, width = 20)
## Option 1: manually fit variogram
# v_mod <- fit.variogram(v, vgm(85, "Exp", 75, 20))
# Option 2: auto fit
library(automap)
v_mod <- autofitVariogram(rainfall ~ 1, as(pts_train, "Spatial"))
plot(v_mod)
# Step 4 - update gstat object with fitted variogram model
ok_gs <- gstat(formula = rainfall ~ 1, model = v_mod$var_model, 
               data = pts_train)
# Step 5 - interpolate
rainfall_ok <- predict(ok_gs, as(rst_template, 'SpatialGrid'))
rainfall_ok <- raster(rainfall_ok['var1.pred']) %>% 
  mask(., gha)
# Step 6 - Evaluation (omitted here)
```

---

```{r, eval=F}
imgs <- stack(rainfall,
              rainfall_nn,
              rainfall_idw,
              rainfall_ok)
names(imgs) <- c('Real max temperature',
                 'NN interpolation',
                 'IDW interpolation',
                 'Ordinary Kirging')
tm_shape(imgs) + 
  tm_raster(title = 'Rainfall',
            breaks = seq(60, 180, 10),
            palette = "-RdYlGn", 
            n = length(seq(60, 180, 10)))
```

---

## Next class

- Review reproducible research
- Review Git commands
- Practice collaborating via GitHub

---

## Homework

- Assignment 5
- Gather the code of your leaflet map and the mini analysis, we will use them in next class
- Decide your final project
