---
title: "Geospatial Analysis with R"
subtitle: Class 12
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message=FALSE, eval=FALSE, warning=FALSE}
library(leaflet)
library(sf)
library(dplyr)
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
bb <- unname(st_bbox(districts))
xy <- st_centroid(districts) %>% st_coordinates() %>% 
  bind_cols(name = districts$distName, .)
slist <- list("color" = "white")
label_opts <- labelOptions(noHide = TRUE, style = slist, direction = 'center',
                           textOnly = TRUE, textsize = "5px")

m <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  fitBounds(bb[1], bb[2], bb[3], bb[4]) %>% 
  addPolygons(data = districts, fill = FALSE, color = "white", 
              group = "Districts", weight = 1) %>% 
  addLabelOnlyMarkers(xy$X, xy$Y, label = xy$name, group = "Names",
                      labelOptions = label_opts) %>%
  addLayersControl(overlayGroups = c("Districts", "Names"),
                   options = layersControlOptions(collapsed = FALSE,
                                                  autoZIndex = FALSE))
```
```{r, eval=FALSE, echo=FALSE}
library(htmlwidgets) 
saveWidget(m, "docs/figures/zambialeaflet.html")
```


---

<iframe seamless src="figures/zambialeaflet.html" width="100%" 
height="500"></iframe>


---

## Today

- More practice with reshaping/analysis/plotting 
- Towards spatial
  - Non-spatial to spatial
  - Read and write spatial
  - CRS
  - Make data from scratch
  - Spatial properties

---


## More practice
### Part 1
- read in the three `geospaar` FAO datasets into a list using `lapply`
- Reshape each of the three datasets by using `pivot_wider` to put harvested area into a new variable **harv_area** and production into **prod**, with the other variables being **crop**, **country**, and **year**. Necessary functions: `%>%`, `select`, `pivot_wider`, `rename`. 
- Combine all these into a single data.frame `crops_df`
- Best: combine the previous two points using `lapply` and `do.call(rbind)`
- Okay: Do this separately for each dataset and then combine with `rbind`

---

- read in the three `geospaar` FAO datasets into a list using `lapply`

```{r, eval=FALSE}
library(geospaar)
fs <- dir(system.file("extdat", package = "geospaar"), 
          pattern = "FAOSTAT", full.names = TRUE)
crops <- lapply(fs, readr::read_csv)
```

Should see...
```
[[1]]
# A tibble: 228 x 14
   `Domain Code` Domain `Area Code` Area  `Element Code` Element
   <chr>         <chr>        <dbl> <chr>          <dbl> <chr>  
 1 QC            Crops          202 Sout…           5312 Area h…
 2 QC            Crops          202 Sout…           5312 Area h…
# … with 218 more rows, and 8 more variables: `Item Code` <dbl>,
...

[[2]]
# A tibble: 228 x 14
   `Domain Code` Domain `Area Code` Area  `Element Code` Element
   <chr>         <chr>        <dbl> <chr>          <dbl> <chr>  
 1 QC            Crops          202 Sout…           5312 Area h…
 2 QC            Crops          202 Sout…           5312 Area h…

...

[[3]]
# A tibble: 228 x 14
   `Domain Code` Domain `Area Code` Area  `Element Code` Element
   <chr>         <chr>        <dbl> <chr>          <dbl> <chr>  
 1 QC            Crops          202 Sout…           5312 Area h…
 2 QC            Crops          202 Sout…           5312 Area h…
...
```
---

- Reshape each of the three datasets in `crops` by using `pivot_wider` to put harvested area into a new variable **harv_area** and production into **prod**, with the other variables being **crop**, **country**, and **year**. 
- Combine all these into a single data.frame `crops_df`


```{r, eval=FALSE}
crops_df <- do.call(rbind, lapply(crops, function(j) {
  x %>% select(Item, Area, Element, Year, Value) %>% 
    pivot_widr(names = Element, values = Value) %>% 
    rename(crop = Item, country = Area, year = Year, 
           harv_area = Area harvested, prod = Production)
}))
crops_df
```

Should see:

```
> crops_df
# A tibble: 342 x 5
   crop  country       year harv_area    prod
   <chr> <chr>        <dbl>     <dbl>   <dbl>
 1 Maize South Africa  1961   4118000 5293000
 2 Maize South Africa  1962   4341000 6024000
 3 Maize South Africa  1963   4339000 6127000
 4 Maize South Africa  1964   4433000 4310000
 5 Maize South Africa  1965   4290000 4608000
 6 Maize South Africa  1966   4241000 5161000
 7 Maize South Africa  1967   4589000 9802000
 8 Maize South Africa  1968   4728000 5358000
 9 Maize South Africa  1969   4387000 5378000
10 Maize South Africa  1970   4418000 6179000
# … with 332 more rows
```

---
#### Pipeline example
```{r, eval = FALSE}
fs <- dir(system.file("extdata", package = "geospaar"), 
          pattern = "maize", full.names = TRUE)
dat <- readr::read_csv(fs) %>% 
  dplyr::select(Item, Area, Element, Year, Value) %>% 
  pivot_wider(names_from = Element, values_from = Value) %>% 
  rename(harv_area = `Area harvested`, prod = Production, country = Area) %>% 
  mutate(yield = prod / harv_area) %>% 
  dplyr::select(country, harv_area, prod, yield) %>% 
  group_by(country) %>% 
  summarise_all(list(mu = mean, sdev = sd))
```

---
### Part 2 - plotting
Using `crops_df`:
- Using `ggplot2`, make a line plot of the annual maize production by country, using red for Zambia and blue for South Africa
- Use `ggplot2` to make three side-by-side line plots of production per year for all three crops by country (red for Zambia, blue South Africa). You need `facet_grid` for this. Do the same for yield (you will need to a new column *yield*)
- Plot the linear regression fit between year for wheat yields, using `ggplot` and `geom_smooth`, with `method = "lm"`. Make a separate series and fit for South African and Zambian yields.

---
- Use `ggplot2` to make three side-by-side line plots of production for all three crops by country (red for Zambia, blue South Africa). Do the same but for yield. You need `facet_grid` for this.

```{r, eval=FALSE}
ggplot(crops_df) + 
  geom_line(aes(x = year, y = prod, color = country)) +
  facet_grid(crop)
```

---

- Adapt the code from the pipeline example to calculate mean wheat yields for Zambia and South Africa. 

---

### Non-spatial to spatial

- Read in farmer data
```{r, eval = FALSE}
library(tidyverse)
library(sf)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() 
farmers_sf <- st_as_sf(farmers, coords = c("x", "y"), crs = 4326)
```

- Write and read the spatial data
```{r, eval=FALSE}
write_sf(farmers_sf, dsn = "external/notebooks/data/farmers_spatial.geojson")
farmers_sf <- read_sf("external/notebooks/data/farmers_spatial.geojson")
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
```

---
  
### Wrangle spatial data
```{r, eval = FALSE}
farmers %>% filter(uuid == "009a8424")
farmers_sf %>% filter(uuid == "009a8424")
farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% 
  arrange(n) %>% st_geometry %>% plot(pch = 16, cex = 0.5)
farmers_sf %>% arrange(uuid)
class(farmers_sf)
# intersection with st_join and left = FALSE to remove plots outside of Zambia
farmers_sf_trim <- farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% 
  st_join(., districts, left = FALSE)
plot(st_geometry(districts))
farmers_sf %>% group_by(uuid) %>% summarize(n = n()) %>% st_geometry %>% 
  plot(pch = 16, add = TRUE, cex = 1)
farmers_sf_trim %>% st_geometry %>% 
  plot(pch = 16, add = TRUE, col = "red", cex = 0.5)
```

---
### Make spatial data

```{r, eval = FALSE}
pts <- st_multipoint(x = cbind(x = c(27.5, 28, 28.5), y = c(-14.5, -15, -15.5)))
pts <- st_geometry(pts) %>% st_as_sf() %>% st_cast("POINT") %>% 
  cbind(ID = 1:nrow(.), .)
pol <- st_polygon(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5), 
                             y = c(-15.5, -16.5, -17, -16, -15.5))))
pol <- st_geometry(pol) %>% st_as_sf() %>% cbind(ID = 1:nrow(.), .)
sline <- st_linestring(cbind(x = c(27, 27.5, 28), y = c(-15, -15.5, -16)))
sline <- st_geometry(pol) %>% st_as_sf() %>% cbind(ID = 1:nrow(.), .)
```

---
### Spatial properties
```{r, eval=FALSE}
pol %>% st_area
pol %>% st_sf(crs = 4326) %>% st_area
sline %>% st_sf(crs = 4326) %>% st_length
```

---
### Spatial transforms
```{r, eval=FALSE}
pol %>% st_sf(crs = 4326) %>% st_transform(., st_crs(roads)) %>% st_area
```




