---
title: "Geospatial Analysis with R"
subtitle: Class 15
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message=FALSE, eval=FALSE, warning=FALSE}
library(leaflet)
library(sf)
library(dplyr)
farmersn <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  readr::read_csv() %>% group_by(uuid) %>% 
  summarize(x = mean(x), y = mean(y), n = n()) %>% 
  st_as_sf(coords = c("x", "y"), crs = 4326)
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
farmersn <- st_join(farmersn, districts, left = FALSE)
bb <- unname(st_bbox(districts))
xy <- st_centroid(districts) %>% st_coordinates() %>% 
  bind_cols(name = districts$distName, .)
slist <- list("color" = "white")
label_opts <- labelOptions(noHide = TRUE, style = slist, direction = 'center',
                           textOnly = TRUE, textsize = "5px")
qpal <- colorQuantile("RdYlBu", farmersn$n, n = 5)

m <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  fitBounds(bb[1], bb[2], bb[3], bb[4]) %>% 
  addPolygons(data = districts, fill = FALSE, color = "white", 
              group = "Districts", weight = 1) %>% 
  addCircles(data = farmersn, popup = as.character(farmersn$n), 
             color = ~qpal(n), group = "Farmers") %>% 
  addLabelOnlyMarkers(xy$X, xy$Y, label = xy$name, group = "Names",
                      labelOptions = label_opts) %>%
  addLayersControl(overlayGroups = c("Districts", "Names", "Farmers"),
                   options = layersControlOptions(collapsed = FALSE,
                                                  autoZIndex = FALSE))
```
```{r, eval=FALSE, echo=FALSE}
library(htmlwidgets) 
saveWidget(m, "docs/figures/zambialeaflet2.html")
```


---

<iframe seamless src="figures/zambialeaflet2.html" width="100%" 
height="500"></iframe>


---

## Today

### Intro to Spatial 2

- Spatial properties

---

Data:  read in `farmers` and make into `sf`, `districts`, and `roads`
```{r, eval=FALSE, echo = FALSE}
library(geospaar)
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() %>% st_as_sf(coords = c("x", "y"))
districts <- st_read(
  system.file("extdata/districts.shp", package = "geospaar")
)
roads <- read_sf(system.file("extdata/roads.shp", package = "geospaar"))
```

---
### Adding to data

Add some categories
```{r, eval=FALSE}
set.seed(2)
districts2 <- cbind(districts, st_centroid(districts) %>% st_coordinates()) %>% 
  mutate(yield = (7 - 0.25 * -Y) * runif(n = nrow(.), min = 0.9, max = 1.2)) %>%
  mutate(yld_cat = ifelse(yield > 4.5, "high", "other"), 
         yld_cat = ifelse(between(yield, 3.5, 4.5), "medium", yld_cat),
         yld_cat = ifelse(yield < 3.5, "low", yld_cat)) %>% 
  dplyr::select(distName, X, Y, yield, yld_cat)
```
---

## Analyses
```{r, eval=FALSE}
# districts2 %>% dplyr::select(yield) %>% summarize_all(funs(mean, sd))
# districts2 %>% summarize(mean(yield), sd(yield))
districts2 %>% group_by(yld_cat) %>% 
  summarize(yield = mean(yield)) %>% 
  ggplot() + geom_sf(aes(fill = yield))
  # geom_sf(aes(fill = yld_cat))
```

---
### Spatial joins
```{r, eval=FALSE}
p <- ggplot(districts2) + geom_sf(fill = "grey")
# farmer_ct <- farmers %>% group_by(uuid) %>% count()
# st_crs(farmer_ct) <- st_crs(districts)
farmer_ct <- farmers %>% group_by(uuid) %>% count() %>% 
  st_set_crs(st_crs(districts))

farmers_districts <- st_join(farmer_ct, districts2, largest = TRUE)
p + geom_sf(data = farmers_districts, aes(color = n)) + 
  scale_color_viridis_c()

# lapply(farmers_per_dist$distName, function(x) {
#   farmers_per_dist %>% filter(distName == x) %>% st_cast("POINT")
# }) %>% do.call(rbind, .)
# farmers_per_dist %>% st_cast("POINT") %>% plot()
farmers_per_dist <- farmers_districts %>% group_by(distName) %>% count()
p + geom_sf(data = farmers_per_dist) + 
  geom_sf(data = farmers_per_dist %>% st_centroid(), col = "blue")
```

---
### Spatial-non-spatial join
```{r, eval=FALSE}
districts3 <- as_tibble(farmers_per_dist) %>% dplyr::select(distName, n) %>% 
  left_join(districts2, .) %>% mutate(n = ifelse(is.na(n), 0, n))

p + geom_sf(data = districts3, aes(fill = log10(n))) +
  scale_fill_brewer(palette = "RdBu")
  # scale_fill_viridis_c()
# p + geom_sf(data = districts3, aes(fill = n))
```

---
### Subsetting

By count
```{r, eval = FALSE}
p + districts3 %>% filter(n > 15) %>% geom_sf(data = ., fill = "red") + 
  districts3 %>% filter(between(n, 1, 15)) %>% geom_sf(data = ., fill = "blue")
p + geom_sf(data = districts3 %>% filter(n > 15), fill = "red") + 
  geom_sf(data = districts3 %>% filter(between(n, 1, 15)), fill = "blue")
```

By area
```{r, eval = FALSE}
dmed <- districts3 %>% #mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  mutate(area = as.numeric(units::set_units(st_area(.), "km2"))) %>% 
  filter(area == quantile(area, probs = 0.5, type = 1)) 
p2 <- p + geom_sf(data = dmed, fill = "red")

dmax <- districts3 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  # arrange(area) %>% slice(nrow(.)) %>% 
  arrange(desc(area)) %>% slice(1) 
  # slice_max(order_by = area) #
p2 + geom_sf(data = dmax, fill = "blue")

districts3 %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  arrange(area) %>% pull(distName)
```

---
## Various

Dataset
```{r, eval=FALSE}
pols <- st_multipolygon(list(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5),
                                        y = c(-15.5, -16.5, -17, -16, -15.5))),
                             list(cbind(x = c(26.5, 27.5, 27, 26, 26.5) + 1,
                                        y = c(-15.5, -16.5, -17, -16, -15.5) -
                                          1))))
pols <- pols %>% st_geometry %>% st_cast("POLYGON") %>% 
  st_as_sf(crs = 4326) %>%   
  mutate(ID = 1:nrow(.)) %>% #as.data.frame() %>%
  dplyr::select(ID)

p + geom_sf(data = pols, aes(fill = as.factor(ID))) + 
  geom_sf(data = st_union(pols), 
          fill = rgb(1, 0, 1, alpha = 0.7), col = "purple")
```

---
### Intersects
```{r, eval=FALSE}
# which(st_intersects(districts2, st_union(pols), sparse = FALSE))
int_dists <- districts2 %>% 
  slice(which(st_intersects(., st_union(pols), sparse = FALSE)))
p + geom_sf(data = int_dists, fill = "red") + 
  geom_sf(data = pols, fill = "transparent", col = "blue")
```

---
### Intersection
```{r, eval=FALSE}
districts_pol <- st_intersection(districts2, st_union(pols))
ggplot(districts_pol) + geom_sf(aes(fill = distName))
p + geom_sf(data = districts_pol, aes(fill = distName))
```

---
### On your own

- Select the smallest district in Zambia from `districts3`, and convert that
to centroid
```{r, eval = FALSE}
smallest_centroid <- districts3 %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter
```

- Create a buffer of width 0.5 around that
```{r, eval = FALSE}
smallest_buffer <- smallest_centroid %>% st_buf...
```

- Use the buffer to select the districts from `districts3` that intersect with it
```{r, eval = FALSE}
dists_in_buff <- districts3 %>% 
  filter(st_intersects(., smallest_buffer...
```


---
- Use the buffer to do an intersection with `districts3`
```{r, eval=FALSE}
buff_int_dists <- st_intersection(...
# p + geom_sf(data = buff_int_dists, aes(fill = distName))
```


- Calculate the area of the intersection



