---
title: "Geospatial Analysis with R"
subtitle: Class 14
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message=FALSE, eval=FALSE, warning=FALSE}
library(leaflet)
library(sf)
library(dplyr)
farmersn <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  readr::read_csv() %>% group_by(uuid) %>% 
  summarize(x = mean(x), y = mean(y), n = n()) %>% 
  st_as_sf(coords = c("x", "y"), crs = 4326)
districts <- read_sf(system.file("extdata/districts.geojson", 
                                 package = "geospaar"))
farmersn <- st_join(farmersn, districts, left = FALSE)
bb <- unname(st_bbox(districts))
xy <- st_centroid(districts) %>% st_coordinates() %>% 
  bind_cols(name = districts$distName, .)
slist <- list("color" = "white")
label_opts <- labelOptions(noHide = TRUE, style = slist, direction = 'center',
                           textOnly = TRUE, textsize = "5px")
qpal <- colorQuantile("RdYlBu", farmersn$n, n = 5)

m <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  fitBounds(bb[1], bb[2], bb[3], bb[4]) %>% 
  addPolygons(data = districts, fill = FALSE, color = "white", 
              group = "Districts", weight = 1) %>% 
  addCircles(data = farmersn, popup = as.character(farmersn$n), 
             color = ~qpal(n), group = "Farmers") %>% 
  addLabelOnlyMarkers(xy$X, xy$Y, label = xy$name, group = "Names",
                      labelOptions = label_opts) %>%
  addLayersControl(overlayGroups = c("Districts", "Names", "Farmers"),
                   options = layersControlOptions(collapsed = FALSE,
                                                  autoZIndex = FALSE))
```

```{r, eval=FALSE, echo=FALSE}
library(htmlwidgets) 
saveWidget(m, "docs/figures/zambialeaflet2.html")
```

---

<iframe seamless src="figures/zambialeaflet2.html" width="100%" 
height="500"></iframe>

---
## Today
- Finish up:
  - split-apply-combine
  - plotting
  - regression
- Intro to spatial 

---
```{r, message = FALSE}
library(sf)
library(dplyr)
sz <- 100
dat <- lapply(1:3, function(x) {
  set.seed(x)
  d <- data.frame(
    id = 1:sz,
    v1 = runif(sz, min = 2, max = 12),
    grp = paste0("g", x)
  ) %>% mutate(v2 = v1 + rnorm(sz, mean = 2, sd = 2)) %>% 
    select(id, v1, v2, grp)
}) %>% bind_rows()
```

---
```{r}
plot(dat$v1, dat$v2)
```

---

## SAC
### `tidyverse`
```{r, eval=FALSE}
dat %>% 
  group_by(grp) %>% 
  # summarize(v2 = mean(v2)) %>%
  # summarize(across(v1:v2, mean)) %>%
  summarize(across(v1:v2, list("mu" = mean, "sd" = sd)))
```

### base
```{r, eval=FALSE}
t(sapply(unique(dat$grp), function(x) {
  colMeans(dat[dat$grp == x, c("v1", "v2")])
}))
```

---
## Graphical SAC

```{r, eval=FALSE}
ggplot(dat) + 
  geom_point(aes(x = v1, y = v2, color = grp))
```

---
```{r, eval=FALSE}
ggplot(dat) + 
  geom_point(aes(x = v1, y = v2)) + 
  facet_wrap(~grp)
  # facet_wrap(~grp, nrow = 2, ncol = 2)
```

```{r, eval=FALSE}
par(mfrow = c(1, 3))
for(i in unique(dat$grp)) {
  plot(v2 ~ v1, data = dat[dat$grp == i, ])
}
```

---
## Regression

### Base
```{r, eval=FALSE}
reg <- lm(v2 ~ v1, data = dat)
summary(reg)

plot(dat$v1, dat$v2)
abline(reg, col = "red")
```

### `ggplot`

```{r, eval=FALSE}
ggplot(dat) + 
  geom_point(aes(x = v1, y = v2)) + 
  geom_smooth(aes(x = v1, y = v2), method = "lm")
```

---
## SAC + Regression

```{r, eval=FALSE}
# ? thoughts
ggplot(dat) + 
  geom_point(aes(x = v1, y = v2, color = grp)) + 
  geom_smooth(aes(x = v1, y = v2), method = "lm") + 
  # geom_smooth(aes(x = v1, y = v2, color = grp), method = "lm") #+
  facet_wrap(~grp)
```


---
## Homework

- Use base `R` functions to repeat the operation performed in the plot in the preceding slide, i.e. split `dat` by group, apply an `lm` to each split, and plot each split as one of three panels in a plot, with the regression line plotted over the points using `abline`.  
- If you want to go a bit further, figure out how to adorn each plot so that each series gets a different color (as with `ggplot`). 



---
## Spatial data
### Reading

- Create spatial points from csv
```{r, message=FALSE, }
library(tidyverse)
library(sf)
## read in csv
farmers <- system.file("extdata/farmer_spatial.csv", package = "geospaar") %>% 
  read_csv() 

## convert tibble to 'sf'
farmers_sf <- st_as_sf(farmers, coords = c("x", "y"), crs = 4326)
```

---

### Reading
- Reading geojsons
- Both `st_read` and `read_sf` work

```{r, message = FALSE}
districts <- read_sf(system.file("extdata/districts.geojson", 
                                 package = "geospaar"))
roads <- st_read(system.file("extdata/roads.geojson", package = "geospaar"))
```

---
### Writing
- `st_write` works as expected
- Use '.geojson' (only one file)
```{r, eval = FALSE}
#st_write(districts, "districts_out.shp")
st_write(districts, "districts2.geojson")
```

---
## Projections (crs)
- Geographic coordinates are: `4326`
```{r, message = FALSE}
st_crs(farmers_sf) <- 4326
farmers_sf %>% st_crs()
# or
farmers_sf %>% st_set_crs(NA)
farmers_sf %>% st_set_crs(4326)
```


---
## projections (use existing)
- you can set projections based on another layer
- but the two layers MUST have the same crs (e.g. geographic)
```{r, message = FALSE}
farmers <- read_csv(system.file("extdata/farmer_spatial.csv", 
                                package = "geospaar"))
farmers <- st_as_sf(farmers, coords = c("x", "y")) %>% 
  st_set_crs(st_crs(districts))

farmers %>% st_crs() ## print new crs
```

---
## projections (transform)
- Use `st_transform()` to project from one crs to another
```{r, message = FALSE}
roads <- st_transform(x = roads, crs = st_crs(districts))
st_crs(roads)
```


---
## Features from scratch
```{r, message = FALSE}
pt <- st_point(x = c(28, -14))
pt
#> POINT (28 -14)
#
# #2
pts <- st_multipoint(x = cbind(x = c(27.5, 28, 28.5), y = c(-14.5, -15, -15.5)))
pts
#> MULTIPOINT ((27.5 -14.5), (28 -15), (28.5 -15.5))
#
# #3
sline <- st_linestring(cbind(x = c(27, 27.5, 28), y = c(-15, -15.5, -16)))
sline
#> LINESTRING (27 -15, 27.5 -15.5, 28 -16)
#
# #4
pol <- st_polygon(list(cbind(x = c(26.5, 27.5, 27, 26, 26.5), 
                             y = c(-15.5, -16.5, -17, -16, -15.5))))
pol
```

---
### Plotting
- `sf`'s generic `plot()` (use `add = TRUE`)
```{r, message = FALSE}
par(mar = c(0, 0, 0, 0))
plot(districts %>% st_geometry(), col = "grey")
plot(pt, add = TRUE, col = "red", pch = 20)
plot(pts, add = TRUE, col = "blue", pch = 20)
plot(sline, add = TRUE, col = "cyan", lwd = 2)
plot(pol, add = TRUE, col = "orange", lwd = 2)
```


---
## Plotting (ggplot)
- Use `geom_sf()`
```{r, message = FALSE}
ggplot(districts) + geom_sf() +
  geom_sf(data = roads, col = "red") + 
  geom_sf(data = farmers_sf, col = "blue")
```

