---
title: "class5 homework"
author: "Lei Song"
date: "3/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Class 5 challenge
### Introduction

This dataset was build by augmenting datasets of rainfall, climate and fertilizer data available for India ([link](https://www.kaggle.com/atharvaingle/crop-recommendation-dataset)).
We use this dataset to practice how to use `dplyr` to clean, modify, and summarize dataset. The packages or functions that are recommended to use are:

- Package `readr` and its function `read_csv` to read data. Install the package if necessary.
- `group_by` function in `dplyr` to group data and then `summarise` or more flexible `summarise_at` to do something to the groups.
- `select` function in `dplyr`  to subset columns, and `filter` function to subset rows.
- `arrange` function in `dplyr` to order the table based on column(s). `desc` function can be used inside of `arrange` to arrange decreasingly.
- `mutate` function in `dplyr` to add a new column to a dataset. Use with `ifelse` might have some surprising effect.
- `slice` function to index rows.
- `pull` function in `dplyr` to read a column as a vector in pipeline.

Read Unit1-Module4 "Data preparation" and the help documentation of each function to learn how to use them.

### Tasks
Create an R markdown file for these tasks.

#### Task 1

1. Install (if in need) and load necessary packages.
1. Click the link to download the csv file [Crop_recomendation.csv](https://www.kaggle.com/atharvaingle/crop-recommendation-dataset/download) and then read it into working environment as object `crop_rmd` using `readr::read_csv`. Hint: set the right data path. One of the safe way is to use the full path.
2. List all crops growing in this area. Hint: which function we learned to clean duplicated elements?

You are expected to get crop types first and assign to an object `crops`. And then print out them and how many of them using **inline code**. 

*Answer example:*

```{r load_packages, warning=F, message=F}
# Above code chunk setting to ignore messages and warnings.
# install.packages('readr')
library(readr)
library(dplyr)
library(here)
```

```{r}
# Task 1
# Load data

# base function read.csv
crop_rmd <- read.csv(here('docs/homeworks/Crop_recommendation.csv'),
                     stringsAsFactors = F)
# read_csv in readr
crop_rmd <- read_csv(here('docs/homeworks/Crop_recommendation.csv'))
crops <- crop_rmd %>% distinct(label) %>% pull(label) # $
## Or
crops <- crop_rmd %>% pull(label) %>% unique()
```

There are `r length(crops)` crops in this region: `r crops`.

#### Task 2

Let's focus on maize:

1. Just select out maize.
2. Calculate the mean values for N, P, K, temperature, humidity, ph, and rainfall.
3. Directly print out the result.

There might be multiple methods to do this work. 

*Answer example:*

```{r}
# Task 2
crop_rmd %>% 
  filter(label == 'maize') %>% 
  dplyr::select(-label) %>% # remove the non-numeric column
  colMeans() # works for numeric
```

#### Task 3

Have a look at papaya this time:

1. Only select out papaya.
2. Add a new column `n_level` and sign values based on the rule: 

   - "low", N <= 45, 
   - "median", N > 45 and N <= 55
   - "high", N > 55
  
    Hint: use nested `ifelse`.
  
3. Then just select out the rows that `n_level` is "median", and rainfall >= 150, and temperature < 30. Hint: which operator is for logical and?
4. Just select rainfall, temperature and humidity.
5. Random select one row out using seed 10.
6. Assign the result to object `papaya` and print it out. Hint: use function `sample_n`.

*Expected result:*

```{r}
# be careful of space in strings.
set.seed(10)
papaya <- crop_rmd %>% 
  mutate(n_level = ifelse(N <= 45, 'low', 
                          ifelse(N <= 55, 
                                 'median', 'high'))) %>% 
  filter(n_level == 'median' & 
           rainfall >= 150 & 
           temperature < 30) %>% 
  dplyr::select(rainfall, temperature, humidity) %>% 
  sample_n(1) # sample_frac() to sample with percentage
papaya

# Another solution modified based on Sarah's work
set.seed(10)
papaya <- filter(crop_rmd, label == 'papaya') %>%
  mutate(n_level = ifelse(N %in% 0:45, "low", 
                          ifelse(N %in% 46:55, 
                                 "medium", "high"))) %>% 
  filter(n_level == 'medium' & 
           rainfall >= 150 & 
           temperature < 30) %>% 
  select(rainfall, temperature, humidity) %>% 
  sample_n(1)
papaya
```

#### Task 4

Let's get some crop that might tolerate heat and drought:

1. Summarize the table to get the mean, max, and min value of temperature, humidity, ph, and rainfall for each crop type.
2. Arrange the table based on mean rainfall increasingly and max temperature decreasingly.
3. Pick up the first 10 crop types.
4. Assign the result to object `crop_hd` and print it out.

*Expected result:*

```{r}
# Task 4
crop_hd <- crop_rmd %>% 
  group_by(label) %>% # summarise
  summarise_at(.vars = c('temperature', 'humidity', 
                         'ph', 'rainfall'),
               .funs = c(mean = "mean",
                         max = "max",
                         min = "min")) %>% 
  arrange(rainfall_mean, desc(temperature_max)) %>% 
  slice(1:10)
crop_hd

# The arguments can also be set like this:
## Based on Sophie's solution
crop_rmd %>% 
  group_by(label) %>% 
  summarise_at(vars(temperature, humidity, ph, rainfall),
               # funs is deprecated from dplyr8.0.0 and later.
               # but could use list instead.
               list(mean = mean, max = max, min = min)) %>% 
  arrange(rainfall_mean) %>% 
  arrange(desc(temperature_max)) %>% 
  head(10) -> crop_hd # head() is a base R function to select out the first n elements
crop_hd
```
