---
title: "Geospatial Analysis with R"
subtitle: Class 12
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class14plus.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, message=FALSE, eval=FALSE, warning=FALSE}
library(leaflet)
library(sf)
library(dplyr)
districts <- read_sf(system.file("extdata/districts.shp", package = "geospaar"))
bb <- unname(st_bbox(districts))
xy <- st_centroid(districts) %>% st_coordinates() %>% 
  bind_cols(name = districts$distName, .)
slist <- list("color" = "white")
label_opts <- labelOptions(noHide = TRUE, style = slist, direction = 'center',
                           textOnly = TRUE, textsize = "5px")

m <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  fitBounds(bb[1], bb[2], bb[3], bb[4]) %>% 
  addPolygons(data = districts, fill = FALSE, color = "white", 
              group = "Districts", weight = 1) %>% 
  addLabelOnlyMarkers(xy$X, xy$Y, label = xy$name, group = "Names",
                      labelOptions = label_opts) %>%
  addLayersControl(overlayGroups = c("Districts", "Names"),
                   options = layersControlOptions(collapsed = FALSE,
                                                  autoZIndex = FALSE))
```

```{r, eval=FALSE, echo=FALSE}
library(htmlwidgets) 
saveWidget(m, "docs/figures/zambialeaflet.html")
```


---

<iframe seamless src="figures/zambialeaflet.html" width="100%" 
height="500"></iframe>


---

## Today

- More practice with reshaping/analysis/plotting 
- Towards spatial
  - Non-spatial to spatial
  - Read and write spatial
  - CRS
  - Make data from scratch
  - Spatial properties

---

## More ggplot examples
- [Datanovia](https://www.datanovia.com/en/blog/ggplot-examples-best-reference/)
- [R Graph Gallery](https://r-graph-gallery.com/index.html)
- [Legend guide](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software)

---
## Load Whittier data
```{r, out.width = "30%"}
library(dplyr)
folder_path <- "../inst/extdata/"
files <- list.files(folder_path, pattern = "whittier")
for(file in files){
  load(paste0(folder_path, file))
}

up_data <- whittier_up_d_h %>% select(date, 
                           lat, 
                           long,
                           device,
                           ndvi_hourly, 
                           gcvi_hourly, 
                           savi_hourly ,
                           evi2_hourly)

mid_data <- whittier_mid_d_h %>% select(date, 
                           lat, 
                           long,
                           device,
                           ndvi_hourly, 
                           gcvi_hourly, 
                           savi_hourly ,
                           evi2_hourly)

down_data <- whittier_down_d_h %>% select(date, 
                           lat, 
                           long,
                           device,
                           ndvi_hourly, 
                           gcvi_hourly, 
                           savi_hourly ,
                           evi2_hourly)
```

---
## Group data
- Ideally all data should be in a single data.frame/tibble
- Let's add a variable `site_name`, and then combine using `rbind`

```{r, out.width = "30%"}
## add site_name column
up_data$site_name <- "up"
mid_data$site_name <- "mid"
down_data$site_name <- "down"

## combine using rbind

all_sites_data <- do.call('rbind', list(up_data, mid_data, down_data))
all_sites_data$year <- lubridate::year(all_sites_data$date)

```


---
## Legends
- Legends must be based on aesthetic in `aes()`
  - Color
  - fill
  - shape
  - size
  - alpha (transparency)
  
```{r, out.width = "30%"}
library(ggplot2)
ggplot(all_sites_data %>% filter(year == 2020)) +
  geom_line(aes(x = date, y = ndvi_hourly, col = site_name))
```

---
## Legends - THIS WILL WORK
- Correct way:
  - data in one table
  - use `aes()` for column you want to make a legend on
  - use `_manual` function if you want more control
  - don't need legend for `x` and `y` (use axis labels/titles)

```{r, out.width = "30%"}
# set up manual
colors <- c("up" = "orange", 
            "mid" = "slateblue",
            "down" = "#AAAAAA")

ggplot(all_sites_data %>% filter(year == 2020)) +
  geom_line(aes(x = date, y = ndvi_hourly, col = site_name)) +
  scale_color_manual(values = colors)
```

---
## Legends - THIS WON'T work
- Incorrect way:
  - data is in multiple tables
  - colors are "hard-coded" (not using aes())
```{r, out.width = "30%"}
library(lubridate)
ggplot() +
  geom_line(data = up_data %>% 
              filter(date >= as_date("2020-06-01")) %>% 
              filter(date <= as_date("2020-09-30")),
            aes(x = date, y = gcvi_hourly), col = 'red') +
  geom_line(data = mid_data %>% 
              filter(date >= as_date("2020-06-01")) %>% 
              filter(date <= as_date("2020-09-30")),
            aes(x = date, y = gcvi_hourly), col = 'blue') +
  geom_line(data = down_data %>% 
              filter(date >= as_date("2020-06-01")) %>% 
              filter(date <= as_date("2020-09-30")),
            aes(x = date, y = gcvi_hourly), col = 'green')
```


---
## Legends
- If you don't provide a "manual", R will do its best to "guess"
- Here R uses different colors and sizes for sites

```{r, out.width = "30%"}
ggplot(all_sites_data %>% filter(year == 2020)) +
  geom_line(aes(x = date, y = ndvi_hourly, size = site_name, col = site_name))

```

---
## Legend manuals
- Manuals provide great control.
- They take time to setup, kind of like a dictionary.
- `ggplot` [shapes]()

```{r, out.width = "30%"}
colors <- c("up" = "orange", 
            "mid" = "slateblue",
            "down" = "#AAAAAA")
alphas <- c("up" = 1, 
            "mid" = 0.3,
            "down" = 0.1)
shapes <- c("up" = 15, 
            "mid" = 16,
            "down" = 17)
```

---
## Legend manuals

```{r, out.width = "30%"}
ggplot(all_sites_data %>% filter(year == 2020)) +
  geom_point(aes(x = date, 
                y = ndvi_hourly,
                col = site_name,
                alpha = site_name,
                shape = site_name)) +
  scale_color_manual(values = colors) + 
  scale_alpha_manual(values = alphas) +
  scale_shape_manual(values = shapes) +
  theme_bw()
```

---
## manually adjusting legend
- [Datanovia article](https://www.datanovia.com/en/blog/ggplot-legend-title-position-and-labels/)

Before
```{r, out.width = "30%"}
p1 <- ggplot(all_sites_data %>% filter(year == 2020)) +
  geom_point(aes(x = date, 
                y = ndvi_hourly,
                col = site_name)) +
  scale_color_manual(values = colors) + theme_bw() 

p1
```


---
## manually adjusting legend
- [Datanovia article](https://www.datanovia.com/en/blog/ggplot-legend-title-position-and-labels/)

After
```{r, out.width = "30%"}
p1 <- ggplot(all_sites_data %>% filter(year == 2020)) +
  geom_point(aes(x = date, 
                y = ndvi_hourly,
                col = site_name)) +
  scale_color_manual(values = colors) + theme_bw() 

p1 + labs(color = "Pod location") ## changes legend title
```

---
## Continuous color ramp (Exercise)
- First we create a scatter plot of NDVI (x) vs GCVI (y)
- Let's use the diamond shape for the down location.
- [shapes link](http://www.sthda.com/english/wiki/ggplot2-point-shapes)

```{r, eval = F, out.width = "30%"}
shapes <- c("up" = 15, 
            "mid" = 16,
            "down" = ...)
p1 <- ggplot(all_sites_data %>% filter(year = 2020)) +
  geom_point(...) +
  scale_shape_manual(...) +
  theme_bw()
p1
```

---
## Continuous color ramp (Exercise)

- Add color ramp using `scale_color_continuous()`
```{r, out.width = "30%"}
shapes <- c("up" = 15, 
            "mid" = 16,
            "down" = 18)
p1 <- ggplot(all_sites_data %>% 
               filter(year == 2020) ) +
  geom_point(aes(x = ndvi_hourly,
                 y = gcvi_hourly, 
                 shape = site_name,
                 col = date)) +
  scale_shape_manual(values = shapes) +
  scale_color_continuous(type = "viridis") +
  theme_bw()
p1
```

---
## Continuous color ramp (Exercise)

- The date is funky, so we need to create a function to tranform the date label.
- Don't worry about details, this is advanced!

```{r, out.width = "30%"}
Date_F <- function(x){
    as.character(x, origin = '1970-01-01',"%Y-%m-%d")
}

p1 <- ggplot(all_sites_data %>% 
               filter(year == 2020) ) +
  geom_point(aes(x = ndvi_hourly,
                 y = gcvi_hourly, 
                 shape = site_name,
                 col = date)) +
  scale_shape_manual(values = shapes) +
  scale_color_continuous(type = "viridis", labels = Date_F) +
  theme_bw()
p1

```



---
## Many other options (binned color etc)
- [ggplot book](https://ggplot2-book.org/scale-colour.html))
- Don't try to memorize all options!
- Figure out what you want to graph, then find an example

---
## Multiple legends
- Examples from [this guide](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software) 

```{r, out.width = "30%"}
mtcars$cyl<-as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
head(mtcars)
p <- ggplot(data = mtcars, 
    aes(x=mpg, y=wt, color=cyl, size=qsec, shape=gear))+
    geom_point()
# Print the plot without guide specification
p
```

---
## Update legend
- change legend order

```{r, out.width = "30%"}
p+guides(color = guide_legend(order=1),
         size = guide_legend(order=2),
         shape = guide_legend(order=3))
```

---
## Update legend
- remove one legend

```{r, out.width = "30%"}
p+guides(color = FALSE, size = FALSE)
```

---
## Themes
- [ggplot book](https://ggplot2-book.org/polishing.html)
Before
```{r, out.width = "30%"}
base <- ggplot(mpg, aes(cty, hwy, color = factor(cyl))) +
  geom_jitter() + 
  geom_abline(colour = "grey50", size = 2)
base
```

---
With labels
```{r, out.width = "30%"}
labelled <- base +
  labs(
    x = "City mileage/gallon",
    y = "Highway mileage/gallon",
    colour = "Cylinders",
    title = "Highway and city mileage are highly correlated"
  ) +
  scale_colour_brewer(type = "seq", palette = "Spectral")
labelled
```

---

With labels + style
```{r, message = F, out.width = "30%"}
styled <- labelled +
  theme_bw() + 
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.background = element_rect(fill = "white", size = 4, colour = "white"),
    legend.justification = c(0, 1),
    legend.position = c(0, 1),
    axis.ticks = element_line(colour = "grey70", size = 0.2),
    panel.grid.major = element_line(colour = "grey70", size = 0.2),
    panel.grid.minor = element_blank()
  )
styled
```


---
## Exercises
- (1) Create SAVI time-series for 2019 with different color for each site.
- (2) Same as above, use facet_grid() to create a separate plot for each year.
- (3) Create scatter plot of EVI vs SAVI. Use a different shape for each site, and a different color for each year.

---
## Challenge
- Create a scatter plot that compares NDVI at the up site (x) and the mid site (y).
- Use a different color for each year
- Hint: you will need to reshape data



---
## Homework
- Read Unit 2 Module 1, parts 1-3
- Review `sf` [overview](https://r-spatial.github.io/sf/articles/sf1.html)
- ggplot review
  - [Legends](https://www.datanovia.com/en/blog/ggplot-legend-title-position-and-labels/) 
  - [Colors](https://ggplot2-book.org/scale-colour.html)
  - [themes](https://ggplot2-book.org/polishing.html)
- Unit 1 Module 4 assignment due next Tuesday (Feb 28)

