---
title: "Geospatial Analysis with R"
subtitle: Class 19
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Today

- Challenge from last class
- Raster intro, continued
  - rasterize, raster to vector
  - raster stats
  - raster math
  - focal functions
- Exercises

---
### Challenge example

```{r, eval=FALSE}
library(tidyverse)
library(sf)

districts <- read_sf(
  system.file("extdata/districts.shp", package = "geospaar")
) %>% st_transform(4326)
smallest_district <- districts %>% 
  mutate(area = as.numeric(st_area(.)) / 10^6) %>% 
  filter(area == min(area))
smallest_centroid <- smallest_district %>% st_centroid()
xy <- st_coordinates(smallest_centroid)

minus <- function(a, b) a - b
rect_pol <- function(width, hgt, xy, crs = 4326) {
  w <- sapply(c(minus, minus, sum, sum, minus), function(x) x(xy[1], width / 2))
  h <- sapply(c(minus, sum, sum, minus, minus), function(x) x(xy[2], hgt / 2))
  l <- list(cbind(w, h))
  p <- st_polygon(l) %>% st_sfc %>% st_sf(ID = 1:length(l), crs = crs)
  return(p)
}  
pol <- rect_pol(0.5, 0.25, xy)  

ggplot(districts) + geom_sf() +
  geom_sf(data = smallest_district, fill = "red") +
  geom_sf(data = pol, col = "blue", fill = "transparent")
```

---
## More on rasters

- create
```{r, eval = FALSE}
library(raster)
r <- raster(extent(30, 31, -14, -13), res = 0.1, crs = 4326)
values(r) <- sample(1:10, size = ncell(r), replace = TRUE)

s <- stack(r, log10(r))
names(s) <- c("dummy", "log10dummy")
plot(s)

b2 <- lapply(1:10, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.1, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  r
}) %>% stack %>% brick
```

---
```{r, eval=FALSE}
b3 <- lapply(1:10, function(x) {
  r <- raster(extent(30, 31, -14, -13), res = 0.001, 
              crs = "+proj=longlat +datum=WGS84")
  set.seed(x)
  values(r) <- sample(1:10, size = ncell(r), replace = TRUE)
  r
}) %>% stack %>% brick

data(chirps)
chirpsz <- mask(chirps, districts)
chirpsz %>% projectRaster(., crs = "ESRI:102022", res = 5000) -> chirpsz_alb
```

---
- Rasterize/vectorize
```{r, eval = FALSE}
distsr <- districts %>% rasterize(., chirpsz)
distsr %>% plot_noaxes
distsr_pol %>% st_as_sf %>% st_geometry %>% plot
distsr_pol %>% st_as_sf %>% slice(49) %>% plot(add = TRUE)
```

---
## Calculations

```{r, eval=FALSE}
rf1 <- chirpsz[[1]] + chirpsz[[2]]
plot(rf1)
plot(rf1 / 1000)
```

- statistics
```{r, eval=FALSE}
cellStats(chirpsz[[1]], stat = mean)
cellStats(chirpsz[[1]], stat = range)
cellStats(chirpsz, stat = mean)
cellStats(chirpsz, stat = sum)
cellStats(chirpsz, stat = range)
```

- z dimension stats
```{r, eval=FALSE}
plot(calc(chirpsz, mean))
plot(calc(chirpsz, sum))
plot(calc(chirpsz, range))
```
---

- neighborhoods

```{r, eval=FALSE}
w <- matrix(1, nrow = 5, ncol = 5)
rainwindow <- focal(raintot, w = w, fun = mean, na.rm = TRUE)
```

---

- reclassify

```{r, eval=FALSE}
raincat <- cut(calc(chirpsz, sum), breaks = c(0, 50, 100, 200))
cols <- c("grey70", "lightblue", "lightgreen")
plot_noaxes(raincat, col = cols,
            legend = FALSE,  main = "Total Rainfall",
            mar = c(0, 0, 1, 0))
legend(x = "bottomright", legend = c("high", "medium", "low"),
       pch = 15, pt.cex = 3, col = rev(cols), bty = "n")
```

```{r, eval=FALSE}
raintot <- calc(chirpsz, fun = sum)
rain <- cellStats(raintot, range)
rain_3 <- cbind(c(floor(rain[1]), 50, 100),
                 c(50, 100, ceiling(rain[2])), 1:3)
raincat <- reclassify(raintot, rain_3, include.lowest = TRUE)
```




---
## Exercises

- Create a dummy raster (sampling from 1:100) using district 49 for extent and res of 0.1
- Reproject to Albers
- Disaggregate to 0.02 degrees (bilinear and NGB)
- Calculate the sum of rainfall from `chirps`
- Identify all areas of total rainfall > 10 mm
- Calculate the mean rainfall in Zambia for the 15th day in `chirps`



