---
title: "Geospatial Analysis with R"
subtitle: rgee setup
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## rgee setup
- Use Google Earth Engine in R!
- All code thanks to Ricardo Dalagnol [github](https://github.com/ricds/DL_RS_GEE/tree/7a3e70758e81b3bc6ef86cc67a231ed4a8827809)
- Shout-out to André de Oliveira Domingues for getting this to work!!

---

# GEE account 

- you need a GEE account
- log in the https://code.earthengine.google.com/ and register for one





---
# Install Miniconda

- the conda environment is where the GEE Python API will be located. The RGEE package uses it.
- first you need to install the Miniconda OUTSIDE of R
- install Miniconda3 at https://docs.conda.io/en/latest/miniconda.html
- open 'anaconda' in the command prompt (window button --> anaconda, you will see anaconda prompt)
- then type in the commands below one-by-one (without the #) to install the rgee_py environment and packages:
```{bash, eval = F}
# run these one line at a time in Miniconda
conda create -n rgee_py python=3.9
activate rgee_py
pip install google-api-python-client
pip install earthengine-api
pip install numpy
pip install gcloud
```


---
# Get conda environment and copy to R variable

- ok conda should now be installed, now lets get the path to the environment, type inside miniconda:
```{bash, eval = F}
conda env list
```

- copy the path to the rgee_py environment, you will need it set in the variable below inside R:
- note the use of double backslashes \\
- this below is where is located in MY computer, you have to use the 'conda env list' command to find where it is located on yours
```{r, eval = F}
rgee_environment_dir = "C:\\Users\\micha\\miniconda3\\envs\\rgee_py"
```

---
# Install R packages (rgee)
```{r, eval = F}
## if you installed everything above, you can now install the packages inside R
#install.packages('pacman')
library(pacman)
# install/load general packages used in the scripts
p_load(raster,
       rgdal,
       rgeos,
       gdalUtils,
       sp,
       sf,
       leaflet,
       mapview,
       caret)
#install.packages('BiocManager')

# now some more specific packages related to using the rgee
p_load(rgee, geojsonio, remotes, reticulate, devtools, googledrive)
install_github("r-spatial/rgee")
```

---
# Set Python environment in R
- sometimes at this point you are required to restart R or the computer before proceeding
- try restarting if the installation do not finish properly and run the installation again after restart

```{r, eval = F}
# set python
reticulate::use_python(rgee_environment_dir, required=T)
rgee::ee_install_set_pyenv(
  py_path = rgee_environment_dir, # Change it for your own Python PATH
  py_env = "rgee_py" # Change it for your own Python ENV
)
Sys.setenv(RETICULATE_PYTHON = rgee_environment_dir)
Sys.setenv(EARTHENGINE_PYTHON = rgee_environment_dir)
```


- Set Global Python Env under Tools --> Global options --> Python --> Select --> Find Conda Env

---
# Run the following

- You may need to run in R (once)
```{r, eval = F}
rgee::ee_install_upgrade()
```

- You also may need to run (in Miniconda, with env activated)
```{bash, eval = F}
conda install earthengine-api==0.1.323 

# run below and follow instructions
earthengine authenticate --quiet
```


---
# Launch rgee

```{r, eval = F}
rgee::ee_clean_credentials()
rgee::ee_Initialize('michaeljcecil@gmail.com',
                    drive = T)
```



