---
title: "Geospatial Analysis with R"
subtitle: rgee setup
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## rgee setup
- Use Google Earth Engine in R!
- All code thanks to Ricardo Dalagnol [github](https://github.com/ricds/DL_RS_GEE/tree/7a3e70758e81b3bc6ef86cc67a231ed4a8827809)
- Shout-out to André de Oliveira Domingues for getting this to work!!

---
## GEE links
- [code editor](https://code.earthengine.google.com/)
- [Tutorials](https://developers.google.com/earth-engine/tutorials/tutorials)

---
## rgee
- [Examples](https://csaybar.github.io/rgee-examples/)
- [Best Practices](https://r-spatial.github.io/rgee/articles/rgee03.html)
- [Video tutorial](https://www.youtube.com/watch?v=SHXuIpjU3YE)
- [rgee with Shiny](https://appsilon.com/r-shiny-google-earth-engine/)

---
## Start rgee
```{r}
library(rgee)
library(raster)
ee_Initialize()
```

---
## Support Vector Machine classification
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Input imagery is a cloud-free Landsat 8 composite.
l8 = ee$ImageCollection("LANDSAT/LC08/C01/T1")

image = ee$Algorithms$Landsat$simpleComposite(
  collection = l8$filterDate("2018-01-01", "2018-12-31"),
  asFloat = TRUE
)

# Use these bands for prediction.
bands = c("B2", "B3", "B4", "B5", "B6", "B7", "B10", "B11")

# Manually created polygons.
forest1 = ee$Geometry$Rectangle(-63.0187, -9.3958, -62.9793, -9.3443)
forest2 = ee$Geometry$Rectangle(-62.8145, -9.206, -62.7688, -9.1735)
nonForest1 = ee$Geometry$Rectangle(-62.8161, -9.5001, -62.7921, -9.4486)
nonForest2 = ee$Geometry$Rectangle(-62.6788, -9.044, -62.6459, -8.9986)

# Make a FeatureCollection from the hand-made geometries.
polygons = ee$FeatureCollection(c(
  ee$Feature(nonForest1, list(class = 0)),
  ee$Feature(nonForest2, list(class = 0)),
  ee$Feature(forest1, list(class = 1)),
  ee$Feature(forest2, list(class = 1))
))

# Get the values for all pixels in each polygon in the training.
training = image$sampleRegions(
  # Get the sample from the polygons FeatureCollection.
  collection = polygons,
  # Keep this list of properties from the polygons.
  properties = list("class"),
  # Set the scale to get Landsat pixels in the polygons.
  scale = 30
)

# Create an SVM classifier with custom parameters.
classifier = ee$Classifier$libsvm(
  kernelType = "RBF",
  gamma = 0.5,
  cost = 10
)

# Train the classifier.
trained = classifier$train(training, "class", bands)

# Classify the image.
classified = image$classify(trained)

# Display the classification result and the input image.
geoviz_image = list(bands = c("B4", "B3", "B2"), max = 0.5, gamma = 2)
geoviz_class = list(min = 0, max = 1, palette = c("red", "green"))

Map$setCenter(-62.836, -9.2399, 9)
Map$addLayer(
  eeObject = image,
  visParams = geoviz_image,
  name = "image"
) +
Map$addLayer(
  eeObject = classified,
  visParams = geoviz_class,
  name = "deforestation"
) +
Map$addLayer(
  eeObject = polygons,
  name = "training polygons"
)
```

---
## Compositing and Mosaicking

```{r, echo = F, warning = F, message = F}
# Load three NAIP quarter quads in the same location, different times.
naip2004_2012 <- ee$ImageCollection("USDA/NAIP/DOQQ")$
  filterBounds(ee$Geometry$Point(-71.08841, 42.39823))$
  filterDate("2004-07-01", "2012-12-31")$
  select(c("R", "G", "B"))

# Temporally composite the images with a maximum value function.
composite <- naip2004_2012$max()
Map$setCenter(lon = -71.12532, lat = 42.3712)
Map$setZoom(zoom = 12)

Map$addLayer(
  eeObject = composite,
  visParams = list(),
  name = "max value composite"
)

# Load four 2012 NAIP quarter quads, different locations.
naip2012 <- ee$ImageCollection("USDA/NAIP/DOQQ")$
  filterBounds(ee$Geometry$Rectangle(-71.17965, 42.35125, -71.08824, 42.40584))$
  filterDate("2012-01-01", "2012-12-31")

# Spatially mosaic the images in the collection and display.
mosaic <- naip2012$mosaic()
Map$addLayer(
  eeObject = mosaic,
  visParams = list(),
  name = "spatial mosaic"
)

# Load a NAIP quarter quad, display.
naip <- ee$Image("USDA/NAIP/DOQQ/m_4207148_nw_19_1_20120710")
Map$setCenter(lon = -71.0915, lat = 42.3443)
Map$setZoom(zoom = 14)

Map$addLayer(
  eeObject = naip,
  visParams = list(),
  name = "NAIP DOQQ"
)

# Create the NDVI and NDWI spectral indices.
ndvi <- naip$normalizedDifference(c("N", "R"))
ndwi <- naip$normalizedDifference(c("G", "N"))

# Create some binary images from thresholds on the indices.
# This threshold is designed to detect bare land.
bare1 <- ndvi$lt(0.2)$And(ndwi$lt(0.3))
# This detects bare land with lower sensitivity. It also detects shadows.
bare2 <- ndvi$lt(0.2)$And(ndwi$lt(0.8))

# Define visualization parameters for the spectral indices.
ndviViz <- list(min = -1, max = 1, palette = c("FF0000", "00FF00"))
ndwiViz <- list(min = 0.5, max = 1, palette = c("00FFFF", "0000FF"))

# Mask and mosaic visualization images.  The last layer is on top.
mosaic <- ee$ImageCollection(list(
  # NDWI > 0.5 is water.  Visualize it with a blue palette.
  ndwi$updateMask(ndwi$gte(0.5))$visualize(ndwiViz),
  # NDVI > 0.2 is vegetation.  Visualize it with a green palette.
  ndvi$updateMask(ndvi$gte(0.2))$visualize(ndviViz),
  # Visualize bare areas with shadow (bare2 but not bare1) as gray.
  bare2$updateMask(bare2$And(bare1$Not()))$visualize(list(palette = c("AAAAAA"))),
  # Visualize the other bare areas as white.
  bare1$updateMask(bare1)$visualize(list(palette = c("FFFFFF")))
))$mosaic()

Map$addLayer(
  eeObject = ndwi,
  visParams = list(),
  name = "Visualization mosaic"
)
```

---
## Download Images for Zambia

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Download five Landsat-8 more cloudless at Chilanga, Zambia
roi <- ee$Geometry$Point(28.24915,	-15.54929)$buffer(1000)$bounds()

# ee_help(ee$ImageCollection$sort)
dataset <- ee$ImageCollection('LANDSAT/LC08/C01/T1_SR') %>%
  ee$ImageCollection$filterBounds(roi) %>% 
  ee$ImageCollection$filterDate("2020-11-01", "2021-05-31") %>% 
  ee$ImageCollection$sort("CLOUD_COVER")

filter <- ee$Filter$lessThanOrEquals("CLOUD_COVER", 50)
dataset <- dataset$filter(filter)

# Function to clip images
clip_roi <- function(image) {
  image$clip(roi)
}

dataset = dataset$map(clip_roi)

## visualization parameters
# Define the visualization parameters.
vizParams_tcc <- list(
  bands = c("B4", "B3", "B2"),
  min = 0,
  max = 3000)

# View spatial objects interactively
Map$centerObject(dataset$first())
Map$addLayers(dataset, vizParams_tcc, 'TCC') ## adds all layers to Leafle
```

---
## Download step
```{r, eval = F}
## Make sure bands have the same resolution
## By default downloads to Google Drive

# From Earth Engine to Local
ee_imagecollection_to_local(
  ic = dataset$select(c('B5', 'B4', 'B3', 'B2')),
  region = roi,
  dsn = "l8_",
  scale = 30
)

```

