---
title: "Geospatial Analysis with R"
subtitle: Ground Sensor Random Forest demo
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "lucy", "middlebury-fonts", "themes/class18.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
## Random Forests
- Used for regression (continuous prediction) or classification (categorical prediction)
- Good for non-linear modeling.
- Does NOT take into account spatial location. Each pixel's values are treated separately.(although [spatial version](https://link.springer.com/article/10.1007/s11004-021-09946-w) exists)

---
## Example 2
- Predict ground sensor NDVI from satellite time-series
- predict ground-based `ndvi_fitted` from Sentinel-1, 2 data
```{r, message = F, warning = F}
load("../inst/extdata/filt_data.rda")
indep_var <- c("s2_ndvi_fitted", "s2_evi2_fitted", "s2_gcvi_fitted",
               "s1_VH_fitted", "s1_VV_fitted", "s1_vh_vv_fitted")
dep_var <- "ndvi_fitted"
```

---
## Create random forest model
- We will validate on Zambia 2020-21 sites.
- So we remove these sites from model training.
```{r, message = F, warning = F}
library(dplyr)
train_data <- filt_data %>% filter(group_id != "Zam_20_21") %>% select(all_of(c(indep_var, dep_var)) )
head(train_data)
```

---
## create model
- Use `ranger` package (much faster than `randomForest`)
- Dep variable - `ndvi_fitted`
- Indep variables - all other variables in `train_data`
```{r, message = F, warning = F}
library(ranger)
rf_model <- ranger(ndvi_fitted ~ ., 
                   data=train_data, 
                   importance='impurity', mtry=3)
print(rf_model)

# 
# library(randomForest)
# rf_model <- randomForest(ndvi_fitted ~ .,
#                      data = train_data, 
#                      proximity=TRUE,
#                      importance = T) 
```

---
## Variable importance

```{r, message = F, warning = F}
library(vip)
v1 <- vip(rf_model)
v1
```


---
## Predict for validation data and plot
```{r, message = F, warning = F}
library(ggplot2)
valid_data <- filt_data %>% filter(group_id == "Zam_20_21")
ndvi_hat <- predict(rf_model, data = valid_data )

valid_data$ndvi_hat <- ndvi_hat[["predictions"]]
sites <- unique(valid_data$Location)
#par(mfrow = c(7,2))
plots <- lapply(sites, function(site){
  p1 <- ggplot(valid_data %>% filter(Location == site)) + 
    geom_point(aes(x = date, y = ndvi_fitted), col = "blue", alpha = 0.2) + 
    geom_point(aes(x = date, y = ndvi_hat), col = "red") + 
    geom_line(aes(x = date, y = s2_ndvi_fitted), alpha = 0.5, col = 'grey') + 
    geom_point(aes(x = date, y = s2_ndvi), alpha = 0.5, col = 'grey') + 
    ggtitle(site) +
  theme_bw()
  print(p1)
  return(p1)
})
  
```

---
## Plots
- red: predicted ndvi
- blue: pod ndvi (dep var)
- grey: S2 ndvi (one of indep var)
```{r, figures-side, fig.show="hold", out.width="50%", message = F, warning = F}
#par(mfrow= c(2,2))
grid.arrange(plots[[1]],
             plots[[2]],
             plots[[3]],
             plots[[4]],
  nrow = 2)
# for (k in 1:4){
#   print(plots[[k]])
# }
```

