---
title: "Unit 2 Practice answers"
subtitle: "GEOG246-346"
date: "`r Sys.Date()` [Back to home](https://agroimpacts.github.io/geospaar/)"
output:
  rmdformats::readthedown:
    highlight: pygments
    fig_caption: yes
    number_sections: yes
    toc_depth: 4
    css: unit.css
vignette: >
  %\VignetteIndexEntry{Unit 2 Practice answers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Module 1 practice answers{#module-1-practice-answers}
## Practice 1
### Questions
1. Answers [here](https://r-spatial.github.io/sf/articles/sf1.html#simple-feature-geometry-types).

2. Assuming the `tibble` has x and y or lat/long coordinates, you apply the function `st_as_sf` with the "coords" argument set to specify which columns contain the x and y coordinates. 

3. `sf::plot` by default plots one panel per variable. You can create a single panel by specifying the variable you want, or by using the `st_geometry` argument to strip out the geometry from object. It also prevent distortions that sometimes occur when overlaying subsequent features on top of a base map. 

4. For a single point you provide an x and y coordinate, otherwise you give an input matrix containing x and y coordinates. A polygon requires that the last point pair in the matrix is the same as the first point pair, to close the polygon. 

### Code
1. 
```{r, eval=FALSE}
farmers %>% st_geometry()
st_geometry(farmers)
```

2. 
```{r, eval=FALSE}
st_crs(farmers) <- st_crs(districts)
# p <- "path/to/your/project/notebooks/data"
p <- "~/Desktop"
st_write(farmers, dsn = file.path(p, "farmers.sqlite"))
rm(farmers)
st_read(file.path(p, "farmers.sqlite"))
```

3. 
```{r, eval = FALSE}
class(roads)
str(roads)
class(districts)
str(districts)
```

4. 
```{r, eval=FALSE}
plot(roads %>% st_geometry(), col = "blue")
```

5. 
```{r, eval=FALSE}
plot(districts %>% select(distName), main = "Zambia Districts")
```

6. 
```{r, eval=FALSE}
pts <- st_multipoint(x = cbind(x = c(27, 28, 29), y = c(-13, -14, -15)))
plot(districts %>% st_geometry(), col = "grey")
plot(pts, add = TRUE, col = "orange", pch = 16)
```

## Practice 2
### Questions
1. At least with this example, pretty negligible--well less than 1% mean absolute error.  It might matter more in other places and with other scales though.  The reason `st_area` knows how to estimate areas is because it invokes `lwgeom::st_geod_area`, which calculates a geodetic surface area.  

2. Because for the time being `sf::plot` is a fair bit faster. However, [this recent twitter thread](https://twitter.com/thomasp85/status/1092412825899810817) suggests that may change soon. 

3. By using `mutate` with `cut` that has break values based on those properties. In our example, we found the breaks using `quantile` and different probabilities/percentile levels that creating tertiles of area. 

### Code
1. 
```{r, eval = FALSE}
set.seed(1)
districts %>% sample_n(20) %>% st_area() %>% units::set_units("ha") %>% mean()
```

2. 
```{r, eval = FALSE}
set.seed(1)
roads %>% sample_n(100) %>% st_length() %>% units::set_units("km") %>% mean()
```

3. 
```{r, eval = FALSE}
plot(st_geometry(districts), col = "lightgrey")
set.seed(1)
farmers %>% filter(season == 2) %>% sample_n(200) %>% st_geometry() %>% 
  plot(col = "red", pch = 20, add = TRUE)
```

4. 
```{r, eval = FALSE}
districts %>% st_transform(st_crs(roads)) %>% st_geometry() %>% 
  plot(col = "lightgrey")
roads %>%  
  mutate(length = as.numeric(st_length(.) / 1000)) %>% 
  filter(length > 50 & length < 100) %>% st_geometry() %>% 
  plot(col = "red", pch = 20, add = TRUE)
```

5.
```{r, eval = FALSE}
deciles <- function(x) quantile(x, probs = seq(0, 1, 0.1))
dist_deciles <- districts %>% mutate(area = as.numeric(st_area(.)) / 10^6) %>%
  mutate(acls = cut(area, breaks = deciles(area), include.lowest = TRUE)) %>% 
  group_by(acls) %>% summarize(sum_area = sum(area))  
dist_deciles
#
# #2
cols <- heat.colors(10)
par(mar = rep(0, 4))
plot(st_geometry(dist_deciles), col = cols)
legend(x = "bottomright", pch = 15, col = cols, bty = "n", 
       legend = paste0(1:10, c("st", "nd", "rd", rep("th", 7))))

```

### Practice 3
#### Questions
1. It changes features from one type to another (e.g. POLYGON to MULTIPOLYGON), either one specified by the user or the simplest possible common feature, if left unspecified. Casting is sometimes necessary to avoid mixed feature types that cause failures for subsequent operations. 

2. `st_union` runs under the hood of `summarise.sf`, so a `summarize` operation on an `sf` will result in a merged/dissolved set of spatial features. 

3. It affects the order of the fields in the resulting unioned `sf` object--the fields from the object passed to the "x" argument appear first. 

#### Code
1. 
```{r, eval = FALSE}
coords <- cbind("x" = c(27, 27.5, 27.5, 27, 27), 
                "y" = c(-13, -13, -13.5, -13.5, -13))
pol2 <- st_polygon(x = list(coords)) %>% st_sfc %>% st_sf(ID = 1, crs = 4326)

par(mar = rep(0, 4))
plot(st_geometry(districts), col = "grey")
plot(pol2, col = "blue", add = TRUE)
```

2. 
```{r, eval = FALSE}
pol2_int_dists <- st_intersection(pol2, districts)
districts[1] %>% plot(col = "grey", main = NULL)
pol2_int_dists[2] %>% plot(col = rainbow(nrow(pol2_int_dists)), add = TRUE)
```

3. 

`pol2` is nearly 26 hectares larger
```{r, eval = FALSE}
pol2_int_dists %>% st_area() %>% as.numeric() %>% sum() / 10000 -
  pol2 %>% st_area() %>% as.numeric() %>% sum() / 10000
```

4. 
```{r, eval = FALSE}
st_difference(districts, pol2)[2] %>% plot(col = "grey", main = NULL)
```

5. 
```{r, eval = FALSE}
# Try compare the
set.seed(1)
farmers_alb %>% filter(season == 2) %>%
  sample_n(size = 5) %>% st_buffer(dist = 30000) %>% st_geometry %>% plot()

# With 
set.seed(1)
farmers_alb %>% filter(season == 2) %>%
  st_sample(size = 5) %>% st_buffer(dist = 30000) %>% st_geometry %>% plot()
```

6.
```{r, eval = FALSE}
roads_gt400 <- roads %>% filter(as.numeric(st_length(.)) / 1000 > 400)
par(mar = rep(0, 4))
st_transform(districts, st_crs(roads)) %>% st_geometry %>% 
  plot(col = "grey", main = NULL)
roads_gt400 %>% st_buffer(25000) %>% plot(col = "green", add = TRUE)
```

7.
```{r, eval = FALSE}
par(mar = rep(0, 4))
st_transform(districts, st_crs(roads)) %>% st_geometry %>% 
  plot(col = "grey", main = NULL)
roads_gt400 %>% st_buffer(25000) %>% plot(col = "tan", add = TRUE)
set.seed(1)
roads_gt400 %>% st_buffer(25000) %>% 
  st_sample(size = rep(20, nrow(.)), exact = TRUE) %>% 
  plot(col = "red", pch = 20, add = TRUE)

```

# Module 2 practice answers
## Practice 1
### Questions
1. `raster` uses the S4 object-oriented system, where `sf` uses S4. Slots in S4 objects are accessed using the `@` operator. 

2. `brick`, because `raster` only allows you to read and write a single layer.  

3. `stack` organizes multiple rasters that might be stored in separation locations into a single multi-layer object, whereas `brick` requires a single file on disk. `bricks` take longer to create, but are faster to work with once they exist. 

4. The output of `raster`'s vectorization functions are `sp` objects, so you have to convert them to `sf` objects using `st_as_sf`. 

### Code
1. 
```{r, eval=FALSE}
# recreate r, r2, r3
e <- extent(c("xmin" = 27, "xmax" = 29, "ymin" = -16, "ymax" = -14))  
r <- raster(x = e, res = 0.25, crs = crs(districts))
set.seed(1)  
values(r) <- sample(1:100, size = ncell(r), replace = TRUE)  # 3
r2 <- r > 50
r3 <- r
set.seed(1)
values(r3) <- rnorm(n = ncell(r3), mean = 10, sd = 2)

# 
r4 <- r3
set.seed(1)
values(r4) <- runif(n = ncell(r4), 0, 1)
r5 <- r4 > 0.5

s2 <- list(r, r2, r3, r4, r5) %>% stack
names(s2) <- c("r", "r2", "r3", "r4", "r5")
plot(s2)

```

2. We use `tempdir()` here, but you should use your `notebooks/data` folder.
```{r, eval = FALSE}
b <- brick(s2, file = file.path(tempdir(), "b2.tif"))
```

3. 
```{r, eval = FALSE}
zamr3 <- raster(x = extent(districts), crs = crs(districts), res = 0.2)
values(zamr3) <- 1:ncell(zamr3)

farmersr <- farmers %>% distinct(uuid, .keep_all = TRUE) %>% select(x, y) %>% 
  mutate(count = 1) %>% st_as_sf(coords = c("x", "y")) %>% 
  rasterize(x = ., y = zamr3, field = "count", fun = sum)

par(mar = c(0, 0, 1, 4))
districts %>% st_union %>% plot(col = "grey", border = "grey")
plot(farmersr, add = TRUE)
```

4. 
```{r, eval = FALSE}
zamr_alb <- projectRaster(from = zamr, res = 20000, crs = crs(roads), 
                          method = "ngb")
farmersr_alb <- projectRaster(from = farmersr, to = zamr_alb, 
                              method = "bilinear")

par(mar = c(0, 0, 1, 4), mfrow = c(1, 2))
districts %>% st_union %>% plot(col = "grey", border = "grey")
plot(farmersr, add = TRUE)
districts %>% st_transform(crs(roads)) %>% st_union %>% 
  plot(col = "grey", border = "grey")
plot(farmersr_alb, add = TRUE)
```

5. 
```{r, eval = FALSE}
par(mar = c(0, 0, 0, 0))
farmersr %>% rasterToPolygons(dissolve = TRUE) %>% st_as_sf %>% 
  plot(main = NULL)
```

## Practice 2
### Questions
1. `cellStats` provides summary statistics over an entire layer; `zonal` calculates statistics within pre-defined zones; `focal` calculates statistics within a moving window. 

2. Use `method = bilinear`.

3. You don't. You have to resample them to a common resolution and extent first. And then you have to `stack` them. 

### Code
1. 
```{r, eval = FALSE}
as.Date("10-11-2017", "%m-%d-%Y")
as.Date("10-11-17", "%m-%d-%y")
as.Date("101117", "%m%d%y")
as.Date("10112017", "%m%d%Y")
lubridate::mdy("10-11-2017")
lubridate::as_date("20171011")
```

2. 
```{r, eval = FALSE}
farmersr2 <- farmers %>% distinct(uuid, .keep_all = TRUE) %>% 
  mutate(count = 1) %>% select(x, y, count) %>% 
  st_as_sf(coords = c("x", "y")) %>% 
  rasterize(., distsr, field = "count", fun = sum)
```

3. 
```{r, eval = FALSE}
zonal(farmersr2, distsr, fun = sum) %>% data.frame %>% 
  subs(distsr, .) %>% plot_noaxes
```

4.
```{r, eval=FALSE}
wmat3 <- matrix(1, nrow = 3, ncol = 3) 
wmat5 <- matrix(1, nrow = 5, ncol = 5) 
fstack <- list(sd3 = focal(x = chirpsz[[20]], w = wmat3, fun = sd), 
               sd5 = focal(x = chirpsz[[20]], w = wmat5, fun = sd),
               max3 = focal(x = chirpsz[[20]], w = wmat3, fun = max), 
               max5 = focal(x = chirpsz[[20]], w = wmat5, fun = max)) %>% stack
fstack %>% plot_noaxes
```

5. 
```{r, eval = FALSE}
chirps_d57 <- chirpsz[[1]] %>% crop(., extent(districts[57, ]))
s <- list(d1 = disaggregate(chirps_d57, fact = 5), 
          d2 = disaggregate(chirps_d57, fact = 5, method = "bilinear")) %>% 
  stack
plot_noaxes(s)
```

6. 
```{r, eval = FALSE}
s <- lapply(list(mean, cv, median), function(x) {
  calc(chirpsz, fun = x)  
}) %>% stack
names(s) <- c("Mean", "CV", "Median")
plot_noaxes(s, nr = 1)
```

## Practice 3
### Questions
1. Using `cut` with vector of breakpoints (e.g. quantiles), or use `reclassify` with a reclassification matrix.

2. `sampleRandom`, `sampleStratified`. There is also `sampleRegular`. 

### Code
1. 
```{r, eval = FALSE}
chirps_sd <- calc(chirpsz, fun = sd)
```

2. 
```{r, eval = FALSE}
(chirps_sd < cellStats(chirps_sd, mean)) %>% plot
```

3. 
```{r, eval = FALSE}
quantile(raintot, probs = seq(0, 1, 0.2)) %>% cut(raintot, .) %>% 
  plot_noaxes
```

4. 
```{r, eval = FALSE}
set.seed(11)
randdistsr <- districts %>% sample_n(size = 15) %>% rasterize(., raintot)
plot_noaxes(randdistsr)

newrandrain <- mask(raintot, randdistsr)
newrandrain %>% plot_noaxes
```

5. 
```{r, eval = FALSE}
set.seed(1)
randsamp <- sampleRandom(x = newrandrain, size = 300)

set.seed(1)
ind <- sampleStratified(x = randdistsr, size = 300 / 15, cells = TRUE)
stratsamp <- newrandrain[ind[, 1]]

rand_rain_stats <- bind_rows(
  tibble(rain = randsamp, dat = "Simple"),
  tibble(rain = stratsamp, dat = "Stratified"),
) %>% drop_na

bp_theme <- theme(legend.title = element_blank(), axis.text.x = element_blank(),
                  axis.ticks.x = element_blank(), 
                  panel.grid.major.x = element_blank(), 
                  panel.grid.minor.x = element_blank(), 
                  panel.background = element_rect(fill = "grey95"))

rand_rain_stats %>% ggplot() +
  geom_boxplot(mapping = aes(y = rain, fill = dat), position = "dodge2") +
  scale_fill_manual(values = c("lightblue", "steelblue")) + 
  ggtitle("Rainfall distributions") + xlab(NULL) + ylab("mm") + bp_theme

```

## Practice 4
### Questions
1. The `area` function is your friend for this. 

2. Use `expression` together combined with `paste` as needed for more complex labels. 

3. Make `names(predstack)` matches the predictor names used by the model. 

### Code
1. 
```{r, eval = FALSE}
demalb42 <- districts %>% filter(ID == 42) %>% st_transform(st_crs(roads)) %>% 
  crop(demalb, .)
vars <- c("slope", "aspect", "flowdir", "tri")
terrvars <- stack(lapply(1:length(vars), function(x) {
  tv <- terrain(x = demalb42, opt = vars[x], unit = "degrees")
}))
names(terrvars) <- vars

plot_noaxes(terrvars)

```

2. 
```{r, eval = FALSE}
library(gstat)

# #1
raintotalb <- projectRaster(from = raintot, res = 5000, crs = crs(roads))
names(raintotalb) <- "rain"
r <- raster(extent(raintotalb), res = res(raintotalb), crs = crs(raintotalb),             vals = 1)

# lapply approach to interpolation
idw_list <- lapply(c(250, 500, 1000), function(x) {
  set.seed(1)
  rainsamp <- sampleRandom(raintotalb, size = 1000, xy = TRUE)
  rainsamp <- as.data.frame(rainsamp)
  invdist <- gstat(id = "rain", formula = rain ~ 1, locations = ~x + y, 
                   data = rainsamp)
  invdistr <- interpolate(object = r, model = invdist)
  invdistrmsk <- mask(x = invdistr, mask = raintotalb)
})

idws <- stack(c(raintotalb, idw_list))

titles <- c("CHIRPS rainfall", "1000 pt IDW", "500 pt IDW", "250 pt IDW")
plot_noaxes(idws, main = titles, zlim = c(0, 150))
```

3. 
```{r, eval = FALSE}
districts %>% filter(ID %in% seq(15, 50, 5)) %>% 
  st_transform(st_crs(roads)) %>% st_geometry %>% st_centroid %>% 
  as_Spatial(.) %>% 
  distanceFromPoints(object = raintotalb, xy = .) %>% 
  mask(., raintotalb) %>% plot_noaxes
```

4. 

Pretty close to results of highest density sample. Slightly higher mean error. 
```{r, eval = FALSE}
# #1
data(zamprec)
zamprecalb <- projectRaster(from = zamprec, to = raintotalb)
names(zamprecalb) <- "rain"
elev <- resample(aggregate(x = demalb, fact = 5), y = raintotalb)

# #2
set.seed(1)
pts <- sampleRandom(x = zamprecalb, size = 25, sp = TRUE) %>% st_as_sf
pts <- pts %>% mutate(elev = raster::extract(x = elev, y = .)) 
pts_dat <- bind_cols(pts %>% data.frame %>% select(-geometry) %>% as_tibble, 
                     st_coordinates(pts) %>% as_tibble) %>% drop_na
  
# #3
p1 <- ggplot(pts_dat) + geom_point(aes(X, rain), col = "steelblue") +
  ylab("Rainfall (mm)")
p2 <- ggplot(pts_dat) + geom_point(aes(Y, rain), col = "blue2") + ylab("")
p3 <- ggplot(pts_dat) + geom_point(aes(elev, rain), col = "darkblue") + ylab("")
cowplot::plot_grid(p1, p2, p3, nrow = 1)

# #4
rain_lm <- lm(rain ~ X + Y + elev, data = pts_dat)
summary(rain_lm)

# #5
xs <- xFromCell(object = raintotalb, cell = 1:ncell(raintotalb))
ys <- yFromCell(object = raintotalb, cell = 1:ncell(raintotalb))
X <- Y <- raintotalb
values(X) <- xs
values(Y) <- ys

# #6
predst <- stack(X, Y, elev)
names(predst) <- c("X", "Y", "elev")
predrainr <- predict(object = predst, model = rain_lm)

# #7
s <- stack(zamprecalb, predrainr, (predrainr - zamprecalb) / zamprecalb * 100)
mae <- round(cellStats(abs(zamprecalb - predrainr), mean), 1)  

pnames <- c("'Observed' Rainfall", "Predicted Rainfall", "% Difference")
par(mfrow = c(1, 3), mar = c(0, 0, 1, 4))
for(i in 1:3) {
  plot_noaxes(s[[i]], main = pnames[i])
  if(i %in% 1:2) {
    pts %>% st_geometry %>% 
      plot(pch = 20, cex = 0.2, col = "grey70", add = TRUE)
  } else {
    mtext(side = 1, line = -3, cex = 0.8, 
          text = paste("Mean abs err =", mae, "mm"))
  }
}
```

***
<center>
[Back to home](https://agroimpacts.github.io/geospaar/)
</center>
***
